---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Distance sampling'
    slide-number: true
    echo: true
lang: en
---


## {content-align="center" style="font-size: 140%;"}

<br>

<center>
  
Lecture 8 -- Hierarchical distance sampling: <br>
simulation, fitting, and prediction

<br>
  
::: {style='font-size: 80%;'}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
:::
  
</center>




# Overview

## Distance sampling overview

Distance sampling is one of the oldest wildlife sampling methods.

<br>

It's based on the simple idea that detection probability should decline with distance between an animal and a transect.
  
<br>

::: {.fragment}
If we can estimate the function describing how $p$ declines with distance $x$, we can estimate abundance... 
:::
::: {.fragment}
... if certain assumptions hold, as always.
:::


## Distance sampling overview

::: {style='font-size: 80%;'}
The simplest estimator of abundance is 
$$
  \hat{N} = \frac{n}{\hat{p}}
$$
where $n$ is the number of individuals detected, $p$ is detection probability, and $E(n)=Np$.
  
  
::: {.fragment}
In distance sampling, detection probability is a *function* of distance, rather than a constant, such that all individuals have unique detection probabilities.
:::
  
  
::: {.fragment}
As a result, we have to replace $p$ with *average* detection probability:
$$
  \hat{N} = \frac{n}{\hat{\bar{p}}}
$$
:::
  
  
::: {.fragment}
How do we compute average detection probability ($\bar{p}$)?
:::
:::



## Detection functions

::: {style='font-size: 80%;'}
To estimate average detection probability ($\bar{p}$), we need:

- A detection function $g(x)$ describing how $p$ declines with distance.
- A probability distribution $p(x)$ describing the distances of all animals (detected and not detected).

The most common detection functions are:

Detection function   | $g(x)$
------------------   | -------
Half normal          | $\exp(-x^2 / (2\sigma^2))$
Negative exponential | $\exp(-x/\sigma)$
Hazard rate          | $1-\exp(-(x/a)^{-b})$
:::


## Half-normal

::: {style='font-size: 80%;'}
$$
    g(x,\sigma) = \exp(-x^2/(2\sigma^2))
$$
  
  
```{r, echo=-1}
#| label: hn
#| fig-width: 7
#| fig-height: 5
#| out-width: "50%"
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))  
sigma1 <- 25; sigma2 <- 50
plot(function(x) exp(-x^2/(2*sigma1^2)), from=0, to=100, 
     xlab="Distance (x)", ylab="Detection probability (p)")
plot(function(x) exp(-x^2/(2*sigma2^2)), from=0, to=100, add=TRUE, col=4)
legend(70, 1, c("sigma=25", "sigma=50"), lty=1, col=c("black","blue"))
```
:::


## Negative exponential

::: {style='font-size: 80%;'}
$$
    g(x,\sigma) = \exp(-x/\sigma)
$$
  
  
```{r, echo=-1}
#| label: nexp
#| fig-width: 7
#| fig-height: 5
#| out-width: "50%"
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))  
sigma1 <- 25; sigma2 <- 50
plot(function(x) exp(-x/sigma1), from=0, to=100, 
     xlab="Distance (x)", ylab="Detection probability (p)")
plot(function(x) exp(-x/sigma2), from=0, to=100, add=TRUE, col=4)
legend(70, 1, c("sigma=25", "sigma=50"), lty=1, col=c("black","blue"))
```
:::




## Hazard rate

::: {style='font-size: 80%;'}
$$
    g(x,a,b) = 1-\exp(-(x/a)^{-b})
$$
  
  
```{r, echo=-1}
#| label: haz
#| fig-width: 7
#| fig-height: 5
#| out-width: "50%"
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))  
a1 <- 25; a2 <- 50; b1 <- 2; b2 <- 10
plot(function(x) 1-exp(-(x/a1)^(-b1)), from=0, to=100, 
     xlab="Distance (x)", ylab="Detection probability (p)", ylim=0:1)
plot(function(x) 1-exp(-(x/a2)^(-b2)), from=0, to=100, add=TRUE, col=4)
legend(70, 1, c("a=25, b=2", "a=50, b=10"), lty=1, col=c("black","blue"))
```
:::


## Average detection probability

::: {style='font-size: 80%;'}
Regardless of the chosen detection function, average detection probability is defined as: 
$$
     \bar{p} = \int_{0}^{B} g(x)p(x) \; \mathrm{d}x
$$
where $B$ is the width of the transect.
  
::: {style='font-size: 80%;'}
All that remains is the specification of $p(x)$, the distribution of distances (between animals and the transect). 
:::
  
::: {style='font-size: 80%;'}
To understand why $p(x)$ is needed, think about it this way:
  
- If most animals are close to the transect, $\bar{p}$ would be high
- If most animals are far from the transect, $\bar{p}$ would be low
:::
:::



## What should we use for *p(x)*?

::: {style='font-size: 70%;'}
What distribution should we use for the distances between animals and transects?

::: {.fragment}
In *line-transect sampling*, it is often assumed that animals are uniformly distributed with respect to the transect.
  
- Consequently, $p(x) = 1/B$, where $x$ is the *perpendicular* distance between animal and transect
- This is guaranteed by random transect placement
- Can also be justified if animals are neither attracted to the transects or avoid them. 
:::
  
::: {.fragment}
In *point-transect sampling*, we make the same assumptions, but we recognize that area increases with distance from a point.
  
- Consequently, $p(x) = 2x/B^2$ (see pg. 408 in AHM)
- Here, $x$ is the *radial* distance to an animal
:::
:::
  

## [Computing average *p* for line transects]{style="font-size: 90%;"}

Half-normal detection function and line-transect sampling.
  
```{r}
#| label: pbar-hn
B <- 100                                         # Transect width
g <- function(x, sigma=25) exp(-x^2/(2*sigma^2)) # g(x)
pdf <- function(x) 1/B                           # p(x), constant
```

<br>

::: {.fragment}
Do the integration
  
```{r}
#| label: pbar-hn-int
gp <- function(x) g(x)*pdf(x)
(pbar <- integrate(gp, lower=0, upper=B)$value)
```

31.3\% chance of detecting an individual within 100 m.
:::


## [Computing average *p* for point transects]{style="font-size: 90%;"}

Half-normal detection function and point-transect sampling.
  
```{r}
#| label: pbar-hn-pt
B <- 100                                         # Transect width
g <- function(x, sigma=25) exp(-x^2/(2*sigma^2)) # g(x)
pdf <- function(x) 2*x/B^2                       # p(x)
```


<br>

::: {.fragment}
Do the integration
  
```{r}
#| label: pbar-hn-int-pt
gp <- function(x) g(x)*pdf(x)
(pbar <- integrate(gp, lower=0, upper=B)$value)
```

12.5\% chance of detecting an individual within 100 m. <br>
:::


## 

<br>

::: {.r-fit-text}
[Distance Sampling Shiny App](https://richard-chandler.shinyapps.io/distance-sampling/)
:::


## [Conventional vs hierarchical distance sampling]{style="font-size: 80%;"}

::: {style='font-size: 80%;'}
*Conventional* distance sampling
  
* Focus is on estimation of detection function parameters and density
* No model for spatial variation in density
* Data are individual-level distances
* We'll deal with CDS in more depth as a prelude to spatial-capture recapture
  
::: {.fragment}
*Hierarchical* distance sampling
  
* Focus is on estimation of detection function parameters and spatial variation in abundance/density
* Data are counts of individuals in each distance bin
* Multinomial $N$-mixture model with a unique function for computing the multinomial cell probabilities 
:::
 
:::
 






## Hierarchical distance sampling
  
::: {style='font-size: 70%;'}
State model (with Poisson assumption)
$$
    \mathrm{log}(\lambda_i) = \beta_0 + \beta_1 {w_{i1}} +
    \beta_2 {w_{i2}} + \cdots \\
    N_i \sim \mathrm{Poisson}(\lambda_i)
$$
  
Observation model
$$
    \mathrm{log}(\sigma_{i}) = \alpha_0 + \alpha_1 {w_{i1}}
    + \alpha_2 {w_{i3}} + \cdots \\
    \{y_{i1}, \dots, y_{iK}\}  \sim \mathrm{Multinomial}(N_i,
    \pi(x, \sigma_i))
$$
:::


::: {.fragment}
::: {style='font-size: 65%;'}

$\lambda_i$ -- Expected value of abundance at site $i$ <br>
$N_i$ -- Realized value of abundance at site $i$ <br>
$\sigma_{i}$ -- Scale parameter of detection function $g(x)$ at site $i$ <br>
$\pi(x,\sigma_i)$ -- Function computing multinomial cell probs <br>
$y_{ij}$ -- count for distance bin $j$ (final count is unobserved) <br>
$w_1$, $w_2$, $w_3$ -- site covariates

:::
:::







# Point transects

## [Multinomial cell probs for point transects]{style="font-size: 80%;"}
  
::: {style='font-size: 70%;'}
Definitions needed for computing \alert{bin-specific} $\bar{p}$ and multinomial cell probabilities. 
    
* $y_{ij}$ -- number of individuals detected at site $i$ in bin $j$
* $\sigma_i$ -- scale parameter of detection function $g(x)$
* $b_1, \dots, b_J$ -- Distance break points defining $J$
      distance intervals
* $\bar{p}_j$ -- Average detection probability in distance interval $j$.
* $\psi_j$ -- Probability of occuring in distance band $j$
:::
::: {style='font-size: 60%;'}
Description   | Multinomial cell probability
----------------------------- | ------------
Pr(occurs and detected in first distance bin)  | $\pi_1 = \psi_1\bar{p}_1$
Pr(occurs and detected in second distance bin) | $\pi_2 = \psi_2\bar{p}_2$
$\cdots$             | $\cdots$
Pr(occurs and detected in last distance bin)  | $\pi_J = \psi_J\bar{p}_J$
Pr(not detected)                  | $\pi_{K} = 1-\sum_{j=1}^J \pi_j$
:::


```{r}
#| include: false
#| echo: false
set.seed(34889243)
```


## Point transects, no covariates
  
::: {style='font-size: 70%;'}
Abundance
  
```{r}
#| label: sim-hds-nocovN3-pt
#| size: 'scriptsize'
B <- 100; Area <- pi*B^2/1e4   ## plot radius (B) and area in ha
nSites <- 100; lambda1 <- 0.9  ## Expected value of density
N3 <- rpois(n=nSites, lambda=lambda1*Area)
```
  
  
Multinomial cell probabilities
  
```{r}
#| label: sim-hds-nocov3-pt
#| size: 'scriptsize'
J <- 5                     # distance bins
sigma <- 50                # scale parameter
b <- seq(0, B, length=J+1) # distance break points
area <- pi*b^2             # area of each circle
psi <- (area[-1]-area[-(J+1)]) / area[J+1]
pbar3 <- numeric(J)        # average detection probability
pi3 <- numeric(J+1)        # multinomial cell probs
for(j in 1:J) {
    pbar3[j] <- integrate(function(x) exp(-x^2/(2*sigma^2))*x,
                          lower=b[j], upper=b[j+1])$value *
                          (2*pi/diff(area)[j])
    pi3[j] <- pbar3[j]*psi[j] }; pi3[J+1] <- 1-sum(pi3[1:J])
```
  
  
Detections in each distance interval
  
```{r}
#| label: sim-hds-y3
#| size: 'scriptsize'
y3.all <- matrix(NA, nrow=nSites, ncol=J+1)
for(i in 1:nSites) {
    y3.all[i,] <- rmultinom(n=1, size=N3[i], prob=pi3)    }
y3 <- y3.all[,1:J]  ## Drop final cell
```

:::




## Observed distances
  
```{r, echo=-1}
#| label: dist-hist3
#| fig.height: 5
#| out.width: "90%"
#| size: 'scriptsize'
par(mai=c(0.9,0.9,0.1,0.1))  
plot(b[-(J+1)]+10, colSums(y3), type="h", lwd=80, lend=2, col="darkcyan", 
     xlim=c(0,100), ylim=c(0, 70), xlab="Distance", ylab="Detections")
```






## Point transects, covariates
  
::: {style='font-size: 70%;'}
Abundance
  
```{r}
#| label: sim-hds-covN4-pt
#| size: 'scriptsize'
elevation <- rnorm(nSites)  
beta0 <- 2; beta1 <- 1
lambda4 <- exp(beta0 + beta1*elevation)   # E(density)
N4 <- rpois(n=nSites, lambda=lambda4*Area)
```
  
  
Multinomial cell probabilities
  
```{r}
#| label: sim-hds-cov4-pt
#| size: 'scriptsize'
noise <- rnorm(nSites)
alpha0 <- 3; alpha1 <- -0.5
sigma4 <- exp(alpha0 + alpha1*noise)
pi4 <- matrix(NA, nSites, J+1) # multinomial cell probs
for(i in 1:nSites) {
  for(j in 1:J) {
      pi4[i,j] <- integrate(function(x) exp(-x^2/(2*sigma4[i]^2))*x,
          lower=b[j], upper=b[j+1])$value*(2*pi/diff(area)[j])*psi[j] }
  pi4[i,J+1] <- 1-sum(pi4[i,1:J]) } 
```
  
  
Detections in each distance interval
  
```{r}
#| label: sim-hds-y4-pt
#| size: 'scriptsize'
y4.all <- matrix(NA, nrow=nSites, ncol=J+1)
for(i in 1:nSites) {
    y4.all[i,] <- rmultinom(n=1, size=N4[i], prob=pi4[i,])    }
y4 <- y4.all[,1:J]
```
:::



## Simulated data

::: {.columns}
::: {.column width='45%'}
      
Observations

  
```{r}
#| label: sim-nocov-dat4-pt
y4[1:25,]
```
:::
::: {.column width='55%'}
    
Summary stats

    
::: {style='font-size: 70%;'}

Proportion of sites known to be occupied
    
```{r}
#| label: sim-nocov-ssO-pt
#| size: 'scriptsize'
# Max count at each site
maxCounts <- apply(y4, 1, max) 
naiveOccupancy <- sum(maxCounts>0)/nSites
naiveOccupancy 
```
  
Total detections in each distance interval
  
```{r}
#| label: sim-nocov-ss4-pt
#| size: 'scriptsize'
colSums(y4)
```
  
  
Naive abundance
  
```{r}
#| label: sim-nocov-ss4sum-pt
#| size: 'scriptsize'
sum(y4)
```
:::
:::
:::









```{r}
#| label: un
#| include: false
library(unmarked)
```



## Prepare data in 'unmarked'
  
::: {style='font-size: 80%;'}
Note the new arguments.
  
```{r}
#| label: un-umf-pt
#| size: 'tiny'
umf4 <- unmarkedFrameDS(y=y4, siteCovs=data.frame(elevation,noise), dist.breaks=b,
                       survey="point", unitsIn="m")
```

```{r}
#| label: wfac-pt
#| size: 'tiny'
summary(umf4)
```
:::


## Fit the model

::: {style='font-size: 80%;'}
```{r}
#| label: un-fit-pt
## fm4 <- distsamp(~noise ~elevation, umf4, keyfun="exp")     # negative exp
## fm4 <- distsamp(~noise ~elevation, umf4, keyfun="hazard")  # hazard rate
fm4 <- distsamp(~noise ~elevation, umf4, keyfun="halfnorm")   # half-normal
fm4
```
:::


## Prediction in 'unmarked'
  
Create `data.frame` with prediction covariates. 
  
```{r}
#| label: preddat
pred.data <- data.frame(noise=seq(-3, 3, by=0.5))
```


Get predictions of $\sigma$ for each row of prediction data.
  
```{r}
#| label: predpsi
sigma.pred <- predict(fm4, newdata=pred.data,
                      type='det', append=TRUE)
```


View $\sigma$ predictions
  
```{r}
#| label: psi-head
print(head(sigma.pred), digits=2)
```







## Prediction in 'unmarked'

```{r, echo=-1}
#| label: pred-sigma
#| fig-width: 7
#| fig-height: 5
#| out-width: '80%'
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))
plot(Predicted ~ noise, sigma.pred, ylab="Scale parameter (sigma)",
     ylim=c(0,100), xlab="Noise level", type="l")
lines(lower ~ noise, sigma.pred, col="grey")
lines(upper ~ noise, sigma.pred, col="grey")
```




## Prediction in 'unmarked'

::: {style='font-size: 80%;'}
```{r, echo=-1}
#| label: pred-sigma2
#| fig-width: 7
#| fig-height: 5
#| out-width: '50%'
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))
plot(Predicted ~ noise, sigma.pred, ylab="Scale parameter (sigma)",
     ylim=c(0,100), xlab="Noise level", type="l")
lines(lower ~ noise, sigma.pred, col="grey")
lines(upper ~ noise, sigma.pred, col="grey")
arrows(-2, sigma.pred$Predicted[3], 0.1, 63, len=0.1, col="blue")
points(-2, sigma.pred$Predicted[3], col="orange", pch=16, cex=2)
par( fig=c(0.5,0.95,0.55,0.95), new=TRUE, las=1)#, mar=c(0,0,0,0) )
plot(function(x) exp(-x^2/(2*40^2)), 0, 100, xlab="Dist", ylab="p",
     col="orange", lwd=2, ylim=c(0,1), las=1)
```
:::



## Prediction in 'unmarked'

::: {style='font-size: 80%;'}
```{r, echo=-1}
#| label: pred-sigma2-2
#| fig-width: 7
#| fig-height: 5
#| out-width: '50%'
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))
plot(Predicted ~ noise, sigma.pred, ylab="Scale parameter (sigma)",
     ylim=c(0,100), xlab="Noise level", type="l")
lines(lower ~ noise, sigma.pred, col="grey")
lines(upper ~ noise, sigma.pred, col="grey")
arrows(1, sigma.pred$Predicted[9], 1.1, 55, len=0.1, col="blue")
points(1, sigma.pred$Predicted[9], col="red", pch=16, cex=2)
par( fig=c(0.5,0.95,0.55,0.95), new=TRUE, las=1)#, mar=c(0,0,0,0) )
plot(function(x) exp(-x^2/(2*15^2)), 0, 100, xlab="Dist", ylab="p",
     col="red", lwd=2, ylim=c(0,1), las=1)
```
:::



##  Conditional-on-*N* and *n~i~*

[distsamp-point-mod.jag](distsamp-point-mod.jag)
```{r}
#| label: bugs-pt
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("distsamp-point-mod.jag"))
```



## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data-pt
jags.data.pt <- list(y=y4, n=rowSums(y4), area=diff(area),
                     b=b,              # Distance break points
                     Area=pi*B^2/1e4,  # Area in ha
                     psi=psi,          # Pr(occuring in bin j)
                     elevation=elevation, noise=noise,
                     nSites=nSites, nBins=J)
```

Initial values
  
```{r}
#| label: bugs-inits-pt
jags.inits.pt <- function() {
    list(beta0=rnorm(1), alpha0=rnorm(1, 5),
         N=rowSums(y4)+rpois(nrow(y4), 2))
}
```

Parameters to monitor
  
```{r}
#| label: bugs-pars-pt
jags.pars.pt <- c("beta0", "beta1",
                  "alpha0", "alpha1", "totalAbundance")
```



## MCMC
  
```{r}
#| label: bugs-mcmc-pt
#| message: false
#| cache: false
#| results: 'hide'
library(jagsUI)
jags.post.pt <- jags.basic(data=jags.data.pt, inits=jags.inits.pt,
                           parameters.to.save=jags.pars.pt,
                           model.file="distsamp-point-mod.jag",
                           n.chains=3, n.adapt=100, n.burnin=0,
                           n.iter=2000, parallel=TRUE)
```

```{r}
#| label: jags-sum-pt
#| cache: true
round(summary(jags.post.pt)$quantile, digits=3)
```


# Line transects

## [Multinomial cell probs for line transects]{style="font-size: 90%;"}
  
Definitions needed for computing \alert{bin-specific} $\bar{p}$ and multinomial cell probabilities. 
  
::: {style='font-size: 80%;'}
* $y_{ij}$ -- number of individuals detected at site $i$ in bin $j$
* $\sigma_i$ -- scale parameter of detection function $g(x)$
* $b_1, \dots, b_{J+1}$ -- Distance break points defining $J$ distance intervals
* $\bar{p}_j$ -- Average detection probability in distance bin j
* $\psi_j$ -- Pr(occuring in distance bin $j$)
:::
  

```{r}
#| include: false
#| echo: false
set.seed(34889243)
```


## Line transects, no covariates
  
::: {style='font-size: 70%;'}
Abundance
  
```{r}
#| label: sim-hds-nocov1
B <- 100; L <- 100; A <- 2*B*L/1e4  ## transect widths, length, and area
nSites <- 100; lambda1 <- 1.3       ## Expected value of density
N1 <- rpois(n=nSites, lambda=lambda1*A)
```
  
  
Multinomial cell probabilities
  
```{r}
#| label: sim-hds-nocov2
J <- 5                     # distance bins
sigma <- 50                # scale parameter
b <- seq(0, B, length=J+1) # distance break points
psi <- diff(b)/B           # Pr(x is in bin j)
pbar1 <- numeric(J)        # average detection probability
pi1 <- numeric(J+1)        # multinomial cell probs
for(j in 1:J) {
    pbar1[j] <- integrate(function(x) exp(-x^2/(2*sigma^2)),
                          lower=b[j], upper=b[j+1])$value / diff(b)[j]
    pi1[j] <- pbar1[j]*psi[j]
}
pi1[J+1] <- 1-sum(pi1[1:J])
```
  
  
Detections in each distance interval
  
```{r}
#| label: sim-hds-y1
y1.all <- matrix(NA, nrow=nSites, ncol=J+1)
for(i in 1:nSites) {
    y1.all[i,] <- rmultinom(n=1, size=N1[i], prob=pi1)    }
y1 <- y1.all[,1:J]  ## Drop final cell
```
:::


## Observed distances
  
```{r, echo=-1}
#| label: dist-hist2
#| fig-height: 5
#| out-width: "90%"
par(mai=c(0.9,0.9,0.1,0.1))  
plot(b[-(J+1)]+10, colSums(y1), type="h", lwd=80, lend=2, col="darkcyan",
     xlim=c(0,100), ylim=c(0, 70), xlab="Distance", ylab="Detections")
```



## Line transects, covariates
  
::: {style='font-size: 70%;'}
Abundance
  
```{r}
#| label: sim-hds-cov1
elevation <- rnorm(nSites)  
beta0 <- 2; beta1 <- 1
lambda2 <- exp(beta0 + beta1*elevation)
N2 <- rpois(n=nSites, lambda=lambda2*A)
```
  
  
Multinomial cell probabilities
  
```{r}
#| label: sim-hds-cov2
noise <- rnorm(nSites)
alpha0 <- 3; alpha1 <- -0.5
sigma2 <- exp(alpha0 + alpha1*noise)
pi2 <- matrix(NA, nSites, J+1) # multinomial cell probs
for(i in 1:nSites) {
  for(j in 1:J) {
      pi2[i,j] <- integrate(function(x) exp(-x^2/(2*sigma2[i]^2)),
          lower=b[j], upper=b[j+1])$value / (b[j+1]-b[j]) * psi[j] }
  pi2[i,J+1] <- 1-sum(pi2[i,1:J]) } 
```
  
  
Detections in each distance interval
  
```{r}
#| label: sim-hds-y2
y2.all <- matrix(NA, nrow=nSites, ncol=J+1)
for(i in 1:nSites) {
    y2.all[i,] <- rmultinom(n=1, size=N2[i], prob=pi2[i,])    }
y2 <- y2.all[,1:J]
```

:::



## Simulated data

::: {.columns}

::: {.column width='45%'}
      
Observations
  
```{r}
#| label: sim-nocov-dat
y2[1:25,]
```
:::
::: {.column width='55%'}
    
::: {style='font-size: 80%;'}
Summary stats
    
    
Proportion of sites known to be occupied
    
```{r}
#| label: sim-nocov-ss1
# Max count at each site
maxCounts <- apply(y2, 1, max) 
naiveOccupancy <- sum(maxCounts>0)/nSites
naiveOccupancy 
```

Total detections in each distance interval
  
```{r}
#| label: sim-nocov-ss2
colSums(y2)
```
  
Naive abundance
  
```{r}
#| label: sim-nocov-ss3
sum(y2)
```
:::
:::
:::







## Prepare data in 'unmarked'
  
::: {style='font-size: 80%;'}
Note the new arguments.
  
```{r}
#| label: un-umf
umf <- unmarkedFrameDS(y=y2, siteCovs=data.frame(elevation,noise), dist.breaks=b,
                       tlength=rep(L, nSites), survey="line", unitsIn="m")
```

```{r}
#| label: wfac
summary(umf)
```
:::


## Fit the model

::: {style='font-size: 70%;'}
```{r}
#| label: un-fit
## fm <- distsamp(~noise ~elevation, umf, keyfun="exp")     # negative exp
## fm <- distsamp(~noise ~elevation, umf, keyfun="hazard")  # hazard rate
fm <- distsamp(~noise ~elevation, umf, keyfun="halfnorm")   # half-normal
fm
```


Compare to actual parameter values:

```{r}
#| label: un-compare
c(beta0=beta0, beta1=beta1); c(alpha0=alpha0, alpha1=alpha1)
```
:::

##  Conditional-on-*N* and *n~i~*

[distsamp-line-mod.jag](distsamp-line-mod.jag)
```{r}
#| label: bugs-line
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("distsamp-line-mod.jag"))
```

## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data2
jags.data.line <- list(y=y2, n=rowSums(y2),
                       b=b,           # Distance break points
                       psi=diff(b)/B, # Pr(occuring in bin j)
                       Area=A,        # Area in ha
                       elevation=elevation, noise=noise,
                       nSites=nSites, nBins=J)
```


Initial values
  
```{r}
#| label: bugs-inits
jags.inits.line <- function() {
    list(beta0=rnorm(1), alpha0=rnorm(1, 5),
         N=rowSums(y2)+rpois(nrow(y2), 2))
}
```


Parameters to monitor
  
```{r}
#| label: bugs-pars
jags.pars.line <- c("beta0", "beta1",
                    "alpha0", "alpha1", "totalAbundance")
```



## MCMC
  
```{r}
#| label: bugs-mcmc-line
#| message: false
#| cache: true
#| results: 'hide'
library(jagsUI)
jags.post.line <- jags.basic(data=jags.data.line, inits=jags.inits.line,
                             parameters.to.save=jags.pars.line,
                             model.file="distsamp-line-mod.jag",
                             n.chains=3, n.adapt=100, n.burnin=0,
                             n.iter=2000, parallel=TRUE)
```

```{r}
#| label: jags-sum-line
#| cache: true
round(summary(jags.post.line)$quantile, digits=3)
```




## Traceplots and density plots

```{r}
#| label: bugs-plot1-rem2
#| out-width: "0.7<br>textwidth"
#| fig-align: 'center'
#| cache: true
plot(jags.post.line[,jags.pars.line[1:3]])
```



## Traceplots and density plots

```{r}
#| label: bugs-plot2-rem2
#| size: 'footnotesize'
#| out-width: "0.7<br>textwidth"
#| fig-align: 'center'
#| cache: true
plot(jags.post.line[,jags.pars.line[4:5]])
```




# Summary



## Distance sampling summary

Assumptions
  
::: {style='font-size: 80%;'}
- Animals don't move during the survey
- Animals are uniformly distributed with respect to the transects
- Detection is certain on the transect, i.e. $p=1$ when $x=0$. 
- Detections are independent
:::

::: {.fragment}
If these assumptions can be met, distance sampling is a powerful method allowing for inference about abundance and density using data from a single visit.
:::




# Assignment


## Assignment

::: {style='font-size: 65%;'}
Create a self-contained R script or Rmarkdown file to do the following:
  
- Fit a distance sampling model with a half-normal detection
     function and the following covariates to the black-throated blue
      warbler data (`btbw\_data\_distsamp.csv`) in 'unmarked' and
      'JAGS':
  * Density covariates: `Elevation, UTM.N, UTM.W`
  * Detection covariates: `Wind, Noise`
  * Response: `btbw0_20, btbw20_40, btbw40_60, btbw60_80, btbw80_100`
      
- Using the model fitted in 'unmarked', create two graphs of
  the predictions: one for density and the other for the scale
  parameter ($\sigma$).
  
Suggestions:
  
- Convert response variables to matrix with \inr{as.matrix}
- Standardize covariates
  

Upload your `.R`, `.Rmd`, or `.qmd` file to ELC by 5:00pm on Tuesday. 

:::
