---
title: "Inference for Models of Population Dynamics"
subtitle: Basic models of population dynamics
author: Richard Chandler
output:
  xaringan::moon_reader:
    self_contained: false
    css: "../../api-lectures.css"
    chakra: "../../libs/remark-latest.min.js"
    nature:
      titleSlideClass: ["center", "middle", "title-slide"]
      countIncrementalSlides: false
---


## Overview

Everyone should be familiar with the following models, but we will review them in some detail:

- Exponential and geometric growth

- Logistic growth

- Stage-structured (matrix) models

- Stochastic versions of the above


For every model, you should know the equations and you should know how to implement it in `R`.


---

## Stage structured models

$${\bf n}(t+1) = {\bf A}{\bf n}(t)$$

where $A$ is the projection matrix and $n$ is the vector of stage-specific abundances

--

Example with 3 stages (juveniles, subadults, and adults)


---

## Stage-structured models in `R`

```{r stage-ex}
f1 <- 0.0   ## Juvenile fecundity
f2 <- 0.5   ## Subadult fecundity
f3 <- 1.5   ## Adult fecundity
s1 <- 0.3   ## Juvenile survival
s2 <- 0.5   ## Subadult survival
s3 <- 0.8   ## Adult survival
A <- matrix(c(f1, f2, f3,
              s1,  0,  0,
               0, s2, s3), nrow=3, ncol=3, byrow=TRUE)
A
```

-- 

Why do we structure the projection matrix this way?

--

The answer is that it allows us to do convenient operations (e.g., eigenanalysis and matrix multiplication) on the computer. 

---

## Matrix multiplication


---

## Eigenanalysis

---

## Asymptotic growth rate

```{r lambda}
eA <- eigen(A)
lambda <- Re(eA$values[1])
lambda ## Asymtotic growth rate
```

---

## Net reproductive rate

---

## Stable stage distribution

---

## Reproductive value


---


## Stage-structured model with transition probabilities

---

## Stochastic stage-structured models


