---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Model selection and model fit'
    slide-number: true
lang: en
---

## {content-align="center" style="font-size: 140%;"}

<br>

<center>
Lecture 6 -- Binomial $N$-mixture models: <br>
model selection <!-- % , spatial prediction, -->
and goodness-of-fit 

<br>
  
::: {style='font-size: 80%;'}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
:::

</center>



# Overview {visibility="hidden"}

## Overview

<br>

How should we choose among competing models?

<br>

What should we do if our model doesn't fit the data well?



# Model selection



## Model selection

::: {style='font-size: 75%;'}
In scientific contexts, we want models that describe natural
processes and allow us to evaluate hypotheses. 

<br>

::: {.fragment}
Models should be predictive, but they shouldn't be crafted with the
sole goal of prediction in mind. 
:::

<br>

::: {.fragment}
Scientists typically don't care about prediction if the model doesn't
help us learn about the processes that gave rise to the data. 
:::

<br>

::: {.fragment}
Nonetheless, predictive performance is often the best way to compare
models and avoid underfitting and overfitting. 
:::

<br>

::: {.fragment}
Make sure the models being compared were motivated by clear
hypotheses. 
:::

:::



## Model selection

::: {style='font-size: 80%;'}
Prediction accuracy increases with model complexity up to a point,
until overfitting kicks in. 
  
::: {.fragment}
The best way to determine if a model is too simplistic or too complex
is to compare predictions to new observations. 
:::
  
::: {.fragment}
However, we rarely have the resources to collect additional data for
the sake of evaluating predictive performance.
:::

::: {.fragment}
A cheaper (albeit less desirable) alternative is to use
cross-validation:
  
- Split data into K partitions
- Fit model to the data in K-1 partitions
- Predict the holdout partition
:::
  

::: {.fragment}
<center>
Information criteria like AIC and WAIC yield similar predictive rankings as cross-validation, but with less computation.
</center>
:::

:::




## Overfitting

::: {style='font-size: 80%;'}
7 regression models with 5 covariates. Only `x2` is in the data generating model.


```{r, echo=2:14}
#| label: overfit
#| fig-width: 8
#| fig-height: 7
#| fig-align: 'center'
#| out-width: "65%"
#| output-location: 'slide'
set.seed(43340)
n <- 50
dat <- data.frame(x1=rnorm(n), x2=rnorm(n), x3=rnorm(n),
                  x4=rnorm(n), x5=rnorm(n))
dat$y <- rnorm(n, mean = -1 + 2*dat$x2, sd = 2) # Data generating model
fm1 <- glm(y~1, gaussian, dat)
fm2 <- glm(y~x1, gaussian, dat)
fm3 <- glm(y~x2, gaussian, dat)                 # Data generating model
fm4 <- glm(y~x1+x2, gaussian, dat)
fm5 <- glm(y~x1+x2+x3, gaussian, dat)
fm6 <- glm(y~x1+x2+x3+x4, gaussian, dat)
fm7 <- glm(y~x1+x2+x3+x4+x5, gaussian, dat)

library(boot)  ## For 'cv.glm'
prediction_error <- c(fm1=cv.glm(dat, fm1)$delta[1],
  fm2=cv.glm(dat, fm2)$delta[1], fm3=cv.glm(dat, fm3)$delta[1],
  fm4=cv.glm(dat, fm4)$delta[1], fm5=cv.glm(dat, fm5)$delta[1],
  fm6=cv.glm(dat, fm6)$delta[1], fm7=cv.glm(dat, fm7)$delta[1])

xbp <- barplot(prediction_error, xlab="Model", ylab="Predication error", cex.lab=1.3)
text(xbp[1], 0.3, as.character(fm1$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[2], 0.3, as.character(fm2$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[3], 0.3, as.character(fm3$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[4], 0.3, as.character(fm4$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[5], 0.3, as.character(fm5$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[6], 0.3, as.character(fm6$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[7], 0.3, as.character(fm7$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[3], prediction_error[3]+0.1, "Data generating model", srt=90,
     pos=4, cex=1.3)
segments(0.3, 9.5, 2.2, 9.5, xpd=TRUE)
text(1.2, 10, "Under-fit", xpd=TRUE, cex=1.3)
segments(4, 4.5, 8.2, 4.5, xpd=TRUE)
text(6, 5, "Over-fit", xpd=TRUE, cex=1.3)
```

:::





## Model selection with AIC
  
::: {style='font-size: 80%;'}

Most information criteria (IC) take on a form like this:
$$
  \mathrm{IC} = \mathrm{fit} + \mathrm{penalty}
$$

where `fit` describes the *in-sample* predictive accuracy and
`penalty` describes model complexity. 
  
::: {.fragment}
In the case of Akaike's "An Information Criterion", which approximates
leave-one-out cross validation, `fit` is described by the likelihood
and the `penalty` is determined by the number of parameters: 
$$
  \mathrm{AIC} = -2 \log(L) + 2P 
$$
where $L$ is the likelihood evaluated at the MLE and $P$ is the number
of parameters. 
:::
   
::: {.fragment}
The lower the AIC, the better the *out-of-sample* predictive
performance. 
:::

:::




## Model selection in unmarked
  
::: {style='font-size: 80%;'}
Import the grouse data
  
```{r}
#| label: grouse-in
#| echo: true
#| cache: false
#| results: 'hide'
#| message: false
library(unmarked)
grouse.data <- read.csv("grouse_data_Nmix.csv", row.names=1)
grouse.umf <- unmarkedFramePCount(
    y=grouse.data[,paste0("grouse",1:3)],
    siteCovs=grouse.data[,c("utmE","utmN","elevation")],
    obsCovs=list(temp=grouse.data[,paste0("Temperature.",1:3)]))
```


::: {.fragment}
Standardize the covariates
  
```{r}
#| label: grouse-stand
#| echo: true
#| cache: false
## scale() only works if all the covariates are continuous  
site.covs.s <- scale(siteCovs(grouse.umf))
colnames(site.covs.s) <- paste0(colnames(site.covs.s), ".s")
siteCovs(grouse.umf) <- cbind(siteCovs(grouse.umf), site.covs.s)
obsCovs(grouse.umf) <- scale(obsCovs(grouse.umf))
```
:::

:::




## Model selection in `unmarked'
  
::: {style='font-size: 80%;'}
Fit some models
  
```{r}
#| label: grouse-mods
#| echo: true
#| warning: false
#| cache: true
fm1 <- pcount(~temp ~ elevation.s+utmE.s+utmN.s, grouse.umf, K=50) 
fm2 <- pcount(~temp ~ elevation.s+utmN.s, grouse.umf, K=50)
fm3 <- pcount(~temp ~ elevation.s, grouse.umf, K=50)
fm4 <- pcount(~1 ~ elevation.s+utmN.s, grouse.umf, K=50)
fm5 <- pcount(~1 ~ elevation.s, grouse.umf, K=50)
fm6 <- pcount(~1 ~ 1, grouse.umf, K=50)
```


::: {.fragment}
Put models in a special type of list
  
```{r}
#| label: grouse-fitlist
#| echo: true
#| warning: false
#| error: true
grouse.models <- fitList('lam(elev+utmE+utmN)p(temp)'=fm1,
                         'lam(elev+utmN)p(temp)'=fm2,
                         'lam(elev)p(ptemp)'=fm3,
                         'lam(elev+utmN)p(.)'=fm4,
                         'lam(elev)p(.)'=fm5,
                         'lam(.)p(.)'=fm6)

```

Uh oh, missing values differ among models.
:::

:::




## Model selection in `unmarked'

::: {style='font-size: 80%;'}
Replace count data with `NA` where associated covariates are missing:

```{r}
#| label: addNA
#| echo: true
na.sites <- apply(is.na(site.covs.s), 1, any)
grouse.counts <- getY(grouse.umf)
grouse.counts[na.sites,] <- NA
grouse.umf@y <- grouse.counts
```

  
  
::: {.fragment}
Fit the models again
  
```{r}
#| label: grouse-mods2
#| echo: true
#| warning: false
#| cache: true
fm1 <- pcount(~temp ~ elevation.s+utmE.s+utmN.s, grouse.umf, K=50)
fm2 <- pcount(~temp ~ elevation.s+utmN.s, grouse.umf, K=50)
fm3 <- pcount(~temp ~ elevation.s, grouse.umf, K=50)
fm4 <- pcount(~1 ~ elevation.s+utmN.s, grouse.umf, K=50)
fm5 <- pcount(~1 ~ elevation.s, grouse.umf, K=50)
fm6 <- pcount(~1 ~ 1, grouse.umf, K=50)
```
:::


::: {.fragment}
Put models in a `fitList`
  
```{r}
#| label: grouse-fitlist2
#| echo: true
#| warning: false
grouse.models <- fitList('lam(elev+utmE+utmN)p(temp)'=fm1,
                         'lam(elev+utmN)p(temp)'=fm2,
                         'lam(elev)p(ptemp)'=fm3,
                         'lam(elev+utmN)p(.)'=fm4,
                         'lam(elev)p(.)'=fm5,
                         'lam(.)p(.)'=fm6)
```
:::

:::




## Model selection in unmarked
  
Create AIC table
```{r}
#| label: aic-table
modSel(grouse.models)
```

<br>

::: {style='font-size: 90%;'}
Use the model with the lowest AIC for prediction, or model-average the predictions.

<br>

::: {.fragment}
<center>
For inference, you should use causal inference methods rather than blind model selection. 
</center>
:::
:::





## WAIC

::: {style='font-size: 80%;'}
Watanabe's widely applicable information criterion (WAIC) is a
generalization of AIC suitable for Bayesian models. 
  
As before, we have 

$$
  \mathrm{WAIC} = \mathrm{fit} + \mathrm{penalty}
$$

  
The measure of `fit` in this case is the log-pointwise-predictive-density:

:::

::: {style='font-size: 70%;'}

$$
   \mathrm{lppd} = \sum_{i=1}^M \log\left(\frac{1}{S} \sum_s p(y_i|\theta_s)\right)
$$
  
  
The penalty is proportional to the posterior variance:
$$
  \mathrm{penalty} = \sum_{i=1}^M \mathrm{Var}(\log p(y_i|\theta_s)))
$$

:::




## Data, inits, and parameters

::: {style='font-size: 80%;'}
Put data in a named list
  
```{r}
#| label: bugs-data
#| echo: true
jags.data <- list(
    y=grouse.counts,
    elevation=site.covs.s[,"elevation.s"],
    utmE=site.covs.s[,"utmE.s"],
    utmN=site.covs.s[,"utmN.s"],
    temp=as.matrix(grouse.data[,paste0("Temperature.",1:3)]),
    nSites=nrow(grouse.counts),
    nOccasions=ncol(grouse.counts))
jags.data$temp <- (jags.data$temp-mean(jags.data$temp, na.rm=TRUE))/
    sd(jags.data$temp, na.rm=TRUE) # standardize temperature
```


Initial values
  
```{r}
#| label: bugs-inits
#| echo: true
jags.inits <- function() {
    list(lambda.intercept=runif(1), alpha0=rnorm(1),
         N=rep(2, jags.data$nSites))
}
```


Parameters to monitor
  
```{r}
#| label: bugs-pars
#| echo: true
jags.pars <- c("beta0", "beta1", "beta2", "beta3",
               "alpha0", "alpha1", "totalAbundance",
               "ld.y.dot", "ld.ydot.N")
```
:::




## The BUGS model
  
::: {style='font-size: 80%;'}
Notice `modswitch`, which we can use to include/exclude predictors. 
:::

[Nmix-model-grouse.jag]("Nmix-model-grouse.jag")
```{r}
#| label: bugs
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model-grouse.jag"))
```

```{r}
#| label: jagsUI
#| include: false
library(jagsUI)
```


## MCMC

::: {style='font-size: 80%;'}
Fit Model 1, corresponding to `fm1`.
  
```{r}
#| label: bugs-mcmc
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
library(jagsUI)
jags.data1 <- jags.data
jags.data1$modswitch <- c(1,1,1,1) ## Include all covariates
jm1 <- jags.basic(data=jags.data1, inits=jags.inits,
                  parameters.to.save=jags.pars, 
                  model.file="Nmix-model-grouse.jag",
                  n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```


Fit Model 2
  
```{r}
#| label: bugs-mcmc2
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jags.data2 <- jags.data; jags.data2$modswitch <- c(1,0,1,1) 
jm2 <- jags.basic(data=jags.data2, inits=jags.inits,
                  parameters.to.save=jags.pars, 
                  model.file="Nmix-model-grouse.jag",
                  n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
:::



## MCMC
  
::: {style='font-size: 70%;'}
Model 3
  
```{r}
#| label: bugs-mcmc3
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jags.data3 <- jags.data; jags.data3$modswitch <- c(1,0,0,1)
jm3 <- jags.basic(data=jags.data3, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 4
  
```{r}
#| label: bugs-mcmc4
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jags.data4 <- jags.data; jags.data4$modswitch <- c(1,0,1,0)
jm4 <- jags.basic(data=jags.data4, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 5
  
```{r}
#| label: bugs-mcmc5
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jags.data5 <- jags.data; jags.data5$modswitch <- c(1,0,0,0)
jm5 <- jags.basic(data=jags.data5, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 6
  
```{r}
#| label: bugs-mcmc6
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jags.data6 <- jags.data; jags.data6$modswitch <- c(0,0,0,0)
jm6 <- jags.basic(data=jags.data6, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```

:::


## WAIC

::: {style='font-size: 80%;'}
$$
  \mathrm{WAIC} = -2\left(\mathrm{lppd} - \sum_i \mathrm{Var}(\log p(y_i|\theta))\right)
$$

```{r}
#| label: waic-fn-ld-op
#| include: false
waic <- function(x, focus=c("y", "yN")) {
    vars <- coda::varnames(x)
    if(focus[1]=="y") {
        ld.samples <- as.matrix(x[,grep("ld.y.dot", vars)])
    } else if(focus[1]=="yN") {
        ld.samples <- as.matrix(x[,grep("ld.ydot.N", vars)])
    } else stop("focus should be either 'y' or 'yN'")
    lppd <- sum(log(colMeans(exp(ld.samples))))
    penalty <- sum(apply(ld.samples, 2, var))
    return(-2*lppd + 2*penalty)
}
```

```{r}
#| label: waic-fn
#| echo: true
waic <- function(x) {
    ## Parameter names
    vars <- coda::varnames(x)
    ## Extract log-density of y at each site for each post sample
    ld.samples <- as.matrix(x[,grep("ld.y.dot", vars)])
    ## Compute log-pointwise-predictive-density
    lppd <- sum(log(colMeans(exp(ld.samples))))
    ## Compute penalty
    penalty <- sum(apply(ld.samples, 2, var))
    ## Return WAIC
    return(-2*(lppd-penalty))
}
```
:::


## WAIC

WAIC for the six models

::: {.columns}
::: {.column width='50%'}
```{r}
#| label: waic1
#| echo: true
#| size: 'scriptsize'
(waic1 <- waic(jm1))
```
```{r}
#| label: waic2
#| echo: true
#| size: 'scriptsize'
(waic2 <- waic(jm2))
```
```{r}
#| label: waic3
#| echo: true
#| size: 'scriptsize'
(waic3 <- waic(jm3))
```
:::
::: {.column width='50%'}
```{r}
#| label: waic4
#| echo: true
#| size: 'scriptsize'
(waic4 <- waic(jm4))
```
```{r}
#| label: waic5
#| echo: true
#| size: 'scriptsize'
(waic5 <- waic(jm5))
```
```{r}
#| label: waic6
#| echo: true
#| size: 'scriptsize'
(waic6 <- waic(jm6))
```
:::
:::
  
  
In practice, you would discard burn-in and ensure convergence before computing WAIC.


## WAIC

Create a WAIC table:
  
<center>
```{r}
#| label: waic-table
#| echo: true
waic.table <- data.frame(WAIC=c(waic1,waic2,waic3,waic4,waic5,waic5))
rownames(waic.table) <- paste("Model", 1:6)
waic.table$delta <- waic.table$WAIC-min(waic.table$WAIC)
waic.table <- waic.table[order(waic.table$WAIC),]
knitr::kable(waic.table, format="latex", digits=2, booktabs=TRUE)
```
</center>




# Goodness-of-fit


## Goodness-of-fit
  
::: {style='font-size: 70%;'}
Distributional assumptions determine the expected values
*and the expected variance* of the random variables, including
the data.

::: {.fragment}
Overdispersion occurs when there is more variance in the data than
expected by the model. Goodness-of-fit methods assess overdispersion and underdispersion.
:::


::: {.fragment}
If the model does not fit the data very well because of
overdispersion, there are several remedial actions:
  
- Scientific approach
  * Figure out why there is unexplained variation
  * Perhaps there were unmeasured covariates or you need a better process model
    
- Statistical approach
  * Soak up variation with random effects
  * Use a different distribution (we'll start here)
:::

:::
    
  
## Changing the distribution for $N$
  
::: {style='font-size: 60%;'}

Standard Poisson-binomial $N$-mixture model (without covariates):
$$
  N_i \sim \mathrm{Poisson}(\lambda) \\
  y_{ij} \sim \mathrm{Binomial}(N_i, p)
$$

::: {.fragment}
We can replace the Poisson distribution with other distributions
that allow for greater variance in $N$.  Two common examples
are the negative  binomial:
$$
    N_i \sim \mathrm{NegBin}(\lambda_i, \kappa)
$$
where $\lambda_i$ is the expected value of $N$ and $\kappa$ is the
dispersion parameter.
<!-- ^[There are several other -->
<!-- parameterizations of the negative binomial]. -->
:::


::: {.fragment}
Another option is the zero-inflated Poisson:

::: {.columns}

::: {.column width='50%'}
$$
  N_i \sim \mathrm{Poisson}(\lambda_i z_i) \\
  z_i \sim \mathrm{Bern}(\psi)
$$
:::
::: {.column width='50%'}
$$
  N_i \sim \mathrm{ZIPoisson}(\lambda_i,\psi)
$$
:::

:::

where $\psi$ is the expected proportion of sites with excess zeros. 

:::
:::


## Simulating data from ZIP model

::: {style='font-size: 80%;'}
Zero-inflated Poisson model

::: {.columns}

::: {.column width='50%'}
$$
  N_i \sim \mathrm{Poisson}(\lambda_i z_i) \\
  z_i \sim \mathrm{Bern}(\psi)
$$
:::

::: {.column width='50%'}
$$
  N_i \sim \mathrm{ZIPoisson}(\lambda_i,\psi)
$$
:::

:::


R code
```{r}
#| label: zip-sim
#| echo: true
nSites <- 20
nVisits <- 3
psi <- 0.5   # Proportion of extra-Poisson zeros
z <- rbinom(n=nSites, size=1, prob=psi) # Extra zeros
lam <- 5     # expected count when z=1
N <- rpois(n=nSites, lambda=lam*z)      # abundance at each site
p <- 0.5     # detection prob
y <- matrix(NA, nSites, nVisits)
for(i in 1:nSites) {
    y[i,] <- rbinom(n=nVisits, size=N[i], prob=p) # count data
}
```

:::


## Goodness-of-fit in unmarked
  
::: {style='font-size: 80%;'}
One of the simplest and best ways of assessing goodness-of-fit is to
look at the residuals.

```{r}
#| label: resid2
#| echo: true
#| fig-width: 9
#| out-width: '60%'
#| fig-align: 'center'
#| fig-keep: 'first'
plot(fm2)
```

No sign of obvious outliers or other problems.
:::


## Goodness-of-fit in unmarked

::: {style='font-size: 80%;'}

The parametric bootstrap can be used to assess model fit:

- Simulate a dataset from the fitted model
- Fit the model to the new dataset
- Compute a fit statistic
- Repeat steps 1-3 several hundred/thousand times
- Compare the distribution of the *expected* fit statistic to the fit statistic associated with the actual data.
  
::: {.fragment}
The fit statistic associated with the actual data should not be in
an extreme quantile of the distribution of the expected values.
:::
  
::: {.fragment}
Assuming a type I error rate of 0.05, the $p$-value should be >0.05 if the model fits the data well, which is the null hypothesis.
:::

:::


## Goodness-of-fit in unmarked

```{r}
#| label: parboot
#| echo: true
#| warning: false
#| cache: true
#| fig-width: 9
#| out-width: '70%'
#| fig-align: 'center'
fitstat <- function(fm) { ## Sum of squared-residuals
    return(c(SSE=sum(residuals(fm)^2, na.rm=TRUE))) 
}
pb <- parboot(fm2, statistic=fitstat, nsim=200, ncores=3); plot(pb)
```



## Goodness-of-fit in unmarked

```{r}
#| label: parboot2
#| echo: true
pb
```

<br>

These results suggest that model `fm2` fits the data well.



## Goodness-of-fit in unmarked
  
::: {style='font-size: 80%;'}
If the model didn't fit the data well, we could try a different
distribution for $N_i$ such as the negative binomial:

```{r}
#| label: negbin
#| echo: true
#| warning: false
#| cache: true
(fm.nb <- pcount(~temp ~ elevation.s+utmN.s, data=grouse.umf, K=50, mixture="NB"))
```
:::



## Goodness-of-fit in unmarked
  
::: {style='font-size: 80%;'}
Or the zero-inflated Poisson:

```{r}
#| label: zip
#| warning: false
#| cache: true
(fm.zip <- pcount(~temp ~ elevation.s+utmN.s, data=grouse.umf, K=50, mixture="ZIP"))
```
:::






## Bayesian fit assessments

<br>

In a Bayesian analysis, we can assess model fit using a
*posterior predictive check* (PPC).

<br>

The PPC compares the posterior distribution of a fit statistic, to
the posterior distribution of a fit statistic computed from
simulated data.
  
<br>
  
The two distributions should look similar if the model fits well.




## BUGS model with fit stats

[Nmix-model-grouse2.jag]("Nmix-model-grouse2.jag")
```{r}
#| label: bugs2
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model-grouse2.jag"))
```



## MCMC

Fit Model 2, corresponding to `fm2`.

```{r}
#| label: bugs-mcmc-gof
#| echo: true
#| message: false
#| cache: true
#| results: 'hide'
jm <- jags.basic(data=jags.data, inits=jags.inits,
                 parameters.to.save=c(jags.pars, "SSE", "SSE.new"), 
                 model.file="Nmix-model-grouse2.jag",
                 n.chains=3, n.adapt=100, n.burnin=0,
                 n.iter=2000, parallel=TRUE)
```




## Bayesian goodness-of-fit

```{r}
#| label: bugs-mcmc-gof-plot
#| fig-width: 8
#| fig-align: 'center'
#| out-width: '55%'
plot(as.matrix(jm[,c("SSE", "SSE.new")]))
abline(a=0, b=1, col="red")
```
  
  
The distributions of `SSE` and `SSE.new` are similar,
indicating that the model fits the data well.









## Assignment

::: {style='font-size: 70%;'}
Create a self-contained R script or Rmarkdown file
to do the following:
  
- Simulate a dataset from the following $N$-mixture model:
  * $N_i \sim \mathrm{ZIPoisson}(\lambda=5, \psi=0.5)$
  * $y_{ij} \sim \mathrm{Binomial}(N_i, p=0.3)$
  * nSites=200, nOccasions=4
      
- Fit a *mis-specified model* using unmarked and `JAGS` with
  $N_i \sim \mathrm{Pois}(\lambda)$ instead of
  $N_i \sim \mathrm{ZIPois}(\lambda, \psi)$.
- Assess goodness-of-fit using the parametric bootstrap and
      the Bayesian posterior predictive check.
- Fit the *correct model* using 'unmarked' and 'JAGS'
      and compare the fit to the mis-specified models.
- Were the fit assessments methods able to identify the
      mis-specified model?

<center>
Upload your `.R` or `.Rmd` file to ELC by 5:00pm on Tuesday. 
</center>

:::


