---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Model selection and model fit'
    slide-number: true
lang: en
---




Lecture 6 -- Binomial $N$-mixture models: <br>
model selection <!-- % , spatial prediction, -->
and goodness-of-fit <br>
  
  
  WILD(FISH) 8390 <br>
  Inference for Models of Fish and Wildlife Population Dynamics <br>


# Overview {visibility="hidden"}

## Overview



# Model selection {visibility="hidden"}



## Model selection

In scientific contexts, we want models that describe natural
processes and allow us to evaluate hypotheses. 
  
  
Models should be predictive, but they shouldn't be crafted with the
sole goal of prediction in mind. 
  
  
Scientists typically don't care about prediction if the model doesn't
help us learn about the processes that gave rise to the data. 
  
Nonetheless, predictive performance is often the best way to compare
models and avoid underfitting and overfitting. 
  
Make sure the models being compared were motivated by clear
hypotheses. 



## Model selection

Prediction accuracy increases with model complexity up to a point,
until overfitting kicks in. 
  
The best way to determine if a model is too simplistic or too complex
is to compare predictions to new observations. 
  
However, we rarely have the resources to collect additional data for
the sake of evaluating predictive performance.

A cheaper (albeit less desirable) alternative is to use
cross-validation:
  
- Split data into K partitions
- Fit model to the data in K-1 partitions
- Predict the holdout partition
  
  
Information criteria like AIC and WAIC yield similar predictive rankings as cross-validation, but with less computation.




## Overfitting

7 regression models with 5 covariates.
  
Only `x2` is in the data generating model.
```{r, echo=2:14}
#| label: overfit
#| fig-width: 8
#| fig-show: 'hide'
#| output-location: 'slide'
set.seed(43340)
n <- 50
dat <- data.frame(x1=rnorm(n), x2=rnorm(n), x3=rnorm(n),
                  x4=rnorm(n), x5=rnorm(n))
dat$y <- rnorm(n, mean = -1 + 2*dat$x2, sd = 2) # Data generating model
fm1 <- glm(y~1, gaussian, dat)
fm2 <- glm(y~x1, gaussian, dat)
fm3 <- glm(y~x2, gaussian, dat)                 # Data generating model
fm4 <- glm(y~x1+x2, gaussian, dat)
fm5 <- glm(y~x1+x2+x3, gaussian, dat)
fm6 <- glm(y~x1+x2+x3+x4, gaussian, dat)
fm7 <- glm(y~x1+x2+x3+x4+x5, gaussian, dat)

library(boot)  ## For 'cv.glm'
prediction_error <- c(fm1=cv.glm(dat, fm1)$delta[1],
  fm2=cv.glm(dat, fm2)$delta[1], fm3=cv.glm(dat, fm3)$delta[1],
  fm4=cv.glm(dat, fm4)$delta[1], fm5=cv.glm(dat, fm5)$delta[1],
  fm6=cv.glm(dat, fm6)$delta[1], fm7=cv.glm(dat, fm7)$delta[1])

xbp <- barplot(prediction_error, xlab="Model", ylab="Predication error", cex.lab=1.3)
text(xbp[1], 0.3, as.character(fm1$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[2], 0.3, as.character(fm2$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[3], 0.3, as.character(fm3$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[4], 0.3, as.character(fm4$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[5], 0.3, as.character(fm5$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[6], 0.3, as.character(fm6$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[7], 0.3, as.character(fm7$call)[2], srt=90, pos=4, cex=1.0)
text(xbp[3], prediction_error[3]+0.1, "Data generating model", srt=90,
     pos=4, cex=1.3)
segments(0.3, 9.5, 2.2, 9.5, xpd=TRUE)
text(1.2, 10, "Under-fit", xpd=TRUE, cex=1.3)
segments(4, 4.5, 8.2, 4.5, xpd=TRUE)
text(6, 5, "Over-fit", xpd=TRUE, cex=1.3)
```





## Model selection with AIC
  
Most information criteria (IC) take on a form like this:
$$
  \mathrm{IC} = \mathrm{fit} + \mathrm{penalty}
$$

where `fit` describes the *in-sample* predictive accuracy and
`penalty` describes model complexity. 
  
In the case of Akaike's "An Information Criterion", which approximates
leave-one-out cross validation, `fit` is described by the likelihood
and the `penalty` is determined by the number of parameters: 
$$
  \mathrm{AIC} = -2 \log(L) + 2P 
$$
where $L$ is the likelihood evaluated at the MLE and $P$ is the number
of parameters. 
   
The lower the AIC, the better the \alert{out-of-sample} predictive
performance. 



% \bgroup
% \let\oldfootnoterule\footnoterule
% \def\footnoterule{\only<1->\oldfootnoterule}
% 
%   ## Model selection with AIC
% %  
%   For an $N$-mixture model, the `integrated likelihood' is computed by
%   marginalizing (summing over all possible values of) $N_i$:
%   \[
%      L = \prod_{i=1}^M \sum_{N^*_i=\max(y_i)}^{K\approx \infty}
%      \left\{\prod_{j=1}^J p(y_{ij}|N^*_i,p)\right\}p(N^*_i|\lambda)
%   \]
% %  
%   where $p(y_{ij}|N_i,p)$ is the binomial probability
%   density\footnote<1->{\scriptsize For discrete random variables, 
%     ``mass'' is often used instead of ``density''.} function 
%   (pdf), and $p(N_i|\lambda)$ is the Poisson (or similar) pdf for
%   local abundance. 
% 
% \egroup



% \bgroup
% \let\oldfootnoterule\footnoterule
% \def\footnoterule{\only<2->\oldfootnoterule}


## Model selection in unmarked
  
Import the grouse data
  
```{r}
#| label: grouse-in
#| cache: false
#| results: 'hide'
#| message: false
library(unmarked)
grouse.data <- read.csv("grouse_data_Nmix.csv", row.names=1)
grouse.umf <- unmarkedFramePCount(
    y=grouse.data[,paste0("grouse",1:3)],
    siteCovs=grouse.data[,c("utmE","utmN","elevation")],
    obsCovs=list(temp=grouse.data[,paste0("Temperature.",1:3)]))
```


Standardize the covariates
  
```{r}
#| label: grouse-stand
#| cache: false
## scale() only works if all the covariates are continuous  
site.covs.s <- scale(siteCovs(grouse.umf))
colnames(site.covs.s) <- paste0(colnames(site.covs.s), ".s")
siteCovs(grouse.umf) <- cbind(siteCovs(grouse.umf), site.covs.s)
obsCovs(grouse.umf) <- scale(obsCovs(grouse.umf))
```





## Model selection in `unmarked'
  
Fit some models
  
```{r}
#| label: grouse-mods
#| warning: false
#| cache: true
fm1 <- pcount(~temp ~ elevation.s+utmE.s+utmN.s, grouse.umf, K=50) 
fm2 <- pcount(~temp ~ elevation.s+utmN.s, grouse.umf, K=50)
fm3 <- pcount(~temp ~ elevation.s, grouse.umf, K=50)
fm4 <- pcount(~1 ~ elevation.s+utmN.s, grouse.umf, K=50)
fm5 <- pcount(~1 ~ elevation.s, grouse.umf, K=50)
fm6 <- pcount(~1 ~ 1, grouse.umf, K=50)
```


Put models in a special type of list
  
```{r}
#| label: grouse-fitlist
#| warning: false
grouse.models <- fitList('lam(elev+utmE+utmN)p(temp)'=fm1,
                         'lam(elev+utmN)p(temp)'=fm2,
                         'lam(elev)p(ptemp)'=fm3,
                         'lam(elev+utmN)p(.)'=fm4,
                         'lam(elev)p(.)'=fm5,
                         'lam(.)p(.)'=fm6)

```

Uh oh, missing values differ among models.





## Model selection in `unmarked'
 
Replace count data with `NA` where associated covariates are missing:
  
```{r}
#| label: addNA
na.sites <- apply(is.na(site.covs.s), 1, any)
grouse.counts <- getY(grouse.umf)
grouse.counts[na.sites,] <- NA
```
@
  
  
Fit the models again
  
```{r}
#| label: grouse-mods2
#| warning: false
#| cache: true
fm1 <- pcount(~temp ~ elevation.s+utmE.s+utmN.s, grouse.umf, K=50)
fm2 <- pcount(~temp ~ elevation.s+utmN.s, grouse.umf, K=50)
fm3 <- pcount(~temp ~ elevation.s, grouse.umf, K=50)
fm4 <- pcount(~1 ~ elevation.s+utmN.s, grouse.umf, K=50)
fm5 <- pcount(~1 ~ elevation.s, grouse.umf, K=50)
fm6 <- pcount(~1 ~ 1, grouse.umf, K=50)
```


Put models in a `fitList`
  
```{r}
#| label: grouse-fitlist2
#| warning: false
grouse.models <- fitList('lam(elev+utmE+utmN)p(temp)'=fm1,
                         'lam(elev+utmN)p(temp)'=fm2,
                         'lam(elev)p(ptemp)'=fm3,
                         'lam(elev+utmN)p(.)'=fm4,
                         'lam(elev)p(.)'=fm5,
                         'lam(.)p(.)'=fm6)

```





## Model selection in `unmarked'
  
Create AIC table
```{r}
#| label: aic-table
modSel(grouse.models)
```
  
  
We could use the model with the lowest AIC for inference and
prediction.
  
  
  Or, we could model-average the predictions. <br>




## Outline
  
## WAIC

Watanabe's widely applicable information criterion (WAIC) is a
generalization of AIC suitable for Bayesian models. 
  
As before, we have 

$$
  \mathrm{WAIC} = \mathrm{fit} + \mathrm{penalty}
$$
  
The measure of `fit` in this case is the log-pointwise-predictive-density:

$$
   \mathrm{lppd} = \sum_{i=1}^M \log\left(\frac{1}{S} \sum_s p(y_i|\theta_s)\right)
$$
  
  
The penalty is proportional to the posterior variance:
$$
  \mathrm{penalty} = \sum_{i=1}^M \mathrm{Var}(\log p(y_i|\theta_s)))
$$





## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data
jags.data <- list(
    y=grouse.counts,
    elevation=site.covs.s[,"elevation.s"],
    utmE=site.covs.s[,"utmE.s"],
    utmN=site.covs.s[,"utmN.s"],
    temp=as.matrix(grouse.data[,paste0("Temperature.",1:3)]),
    nSites=nrow(grouse.counts),
    nOccasions=ncol(grouse.counts))
jags.data$temp <- (jags.data$temp-mean(jags.data$temp, na.rm=TRUE))/
    sd(jags.data$temp, na.rm=TRUE) # standardize temperature
```


Initial values
  
```{r}
#| label: bugs-inits
jags.inits <- function() {
    list(lambda.intercept=runif(1), alpha0=rnorm(1),
         N=rep(2, jags.data$nSites))
}
```


Parameters to monitor
  
```{r}
#| label: bugs-pars
jags.pars <- c("beta0", "beta1", "beta2", "beta3",
               "alpha0", "alpha1", "totalAbundance",
               "ld.y.dot", "ld.ydot.N")
```




## The BUGS model
  
Notice `modswitch`, which we can use to include/exclude predictors. 

```{r}
#| label: bugs
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model-grouse.jag"))
```

```{r}
#| label: jagsUI
#| include: false
library(jagsUI)
```


## MCMC

Fit Model 1, corresponding to \inr{fm1}. 
  
```{r}
#| label: bugs-mcmc
#| message: false
#| cache: true
#| results: 'hide'
library(jagsUI)
jags.data1 <- jags.data
jags.data1$modswitch <- c(1,1,1,1) ## Include all covariates
jm1 <- jags.basic(data=jags.data1, inits=jags.inits,
                  parameters.to.save=jags.pars, 
                  model.file="Nmix-model-grouse.jag",
                  n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```


Fit Model 2
  
```{r}
#| label: bugs-mcmc2
#| message: false
#| cache: true
#| results: 'hide'
jags.data2 <- jags.data; jags.data2$modswitch <- c(1,0,1,1) 
jm2 <- jags.basic(data=jags.data2, inits=jags.inits,
                  parameters.to.save=jags.pars, 
                  model.file="Nmix-model-grouse.jag",
                  n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```

## MCMC
  
Model 3
  
```{r}
#| label: bugs-mcmc3
#| message: false
#| cache: true
#| results: 'hide'
jags.data3 <- jags.data; jags.data3$modswitch <- c(1,0,0,1)
jm3 <- jags.basic(data=jags.data3, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 4
  
```{r}
#| label: bugs-mcmc4
#| message: false
#| cache: true
#| results: 'hide'
jags.data4 <- jags.data; jags.data4$modswitch <- c(1,0,1,0)
jm4 <- jags.basic(data=jags.data4, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 5
  
```{r}
#| label: bugs-mcmc5
#| message: false
#| cache: true
#| results: 'hide'
jags.data5 <- jags.data; jags.data5$modswitch <- c(1,0,0,0)
jm5 <- jags.basic(data=jags.data5, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```
Model 6
  
```{r}
#| label: bugs-mcmc6
#| message: false
#| cache: true
#| results: 'hide'
jags.data6 <- jags.data; jags.data6$modswitch <- c(0,0,0,0)
jm6 <- jags.basic(data=jags.data6, inits=jags.inits, parameters.to.save=jags.pars,
                  model.file="Nmix-model-grouse.jag", n.chains=3, n.adapt=100, n.burnin=0,
                  n.iter=2000, parallel=TRUE)
```



## WAIC
$$
    \mathrm{WAIC} = -2\left(\mathrm{lppd} - \sum_i \mathrm{Var}(\log p(y_i|\theta))\right)
$$

```{r}
#| label: waic-fn-ld-op
#| include: false
waic <- function(x, focus=c("y", "yN")) {
    vars <- coda::varnames(x)
    if(focus[1]=="y") {
        ld.samples <- as.matrix(x[,grep("ld.y.dot", vars)])
    } else if(focus[1]=="yN") {
        ld.samples <- as.matrix(x[,grep("ld.ydot.N", vars)])
    } else stop("focus should be either 'y' or 'yN'")
    lppd <- sum(log(colMeans(exp(ld.samples))))
    penalty <- sum(apply(ld.samples, 2, var))
    return(-2*lppd + 2*penalty)
}
```

```{r}
#| label: waic-fn
waic <- function(x) {
    ## Parameter names
    vars <- coda::varnames(x)
    ## Extract log-density of y at each site for each post sample
    ld.samples <- as.matrix(x[,grep("ld.y.dot", vars)])
    ## Compute log-pointwise-predictive-density
    lppd <- sum(log(colMeans(exp(ld.samples))))
    ## Compute penalty
    penalty <- sum(apply(ld.samples, 2, var))
    ## Return WAIC
    return(-2*(lppd-penalty))
}
```


## WAIC

WAIC for the six models

::: {.columns}
::: {.column width='50%'}
```{r}
#| label: waic1
#| size: 'scriptsize'
(waic1 <- waic(jm1))
```
```{r}
#| label: waic2
#| size: 'scriptsize'
(waic2 <- waic(jm2))
```
```{r}
#| label: waic3
#| size: 'scriptsize'
(waic3 <- waic(jm3))
```
:::
::: {.column width='50%'}
```{r}
#| label: waic4
#| size: 'scriptsize'
(waic4 <- waic(jm4))
```
```{r}
#| label: waic5
#| size: 'scriptsize'
(waic5 <- waic(jm5))
```
```{r}
#| label: waic6
#| size: 'scriptsize'
(waic6 <- waic(jm6))
```
:::
:::
  
  
In practice, you would discard burn-in and ensure convergence before computing WAIC.


## WAIC

Create a WAIC table:
  
<center>
```{r}
#| label: waic-table
waic.table <- data.frame(WAIC=c(waic1,waic2,waic3,waic4,waic5,waic5))
rownames(waic.table) <- paste("Model", 1:6)
waic.table$delta <- waic.table$WAIC-min(waic.table$WAIC)
waic.table <- waic.table[order(waic.table$WAIC),]
knitr::kable(waic.table, format="latex", digits=2, booktabs=TRUE)
```
</center>






# Goodness-of-fit




## Goodness-of-fit
  
Distributional assumptions determine the expected values
*and the expected variance* of the random variables, including
the data.
  
  
Overdispersion occurs when there is more variance in the data than
expected by the model.
  
  
Goodness-of-fit methods assess overdispersion and underdispersion.
  
  
If the model does not fit the data very well because of
overdispersion, there are several remedial actions:
  
- Scientific approach
  * Figure out why there is unexplained variation
  * Perhaps there were unmeasured covariates or you need a better process model
    
- Statistical approach
  * Soak up variation with random effects
  * Use a different distribution (we'll start here)
    
  
## Changing the distribution for $N$
  
  Standard Poisson-binomial $N$-mixture model (without covariates):
$$
    N_i \sim \mathrm{Poisson}(\lambda) \\
    y_{ij} \sim \mathrm{Binomial}(N_i, p)
$$
  
We can replace the Poisson distribution with other distributions
that allow for greater variance in $N$.  Two common examples
are the negative  binomial:
$$
    N_i \sim \mathrm{NegBin}(\lambda_i, \kappa)
$$
where $\lambda_i$ is the expected value of $N$ and $\kappa$ is the
dispersion parameter^[There are several other
parameterizations of the negative binomial].
  
Another option is the zero-inflated Poisson:
::: columns
::: {.column width='50%'}
$$
  N_i \sim \mathrm{Poisson}(\lambda_i z_i) \\
  z_i \sim \mathrm{Bern}(\psi) <br>
$$
:::
::: {.column width='50%'}
::: {.column width='50%'}
$$
  N_i \sim \mathrm{ZIPoisson}(\lambda_i,\psi)
$$
:::
:::
where $\psi$ is the expected proportion of sites with excess zeros. 


## Simulating data from ZIP model

Zero-inflated Poisson model
::: columns
::: {.column width='50%'}
$$
  N_i \sim \mathrm{Poisson}(\lambda_i z_i) <br>
  z_i \sim \mathrm{Bern}(\psi) <br>
$$
:::
::: {.column width='50%'}
$$
  N_i \sim \mathrm{ZIPoisson}(\lambda_i,\psi)
:::
:::


R code
```{r}
#| label: zip-sim

nSites <- 20
nVisits <- 3
psi <- 0.5   # Proportion of extra-Poisson zeros
z <- rbinom(n=nSites, size=1, prob=psi) # Extra zeros
lam <- 5     # expected count when z=1
N <- rpois(n=nSites, lambda=lam*z)      # abundance at each site
p <- 0.5     # detection prob
y <- matrix(NA, nSites, nVisits)
for(i in 1:nSites) {
    y[i,] <- rbinom(n=nVisits, size=N[i], prob=p) # count data
}
```



## Goodness-of-fit in unmarked
  
One of the simplest and best ways of assessing goodness-of-fit is to
look at the residuals.
```{r}
#| label: resid2
#| fig-width: 9
#| out-width: '60%'
#| fig-align: 'center'
#| fig-keep: 'first'
plot(fm2)
```


No sign of obvious outliers or other problems.


## Goodness-of-fit in unmarked

The parametric bootstrap can be used to assess model fit:
- Simulate a dataset from the fitted model
- Fit the model to the new dataset
- Compute a fit statistic
- Repeat steps 1-3 several hundred/thousand times
- Compare the distribution of the *expected* fit statistic to
the fit statistic associated with the actual data.
  
The fit statistic associated with the actual data should not be in
an extreme quantile of the distribution of the expected values.
  
Assuming a type I error rate of 0.05, the $p$-value should be
$>0.05$ if the model fits the data well, which is the null hypothesis.


## Goodness-of-fit in unmarked

```{r}
#| label: parboot
#| warning: false
#| cache: true
#| fig-width: 9
#| out-width: '70%'
#| fig-align: 'center'
fitstat <- function(fm) { ## Sum of squared-residuals
    return(c(SSE=sum(residuals(fm)^2, na.rm=TRUE))) 
}
pb <- parboot(fm2, statistic=fitstat, nsim=200, ncores=3); plot(pb)
```



## Goodness-of-fit in unmarked
```{r}
#| label: parboot2
pb
```

These results suggest that model `fm2` fits the data well.



## Goodness-of-fit in unmarked
  
If the model didn't fit the data well, we could try a different
distribution for $N_i$ such as the negative binomial:

```{r}
#| label: negbin
#| warning: false
#| cache: true
(fm.nb <- pcount(~temp ~ elevation.s+utmN.s, data=grouse.umf, K=50, mixture="NB"))
```



## Goodness-of-fit in unmarked
  
Or the zero-inflated Poisson:

```{r}
#| label: zip
#| warning: false
#| cache: true
(fm.zip <- pcount(~temp ~ elevation.s+utmN.s, data=grouse.umf, K=50, mixture="ZIP"))
```






## Bayesian fit assessments

In a Bayesian analysis, we can assess model fit using a
*posterior predictive check* (PPC).
  
  
The PPC compares the posterior distribution of a fit statistic, to
the posterior distribution of a fit statistic computed from
simulated data.
  
  
The two distributions should look similar if the model fits well.




## BUGS model with fit stats

```{r}
#| label: bugs2
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model-grouse2.jag"))
```



## MCMC
Fit Model 2, corresponding to \inr{fm2}. 


```{r}
#| label: bugs-mcmc-gof
#| message: false
#| cache: true
#| results: 'hide'
jm <- jags.basic(data=jags.data, inits=jags.inits,
                 parameters.to.save=c(jags.pars, "SSE", "SSE.new"), 
                 model.file="Nmix-model-grouse2.jag",
                 n.chains=3, n.adapt=100, n.burnin=0,
                 n.iter=2000, parallel=TRUE)
```




## Bayesian goodness-of-fit

```{r}
#| label: bugs-mcmc-gof-plot
#| fig-width: 8
#| fig-align: 'center'
#| out-width: '55%'
plot(as.matrix(jm[,c("SSE", "SSE.new")]))
abline(a=0, b=1, col="red")
```
  
  
The distributions of `SSE` and `SSE.new` are similar,
indicating that the model fits the data well.









## Assignment

Create a self-contained R script or Rmarkdown file
to do the following:
  
- Simulate a dataset from the following $N$-mixture model:
  * $N_i \sim \mathrm{ZIPoisson}(\lambda=5, \psi=0.5)$
  * $y_{ij} \sim \mathrm{Binomial}(N_i, p=0.3)$
  * nSites=200, nOccasions=4
      
- Fit a \alert{mis-specified model} using unmarked and `JAGS' with
      $N_i \sim \mathrm{Pois}(\lambda)$ instead of
      $N_i \sim \mathrm{ZIPois}(\lambda, \psi)$.
- Assess goodness-of-fit using the parametric bootstrap and
      the Bayesian posterior predictive check.
- Fit the *correct model* using 'unmarked' and 'JAGS'
      and compare the fit to the mis-specified models.
- Were the fit assessments methods able to identify the
      mis-specified model?
  
  
Upload your `.R` or `.Rmd` file to ELC by 5:00pm on Tuesday. 







