---
title: "Movement models"
author: "Richard Chandler"
date: "2025-11-18"
format: revealjs
description: |
  Random walks
engine: knitr
---

::: {.hidden}
$$
 \def\RR{{\bf R}}
 \def\bold#1{{\bf #1}}
 \def\but{{\boldsymbol u_t}}
 \def\butp{{\boldsymbol u_{t+1}}}
$$
:::


# Overview

## Movement models

::: {.incremental}

- Random walks
  * Simple random walk
  * Correlated random walk
  * Biased random walk
  * PDE

:::

. . .

- Point process models
  * Step-selection models


# Random walks

## Simple random walk

$$
  {\bf u_{t+1}} = \but + {\bf \epsilon}
$$

where

$$
  {\boldsymbol \epsilon} = \text{Norm}(0, \sigma^2 {\bf I})
$$

## Simple random walk simulation


:::: {.columns}

::: {.column width="50%"}

```{r}
#| label: rw-simple-code
#| echo: TRUE
#| attr-source: 'style="font-size: 0.7em;"'

T <- 50
u <- matrix(NA, nrow=T, ncol=2)
u[1,] <- c(0.5,0.5)
for(t in 2:T) {
  u[t,] <- u[t-1,] + rnorm(n=2, mean=c(0,0), sd=0.05)
  ## u[t,] <- rnorm(n=2, mean=u[t-1,], sd=0.1)  # Equivalent to line above
}
```

:::

::: {.column width="50%"}

```{r}
#| label: rw-simple-plot
#| echo: TRUE
#| out.height: 800px
#| out.width: 800px
#| fig.align: center
#| attr-source: 'style="font-size: 0.7em;"'
plot(u, asp=1, xlim=c(0,1), ylim=c(0,1))
arrows(u[1:49,1], u[1:49,2],
       u[2:50,1], u[2:50,2], length=0.1, col="blue")

```

:::

## Biased random walk

\begin{equation}
  {\boldsymbol u_{t+1}} = {\bf u_t} + {\bf B}({\bf u_{t}}-{\bf s}) + {\bf \epsilon}
  \label{eq:brwalk}
\end{equation}



## Correlated random walk

$$
\begin{equation}
  {\bf R} = \gamma
  \begin{pmatrix}
    \cos(\theta) & -\sin(\theta) \\
    \sin(\theta) & \cos(\theta) 
  \end{pmatrix}
  \label{eq:crw-rotmat}
\end{equation}
$$
where $\theta \in [-\pi,\pi]$ is the expected turn angle, and
$\gamma \in [0,1]$ dampens the persistence. 

