---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Spatial capture-recapture I'
    slide-number: true
    echo: true
lang: en
---

  
## {content-align="center" style="font-size: 150%;"}

<br>

<center>

**Spatial capture-recapture for <br> closed populations**

<br>

::: {style='font-size: 60%;'}
  WILD(FISH) 8390 <br>
  Inference for Models of Fish and Wildlife Population Dynamics <br>
:::
  
</center>




# Overview {visibility="hidden"}



## [SCR overview -- Two motivations]{style='font-size: 85%;'}

::: {style='font-size: 65%;'}

1. Improved inference
    - Non-spatial models don't account for important sources of variation in $p$.
        * Distance between animals and traps
        * Trap-specific covariates
    - SCR makes it possible to estimate *density*, not just $N$ in an unknown region.

::: {.fragment}
2. Improved science
    - We can ask new questions, such as:
        * What influences spatial variation in density?
        * How do survival and recruitment vary in space and time?
        * How does movement influence density and detectability?
:::

::: {.fragment}
<center>
**Rather than think of SCR as a new estimation tool, you can think of it as an individual-based framework for inference on spatial population dynamics.**
</center>
:::

:::


## SCR data
  
::: {style='font-size: 70%;'}
SCR capture histories can be organized as a 3D array where $y_{ijk}$ indicates if individual $i=1,\dots,n$, was captured in trap $j=1,\dots,J$ on occasion $k=1,\dots,K$.

<!-- Here's an example of a "flattened array" with $n=4$ animals captured at $J=3$ traps on $K=2$ occasions. -->
  
::: {style='font-size: 60%;'}

+----------------+----------------------------------+
|                | Occasion                         |
+:===============+:==:+:==:+:==:+:==:+:==:+:==:+:==:+
|                | Trap         |    | Trap         |
+----------------+--------------+----+--------------+
|                | 1            |    | 2            |
+----------------+----+----+----+----+----+----+----+
| Individual     | 1  | 2  | 3  |    | 1  | 2  | 3  |
+----------------+----+----+----+----+----+----+----+
| 1              | 0  | 0  | 0  |    | 1  | 0  | 1  |
+----------------+----+----+----+----+----+----+----+
| 2              | 1  | 1  | 1  |    | 0  | 1  | 1  |
+----------------+----+----+----+----+----+----+----+
| 3              | 0  | 1  | 0  |    | 0  | 0  | 0  |
+----------------+----+----+----+----+----+----+----+
| 4              | 0  | 1  | 1  |    | 1  | 0  | 0  |
+----------------+----+----+----+----+----+----+----+

:::

<br>

::: {.fragment}
Because we know the coordinates of the traps, we also know when and where each individual was detected.
:::
  
  
::: {.fragment}
This spatial information has been available all along, but it wasn't utilized to estimate density until Efford (2004, Oikos).
:::

:::



## [Closed population model (*N* known hypothetically) ]{style="font-size: 60%;"}

::: {style='font-size: 50%;'}
State model (a spatial point process model) 
$$
    \lambda({\bf s}) = \exp(\beta_0 + \beta_1 w_1({\bf s}) + \beta_2 w_2({\bf s}) + \cdots) \\
    \Lambda = \int_{\mathcal{S}} \lambda({\bf s}) \; \mathrm{d}{\bf s} \\
    N \sim \mathrm{Pois}(\Lambda) \\
    {\bf s}_i \propto \lambda({\bf s}) \;\; \mathrm{for}\; i=1,\dots,N 
$$

Observation model (supposing $N$ was known)
$$
    p_{ij} = g_0\exp(-\|{\bf s}_i - {\bf x}_j\|^2/(2\sigma^2))  \;\; \mathrm{for}\, j=1,\dots,J \\
    y_{ijk} \sim \mathrm{Bernoulli}(p_{ij})
$$

:::

::: {style='font-size: 50%;'}
  $\lambda({\bf s})$ -- "Intensity function" describing expected density of individuals at location $\bf s$ <br> 
  $\Lambda$ -- Expected number of individuals in the spatial region $\mathcal{S}$ <br>
  $N$ -- Realized number of individuals (ie, population size) <br>
  ${\bf s}_i$ -- Location of the $i$th activity center <br>
  ${\bf x}_j$ -- Location of trap $j$ <br>
  $\|{\bf s}_i - {\bf x}_j\|$ -- Euclidean distance between ${\bf s}_i$ and ${\bf x}_j$ <br>
  $g_0$ -- Capture probability when distance between activity centers
  and traps is 0 <br>
  $\sigma$ -- Scale parameter of encounter function <br>
  $p_{ij}$ -- Capture probability <br>
  $y_{ijk}$ -- Spatial capture histories <br>
:::






## [Closed population model (*N* known hypothetically)]{style="font-size: 60%;"}

::: {style='font-size: 80%;'}
Many variations and extensions are possible for both the state and observation models, but today we will focus on the simplest case with no spatial variation in the expected value of density.

::: {style='font-size: 80%;'}
$$
    N \sim \mathrm{Pois}(\lambda) \\
    {\bf s}_i \sim \mathrm{Unif}(\cal S) \\
    p_{ij} = g_0\exp\left(\frac{-\|{\bf s}_i - {\bf x}_j\|^2}{2\sigma^2}\right) \\
    y_{ijk} \sim \mathrm{Bernoulli}(p_{ij})
$$
::: 

<br>

::: {.fragment}
Just remember, this formulation isn't exactly correct because we can't observe the "all zero" encounter histories (and thus we don't know $N$).
:::


:::



## {.smaller}

<br>

::: {.r-fit-text}
[Shiny App](https://rbchan-scr-cap-prob.share.connect.posit.cloud/)
:::

<center>
[https://rbchan-scr-cap-prob.share.connect.posit.cloud/](https://rbchan-scr-cap-prob.share.connect.posit.cloud/)
</center>




# Simulation


## Homogeneous Poisson point process


::: {style='font-size: 80%;'}

```{r, echo=-1}
#| label: ppp1
#| size: 'footnotesize'
#| fig-height: 7
#| fig-width: 7
#| out-width: "45%"
#| fig-align: "center"
set.seed(439)
lambda1 <- 25; A <- 1     ## lambda1=density, A=area
N <- rpois(1, lambda1*A)  
s <- cbind(runif(N, 0, 1), runif(N, 0, 1))
plot(s, pch=16, col="blue", xlab="Easting", ylab="Northing",
     xlim=c(0,1), ylim=c(0,1), cex.lab=1.5, asp=1)
```

:::

```{r}
#| include: false
#| eval: false
#| purl: false
## Simulate covariate from normal distribution with exponential covariance
delta <- 0.01
grid0 <- seq(0+delta/2, 1-delta/2, delta)
grid <- cbind(rep(grid0, each=length(grid0)),
              rep(grid0, times=length(grid0)))
dist.grid <- as.matrix(dist(grid))
V <- exp(-dist.grid*10)
set.seed(4984)
w <- t(chol(V)) %*% rnorm(nrow(V))

elevation <- 1000 + w*100

library(raster)
elevation.r <- rasterFromXYZ(cbind(grid[,1], grid[,2], elevation))
plot(elevation.r)
writeRaster(elevation.r, filename="elevation.tif", overwrite=TRUE)

## library(lattice)

## levelplot(w ~ grid[,1]+grid[,2], aspect="iso", at=seq(min(w), max(w), len=100))

## lambda <- exp(-7 + 1.5*w)
## (Lambda <- sum(lambda*delta^2*1e4))
## (N <- rpois(1, Lambda))

## levelplot(lambda ~ grid[,1]+grid[,2], aspect="iso",
##           at=seq(min(lambda), max(lambda), len=100),
##           col.regions=topo.colors(100))

## pi <- lambda/sum(lambda)

## s <- grid[sample(nrow(grid), size=N, replace=TRUE, prob=pi),]

## levelplot(lambda ~ grid[,1]+grid[,2], aspect="iso",
##           panel=function(...) {
##               panel.levelplot(...)
##               lpoints(s, pch=16, col="blue")
##           }, 
##           at=seq(min(lambda), max(lambda), len=100),
##           col.regions=terrain.colors(100))
```







## Traps

::: {style='font-size: 80%;'}

```{r, echo=-(2:4)}
#| label: traps1
#| fig-height: 7
#| fig-width: 7
#| out-width: "45%"
#| fig-align: "center"
x <- cbind(rep(seq(0.15, 0.85, by=0.1), each=8),
           rep(seq(0.15, 0.85, by=0.1), times=8))  ## Trap locations
##plot(lambda, col=terrain.colors(100),
##     main="Density surface with activity centers and traps")
##points(s, pch=16, col="blue") ## Activity center locations
plot(s, pch=16, col="blue", xlab="Easting", ylab="Northing", asp=1,
     main="Activity centers and traps")
points(x, pch=3)              ## Trap locations
```
:::



## [Distance between traps and activity centers]{style="font-size: 80%;"}

::: {style='font-size: 80%;'}
Compute distances between activity centers (${\bf s}_1, \dots, {\bf s}_N$) and traps (${\bf x}_1, \dots, {\bf x}_J$).
```{r}
#| label: dist1
J <- nrow(x)                 ## nTraps
dist.sx <- matrix(NA, N, J)  
for(i in 1:N) {
    dist.sx[i,] <- sqrt((s[i,1]-x[,1])^2 + (s[i,2]-x[,2])^2)
}
```

Look at distances between first 4 individuals and first 5 traps.

```{r}
#| label: dist2
round(dist.sx[1:4,1:5], digits=2)
```
:::





## Capture probability

::: {style='font-size: 80%;'}
Compute capture probability

```{r}
#| label: p1
g0 <- 0.2
sigma <- 0.05
p <- g0*exp(-dist.sx^2/(2*sigma^2))
```

<br>

Look at capture probs for first 4 individuals and first 5 traps.

```{r}
#| label: p2
round(p[1:4,1:5], digits=3)
```
:::



## Capture histories

::: {style='font-size: 80%;'}
Simulate capture histories for all $N$ individuals

```{r}
#| label: y1
K <- 5                          # nOccasions
y.all <- array(NA, c(N, J, K))
for(i in 1:N) {
    for(j in 1:J) {
        y.all[i,j,] <- rbinom(K, 1, prob=p[i,j])
    }
}
```

Discard individuals not captured

```{r}
#| label: y2
captured <- rowSums(y.all)>0
y <- y.all[captured,,]
```

Capture histories for first 2 individuals and first 5 traps on first occasion.

```{r}
#| label: y3
y[1:2,1:5,1]
```
:::


## Summary stats

::: {style='font-size: 80%;'}
Individuals captured


```{r}
#| label: n-y
(n <- nrow(y))
```
 
Capture frequencies
    

```{r}
#| label: cap-freq
y.tilde <- rowSums(y)
table(y.tilde)
```


Spatial recaptures
    
```{r}
#| label: sp-cap-freq
y.nok <- apply(y, c(1, 2), sum)
y.nojk <- apply(y.nok>0, 1, sum)
table(y.nojk)
```

:::



## Spider plot

::: {style='font-size: 80%;'}
```{r}
#| label: spider
#| fig-width: 7
#| fig-height: 7
#| out-width: '55%'
#| fig-align: 'center'
#| echo: false
#| include: true
## plot(lambda, col=terrain.colors(100),
##      main="Density surface, activity centers, traps, and capture locs")
plot(0, xlim=c(0,1), ylim=c(0,1), asp=1, xlab="", ylab="",
     main="Activity centers, traps, and capture locs")
s.cap <- s[captured,]
for(i in 1:n) {
    traps.i <- which(rowSums(y[i,,])>0)
    for(j in 1:length(traps.i)) {
        segments(s.cap[i,1], s.cap[i,2],
                 x[traps.i[j],1], x[traps.i[j],2], col=gray(0.3))
    }
}
points(s[captured,], pch=16, col="blue") ## Activity center locations
points(s[!captured,], pch=1, col="blue") ## Activity center locations
points(x, pch=3)                         ## Trap locations
```
:::




# Likelihood


## Software options

-  Program DENSITY
  * Windows program with GUI
  
-  R package 'secr'
  * The oldest R package with the most options
  
- R package 'oSCR'
  * A newer R package with similar functionality
  




```{r}
#| label: write
#| include: false
#| results: "hide"
#| purl: false
ch.out <- data.frame(session=1,
                     individual=rep(slice.index(y, 1), y),
                     occasion=rep(slice.index(y, 3), y),
                     trap=rep(slice.index(y, 2), y))
write.table(ch.out, file="encounter_data_file.csv",
            row.names=FALSE, col.names=FALSE, sep=",")
traps.out <- data.frame(trap=1:nrow(x), x*1000)
write.table(traps.out, file="trap_data_file.csv",
            row.names=FALSE, col.names=FALSE, sep=",")
library(secr)
```


## R package 'secr'

::: {style='font-size: 80%;'}
Import data from two text files
```{r}
#| label: secr-in
#| warning: false
library(secr)  
sch <- read.capthist(captfile="encounter_data_file.csv",
                     trapfile="trap_data_file.csv",
                     detector="proximity", fmt="trapID")
summary(sch)
```
:::



## R package 'secr'

```{r}
#| label: secr-plot
#| out-width: "70%"
#| fig-align: "center"
plot(sch)
```



## R package 'secr'

::: {style='font-size: 80%;'}
Fit the SCR equivalent of Model $M_0$
```{r}
#| label: secr-M0
#| cache: true
#| warning: false
fm.M0 <- secr.fit(sch, model=list(D=~1, g0=~1, sigma=~1),
    details = list(fastproximity = FALSE), ## For AIC comparisons
    buffer=150, trace=FALSE, ncores=3)
coef(fm.M0)
```


Estimates on original scale
```{r}
#| label: secr-M0-real
#| cache: true
predict(fm.M0)
```
<br>

<center>
`D` is density ($\lambda$, animals/ha), `g0` is $g_0$, and `sigma` is 
$\sigma$.
</center>

:::


## R package 'secr'

::: {style='font-size: 80%;'}
Fit the SCR equivalent of Model $M_t$

```{r}
#| label: secr-Mt
#| cache: true
#| warning: false
fm.Mt <- secr.fit(sch, model=list(D=~1, g0=~t, sigma=~1),
                  buffer=150, trace=FALSE, ncores=3)
coef(fm.Mt)
```


Estimates on original scale
```{r}
#| label: secr-Mt-real
#| cache: true
predict(fm.Mt)
```
:::


## R package 'secr'

::: {style='font-size: 80%;'}
Fit the SCR equivalent of Model $M_b$
```{r}
#| label: secr-Mb
#| cache: true
#| warning: false
fm.Mb <- secr.fit(sch, model=list(D=~1, g0=~b, sigma=~1),
                  buffer=150, trace=FALSE, ncores=3)
coef(fm.Mb)
```


Estimates on original scale
```{r}
#| label: secr-Mb-real
#| cache: true
predict(fm.Mb)
```
:::


## R package 'secr'

::: {style='font-size: 75%;'}
What about $N$? `E.N` is the expected value of $N$. `R.N` is the realized value of $N$. 
:::


```{r}
#| label: regionN-M0
region.N(fm.M0)
```
  
```{r}
#| label: regionN-Mt
region.N(fm.Mt)
```
  
```{r}
#| label: regionN-Mb
region.N(fm.Mb)
```


## R package 'secr'
  
Was `buffer` big enough?

::: {style='font-size: 75%;'}
```{r}
#| label: buffer1
#| warning: false
#| cache: true
predict(update(fm.M0, buffer=100))[1,]
```


```{r}
#| label: buffer2
#| warning: false
#| cache: true
predict(update(fm.M0, buffer=150))[1,]
```


```{r}
#| label: buffer3
#| warning: false
#| cache: true
predict(update(fm.M0, buffer=200))[1,]
```


```{r}
#| label: buffer4
#| warning: false
#| cache: true
predict(update(fm.M0, buffer=250))[1,]
```


```{r}
#| label: buffer5
#| warning: false
#| cache: true
predict(update(fm.M0, buffer=300))[1,]
```

:::




## R package 'secr'

AIC
```{r}
#| label: aic
#| cache: true
AIC(fm.M0, fm.Mt, fm.Mb)
```



# Data augmentation



## Data augmentation model

The DA version of a basic SCR model is:

::: {style='font-size: 80%;'}
$$
    {\bf s}_i \sim \mathrm{Unif}(\mathcal{S}) \\
    z_i \sim \mathrm{Bern}(\psi) \\
    p_{ij} = g_0\exp\left(\frac{-\|{\bf s}_i-{\bf x}_j\|^2}{2\sigma^2}\right) \\
    y_{ijk} \sim \mathrm{Bern}(z_i \times p_{ij}) \\
    N=\sum_{i=1}^M z_i
$$
:::





## Model SCR~0~ -- data augmentation

<center>
[SCR0.jag](SCR0.jag)
</center>


```{r}
#| label: bugs-SC0-R
#| eval: false
#| include: false
#| echo: false
writeLines(readLines("SCR0.jag"))
```
```{r}
#| label: bugs-SC0
#| comment: ''
#| background: 'beige'
#| echo: false
writeLines(readLines("SCR0.jag"))
```



```{r}
#| label: rjags
#| include: false
#| results: "hide"
library(rjags)
```


## Model SCR~0~ -- data augmentation

::: {style='font-size: 75%;'}
Data
  
```{r}
#| label: jd-SCR0-aug
M <- 150
y.aug <- array(0, c(M, J, K))
y.aug[1:nrow(y),,] <- y
jags.data.SCR0 <- list(y=y.aug, M=M, J=J, K=K, x=x, xlim=c(0,1), ylim=c(0,1))
```


Inits and parameters
  
```{r, echo=1:2}
#| label: ji-M0-aug
#| message: false
#| report: false
ji.SCR0 <- function() {
  si <- cbind(runif(M), runif(M))
  for(i in 1:nrow(y)) {
    traps_with_dets_i <- rowSums(y[i,,]>0)
    si[i,] <- colMeans(x[traps_with_dets_i,,drop=FALSE])
  }
  list(z=rep(1,M), psi=runif(1), s=si, g0=runif(1), sigma=runif(1, 0.05, 0.1)) }
jp.SCR0 <- c("g0", "sigma", "EN", "N")
library(jagsUI)
```


MCMC
  
```{r}
#| label: mcmc-M0-aug
#| results: 'hide'
#| cache: true
jags.post.SCR0 <- jags.basic(data=jags.data.SCR0, inits=ji.SCR0,
                             parameters.to.save=jp.SCR0, model.file="SCR0.jag",
                             n.chains=3, n.adapt=100, n.burnin=0,
                             n.iter=2000, parallel=TRUE)
```
:::


## Posterior summaries

```{r}
#| label: summary-mcmc-SCR0
summary(jags.post.SCR0)
```

## Traceplots and density plots

```{r}
#| label: plot-mcmc-SCR0
#| fig-height: 7
#| fig-width: 9
#| out-width: "70%"
#| fig-align: 'center'
plot(jags.post.SCR0[,jp.SCR0])
```

## [Easy tricks to speed up Bayesian inference]{style="font-size: 80%;"}
  
- Treat $z_i=1$ as data for first $n$ individuals.
- If there are no occasion-specific covariates, collapse data and use binomial instead of Bernoulli distribution.
- Use a single zero for each augmented individual, instead of an array of zeros. Then compute probability of detecting an individual at least once.
  

## Model SCR~0~ -- data augmentation

```{r}
#| label: bugs-SC0-faster-R
#| eval: false
writeLines(readLines("SCR0-faster.jag"))
```
```{r}
#| label: bugs-SC0-faster
#| comment: ''
#| background: 'beige'
#| echo: false
writeLines(readLines("SCR0-faster.jag"))
```




## Model SCR~0~ -- faster

::: {style='font-size: 80%;'}
Data
  
```{r}
#| label: jd-SCR0-aug-faster
y.tilde <- apply(y, c(1,2), sum)
n <- nrow(y)
jags.data.SCR0.faster <- list(y.tilde=y.tilde, n=n, M=M, J=J,
                              z=c(rep(1, n), rep(NA, M-n)),
                              K=K, zero=rep(0, M), x=x,
                              xlim=c(0,1), ylim=c(0,1))
```


Inits and parameters (same as before)


```{r}
#| label: ji-M0-aug-faster
ji.SCR0.faster <- function() {
    list(z=c(rep(NA, n), rep(0,M-n)), psi=runif(1),
         s=cbind(runif(M), runif(M)),
         g0=runif(1), sigma=runif(1, 0.05, 0.1)) }
```
MCMC
  
```{r}
#| label: mcmc-M0-aug-faster
#| results: 'hide'
#| cache: true
jags.post.SCR0.faster <- jags.basic(data=jags.data.SCR0.faster,
                                    inits=ji.SCR0.faster,
                                    parameters.to.save=jp.SCR0,
                                    model.file="SCR0-faster.jag",
                                    n.chains=3, n.adapt=100, n.burnin=0,
                                    n.iter=2000, parallel=TRUE)
```
:::


## Posterior summaries

```{r}
#| label: summary-mcmc-SCR0-faster
summary(jags.post.SCR0.faster)
```

## Traceplots and density plots

```{r}
#| label: plot-mcmc-SCR0-faster
#| fig-width: 9
#| fig-height: 7
#| out-width: '70%'
#| fig-align: 'center'
plot(jags.post.SCR0.faster[,jp.SCR0])
```



# Summary




## Alternative observation models

Common options include:
  
::: {style='font-size: 70%;'}
  Model                     Detector/Trap   Examples
  ------------------------- --------------- --------------------
  Bernoulli                 Proximity       Hair-snares
  Poisson                   Count           Camera trap
  Multinomial/Categorical   Multi-catch     Mist net, crab pot
  ---                       Single-catch    Sherman trap
:::

<br>

::: {.fragment}
There are other observation models for data from area searches, transects, or acoustic recording units.
:::



## SCR study design

::: {style='font-size: 80%;'}

<br>

SCR involves model-based, rather than design-based, inference (see Ch. 10).

<br>

::: {.fragment}
Random placement of traps is not required, but it's a good idea to randomly sample locations along the environmental gradients you're interested in.
:::
   
<br>

::: {.fragment}
The other two key design considerations are:
- Capture as many individuals as you can (i.e., maximize $n$)
- Obtain as many *spatial* recaptures as possible
:::
  
<br>

::: {.fragment}
There is a tradeoff between these two objectives. Simulation is often the best option for finding the right balance.
:::

:::


## SCR summary

<br>

We assume that variation in $p$ arises from distance between animals and traps.

<br>

We can estimate abundance and model distribution (i.e., spatial variation in density)
   
<br>

Next time, we'll see how to do that using secr and JAGS, and we'll make a bunch of maps.



## Assignment

::: {style='font-size: 80%;'}

<br>

Create a self-contained R script or Rmarkdown file to do the following: 
  
- Import the bear data formatted for 'secr' and fit the null model and a model where $g_0$ varies among the 8 occasions. Use 'secr' to determine what an appropriate buffer should be.
- Import the bear data formatted for JAGS (using the command `load("jags_bear_data.gzip")`) and tabulate the capture frequencies and the number of spatial recapture. Fit the model with occasion-specific $g_0$ in 'JAGS', and compare estimates and 95% CIs for $g_{0}$, $\sigma$, $E(N)$ and $N$ with the results from secr. 
  
Upload your work to ELC by 5:00pm on Tuesday. 
:::

