---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Dog'
    slide-number: true
lang: en
---




  
  { Lecture 3 -- Occupancy models: simulation and fitting}
  {\color{default} \rule{\textwidth}{0.1pt}}
  
  
  WILD(FISH) 8390 \\
%  Estimation of Fish and Wildlife Population Parameters
  Inference for Models of Fish and Wildlife Population Dynamics
  







# Motivation {visibility="hidden"}




## Outline
  

## Motivation

We often want to do logistic regression, but the data are subject to measurement error.
  
  
Specifically, wildlife data are often fraught with false negatives (ie, the species was there but we didn't detect it).
  
  
Ignoring this source of measurement error can result in misleading inferences.
  * Occupancy will be under-estimated
  * Habitat associations may be incorrect
  

## Accounting for detection probability

  The model for the state process is the same as logistic regression:

$$
  \begin{gather*}
    \mathrm{logit}(\psi_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots \\
    z_i \sim \mathrm{Bern}(\psi_i)
  \end{gather*}
$$

Model for the observation process conditional on the state process:

$$
\begin{gather*}
    \mathrm{logit}(p_{ij}) = \alpha_0 + \alpha_1 x_{ij1} + \alpha_2 x_{ij2} + \cdots \\
    y_{ij} \sim \mathrm{Bern}(z_i\times p_{ij})
  \end{gather*}
$$

  
Definitions
  $\psi_i$ -- probability that the species occurs at site $i$ 
  $z_i$ -- binary presence/absence variable at site $i$ 
  $p_{ij}$ -- probability of detecting the species at site $i$ on occasion $j$ 
  $y_{ij}$ -- binary detection/non-detection data




 ## Assumptions

Assumptions of the basic single-season occupancy models:
  
    * []
    * Population closure
    
      * Occupancy state (presence/absence) constant over time
    
    * Statistical independence
    
      * Occupancy state at one site has no impact on occupancy of other sites
      * Detections are independent, conditional on occupancy state 
    
    * Variation in $\psi$ and $p$ described by covariates
  



# Simulation {visibility="hidden"}




## Outline
  

## Simulation

Simplest case with no covariates: 
  

  * Specify the parameters and dimensions
```{r}
#| label: sim-nocov1
psi <- 0.5       ## Occurrence probability
p <- 0.2         ## Detection probability
nSites <- 20
nVisits <- 4
```
  * Now, simulate presence-absence
```{r}
#| label: sim-nocov2
set.seed(3439)    ## Just to make it reproducible
z <- rbinom(nSites, size=1, psi) ## pres/absence
``` 

  * Simulate observations
```{r}
#| label: sim-nocov3
y <- matrix(NA, nrow=nSites, ncol=nVisits)
for(i in 1:nSites) {
    y[] <- rbinom(nVisits, size=1, prob=z[]*p)
}
```   





## Simulated data

::: columns
::: {.column width='50%'}

     Observations
```{r}
#| label: sim-nocov-dat
y
``` 
:::
::: {.column width='50%'}
    
Summary stats

Detections at each site

```{r}
#| label: sim-nocov-ss1
siteDets <- rowSums(y) # Dets at each site
table(siteDets)        # Frequency
```


Proportion of sites known to be occupied

```{r}
#| label: sim-nocov-ss2
naiveOccupancy <- sum(siteDets>0)/nSites
naiveOccupancy 
```
:::
:::




# Model fitting {visibility="hidden"}


## Outline
  

## R package `unmarked'
  
Install and load the package 
  
```{r}
#| label: un-install
#| message: false
## install.packages("unmarked")  
library(unmarked)
```


Documentation (help files) 
  
```{r}
#| label: un-help
#| eval: false
help("unmarked")
```

Documentation (vignettes) 
  
```{r}
#| label: un-vig
#| eval: false
vignette(package="unmarked")
## vignette("spp-dist")
```





## R package `unmarked'

Unlike standard model fitting functions such as `lm` and `glm` 
that require data to be in `data.frames`, data for `unmarked'
has to be formatted in an `unmarkedFrame`.

```{r}
#| label: un-umf
umf <- unmarkedFrameOccu(y=y)
summary(umf)
```


## R package `unmarked'


Fit the single-season occupancy model

```{r}
#| label: un-fit
fm <- occu(~1 ~1, umf)
summary(fm)
```

Estimates obtained using maximum likelihood


## R package `unmarked'

Occupancy estimate ($\hat{\psi}$)

```{r}
#| label: un-back1
backTransform(fm, type="state")
```


Detection probability estimate ($\hat{p}$)

```{r}
#| label: un-back2
backTransform(fm, type="det")
```


## Two types of occupancy estimates

* Unconditional occupancy $\psi=\Pr(z=1)$

* The estimate $\hat{\psi}$ that we computed earlier can be
        thought of as a prediction that we could use for a new
        site or a new time period
    
* Conditional occupancy $\psi^* = \Pr(z_i=1|y_{i})$
    
* Conditional occupancy estimates only apply to sites in the
        sample, at the time of sampling

* In `unmarked', conditional occupancy is computed using
        Empirical Bayes methods, with the `ranef` and `bup`
        functions. In JAGS, full Bayesian inference is used instead. 

* Conditional occupancy estimates can be used to answer
        questions like: 
        
          * Which sites were occupied?
          * How many sites were occupied?

```{r}
#| label: bup
## Posterior probs Pr(z_i=1 | y_i)       
z.post <- ranef(fm)
## Extract posterior means
psi.conditional <-  bup(z.post, stat="mean") 
```
  


## Occupancy estimates

```{r}
#| label: psi-zpm
round(data.frame(y=y, psi.unconditional=predict(fm, type="state")[],
                 psi.conditional), 3)
```




## Occupancy estimates

  How many sites were occupied? 

  We can answer this using posterior prediction.
```{r}
#| label: bup-hist
#| fig.height: 5
#| out.width: "70%"
#| fig.align: 'center'
nsim <- 1000
sites.occupied.post <- predict(z.post, func=sum, nsim=nsim)
par(mai=c(0.9,0.9,0.1,0.1))
plot(table(sites.occupied.post)/nsim, lwd=5, xlab="Sites occupied",
    ylab="Empirical Bayes posterior probability")
abline(v=sum(z), col="red") ## Actual number occupied
```





## Occupancy estimates

  The posterior median (or mean or mode) can be used as a point estimate:

```{r}
#| label: post-mean-eb
psi.cond.estimate <- median(sites.occupied.post)
psi.cond.estimate
```

Quantiles can be used for the 95\% confidence interval:

```{r}
#| label: post-ci-eb
quantile(sites.occupied.post, prob=c(0.025, 0.975))
```


## Bayesian methods

Bayesian methods incorporate prior distributions into inferences.
  
We often have prior information, and it's easy to use.
  
If we don't have prior information, we can use vague priors and obtain
similar inferences as likelihood-based methods.
  
Ultimately, we want the posterior distribution of the model parameters. We'll use MCMC to obtain samples from this distribution. 


## JAGS and the jagsUI package

A reminder...
  
The basic steps of a Bayesian analysis with JAGS are:
  * Write a model in the BUGS language in a text file
  * Put data in a named list
  * Create a function to generate random initial values
  * Specify the parameters you want to monitor
  * Use MCMC to draw posterior samples
  * Use posterior samples for inference and prediction
  

## The BUGS language

The BUGS model file: {\tt occupancy-model.jag}

```{r}
#| label: bugs
#| comment: ''
#| background: 'beige'
writeLines(readLines("occupancy-model.jag"))
``` 




## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data
jags.data <- list(y=y, nSites=nSites, nOccasions=nVisits)
```


Initial values
  
```{r}
#| label: bugs-inits
jags.inits <- function() {
    list(psi=runif(1), p=runif(1), z=rep(1, nSites))
}
``` 


Parameters to monitor
  
```{r}
#| label: bugs-pars
jags.pars <- c("psi", "p", "sitesOccupied")
``` 




## MCMC

This function will compile the model, run adaptive MCMC, and then draw
posterior samples using 3 Markov chains run in parallel.

```{r}
#| label: bugs-mcmc
#| warning: false
#| results: 'hide'
#| cache: false
## install.packages("jagsUI")
library(jagsUI)
jags.post.samples <- jags.basic(data=jags.data, inits=jags.inits,
                                parameters.to.save=jags.pars,
                                model.file="occupancy-model.jag",
                                n.chains=3, n.adapt=100, n.burnin=0,
                                n.iter=2000, parallel=TRUE)
``` 


## Summarize output

```{r}
#| label: bugs-sum
summary(jags.post.samples[,1:4])
``` 



## Traceplots and density plots

```{r}
#| label: bugs-plot
#| out.width: "70%"
#| fig.align: 'center'
plot(jags.post.samples[,1:4])
``` 



# Assignment {visibility="hidden"}


## Assignment

Create a self-contained R script or Rmarkdown file to do the
following: 
  
- Import the Canada Warbler data (`"cawa\_data\_2017\_occu.csv"`) and compute basic summary stats like we did earlier. 
  * Data are ``point count data'' from North Carolina
  * Each 10-min survey was divided into 4, 2.5-min intervals 
    
- Fit the single-season (static) occupancy model using `unmarked' and `JAGS'. 

- Create a `data.frame` to compare the estimates of $\psi$, $p$, and
      ``nSitesOccupied'' from the two analyses. Compare point
      estimates and 95\% CIs. Describe any differences you see in the
      estimates. 
  
<center>
  Upload your {\tt .R} or {\tt .Rmd} file to ELC by 5:00pm on Monday. 
</center>








