---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Occupancy Models Part I'
    slide-number: true
#    smaller: true
    css: ../../custom.css
lang: en
---



## {content-align="center" style="font-size: 160%;"}

<br>

<center>

Occupancy models:<br>simulation and fitting

<br>
  
::: {style="font-size: 70%;"}
WILD(FISH) 8390<br>
Inference for Models of Fish and Wildlife Population Dynamics
:::

</center>


# Motivation {visibility="hidden"}




## Outline

- Motivation

- Simulation

- Classical analysis with 'unmarked'

- Bayesian analysis with 'JAGS'



## Motivation


::: {style="font-size: 90%;"}

We often want to do logistic regression, but the data are subject to measurement error.

<br>
  
::: {.fragment}
Specifically, wildlife data are often fraught with false negatives (ie, the species was there but we didn't detect it).
:::

<br>
  
::: {.fragment}

Ignoring this source of measurement error can result in misleading inferences. 

- Occupancy will be under-estimated

- Habitat associations may be incorrect

:::

:::



## Accounting for detection probability

::: {style="font-size: 75%;"}

The model for the state process is the same as logistic regression:

$$
  \begin{gather*}
    \mathrm{logit}(\psi_i) = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \cdots \\
    z_i \sim \mathrm{Bern}(\psi_i)
  \end{gather*}
$$

::: {.fragment}

Model for the observation process conditional on the state process:

$$
\begin{gather*}
    \mathrm{logit}(p_{ij}) = \alpha_0 + \alpha_1 x_{ij1} + \alpha_2 x_{ij2} + \cdots \\
    y_{ij} \sim \mathrm{Bern}(z_i\times p_{ij})
  \end{gather*}
$$

  
- $\psi_i$ -- probability that the species occurs at site $i$
- $z_i$ -- binary presence/absence variable at site $i$
- $p_{ij}$ -- probability of detecting the species at site $i$ on occasion $j$
- $y_{ij}$ -- binary detection/non-detection data

:::


:::



## Occupancy model without covariates

<br>

$$
  \begin{gather*}
    z_i \sim \mathrm{Bern}(\psi) \\
    y_{ij} \sim \mathrm{Bern}(z_i\times p)
  \end{gather*}
$$

::: {style="font-size: 90%;"}
  
- $\psi$ -- probability that the species occurs at site $i$
- $z_i$ -- binary presence/absence variable at site $i$
- $p$ -- probability of detecting the species at site $i$ on occasion $j$
- $y_{ij}$ -- binary detection/non-detection data

:::




## Assumptions

Assumptions of the basic single-season occupancy models:


::: {.incremental}

- Population closure
    
- Occupancy state (presence/absence) constant over time
    
- Statistical independence
    
- Occupancy state at one site has no impact on occupancy of other sites
 
- Detections are independent, conditional on occupancy state 
    
- Variation in $\psi$ and $p$ described by covariates
  
:::




# Simulation {visibility="hidden"}


## Simulation with no covariates

::: {style="font-size: 75%;"}

1. Specify the parameters and dimensions

```{r}
#| label: sim-nocov1
#| echo: true
psi <- 0.5       ## Occurrence probability
p <- 0.3         ## Detection probability
nSites <- 40
nVisits <- 4
```

::: {.fragment}

2. Simulate presence-absence

```{r}
#| label: sim-nocov2
#| echo: true
set.seed(3439)    ## Just to make it reproducible
z <- rbinom(nSites, size=1, psi) ## pres/absence
```

:::

::: {.fragment}

3. Simulate observations

```{r}
#| label: sim-nocov3
#| echo: true
y <- matrix(NA, nrow=nSites, ncol=nVisits)
for(i in 1:nSites) {
    y[i,] <- rbinom(nVisits, size=1, prob=z[i]*p)
}
```

:::

:::


## Simulated data

::: {.columns style="font-size: 80%;"}
::: {.column width='40%'}

Observations

```{r}
#| label: sim-nocov-dat
#| echo: true
y[1:15,]
``` 

:::

::: {.column width='5%'}
:::

::: {.column width='55%'}
    
<center>Summary stats</center>

Detections at each site

```{r}
#| label: sim-nocov-ss1
#| echo: true
siteDets <- rowSums(y) # Dets at each site
table(siteDets)        # Frequency
```

<br>

Proportion of sites known to be occupied

```{r}
#| label: sim-nocov-ss2
#| echo: true
naiveOccupancy <- sum(siteDets>0)/nSites
naiveOccupancy 
```

:::
:::




# Model fitting
 

## R package 'unmarked'
  
Install and load the package 
  
```{r}
#| label: un-install
#| echo: true
#| message: false
## install.packages("unmarked")  
library(unmarked)
```

<br>

Documentation (help files) 
  
```{r}
#| label: un-help
#| echo: true
#| eval: false
help("unmarked")
```
<br>

Documentation (vignettes) 
  
```{r}
#| label: un-vig
#| echo: true
#| eval: false
vignette(package="unmarked")
## vignette("spp-dist")
```





## R package 'unmarked'

::: {style="font-size: 90%;"}
Unlike standard model fitting functions such as `lm` and `glm` 
that require data to be in `data.frames`, data for 'unmarked'
has to be formatted in an `unmarkedFrame`.
:::

<br>

```{r}
#| label: un-umf
#| echo: true
umf <- unmarkedFrameOccu(y=y)
summary(umf)
```


## R package 'unmarked'


Fit the single-season occupancy model

```{r}
#| label: un-fit
#| echo: true
fm <- occu(~1 ~1, umf)
summary(fm)
```

<!-- Estimates obtained using maximum likelihood -->


## R package 'unmarked'

::: {style="font-size: 95%;"}


Occupancy estimate ($\hat{\psi}$)

```{r}
#| label: un-back1
#| echo: true
backTransform(fm, type="state")
```

::: {.fragment}

Detection probability estimate ($\hat{p}$)

```{r}
#| label: un-back2
#| echo: true
backTransform(fm, type="det")
```

:::

:::


## Two types of occupancy estimates

::: {style="font-size: 75%;"}

1. Expected occupancy $\psi=\Pr(z=1)$

- The estimate $\hat{\psi}$ that we computed earlier can be thought of as a prediction that we could use for a new site or a new time period.

<br>

::: {.fragment}

2. Realized occupancy $\psi^* = \Pr(z_i=1|y_{i})$
    
- Realized occupancy estimates only apply to sites in the sample, at the time of sampling.

- Realized occupancy estimates can be used to answer questions like: 
        
  * Which sites were occupied?
  * How many sites were occupied?

:::



:::



## Realized occupancy in 'unmarked'

In 'unmarked', realized occupancy is computed using Empirical Bayes methods, with the `ranef` and `bup` functions. In JAGS, full Bayesian inference is used instead. 

<br>

```{r}
#| label: bup
#| echo: true
## Posterior probs Pr(z_i=1 | y_i)       
z.post <- ranef(fm)
## Extract posterior means
psi_realized <-  bup(z.post, stat="mean") 
```



## Occupancy estimates

```{r}
#| label: psi-zpm
#| echo: true
round(data.frame(y=y, psi_expected=predict(fm, type="state")[,1],
                 psi_realized), 3)
```




## Occupancy estimates

::: {style="font-size: 80%;"}

How many sites were occupied? 

We can answer this using posterior prediction.

:::


```{r}
#| label: bup-hist
#| echo: true
#| fig.height: 5
#| out.width: "70%"
#| fig.align: 'center'
nsim <- 1000
sites.occupied.post <- predict(z.post, func=sum, nsim=nsim)
par(mai=c(0.9,0.9,0.1,0.1))
plot(table(sites.occupied.post)/nsim, lwd=5, xlab="Sites occupied",
    ylab="Empirical Bayes posterior probability")
abline(v=sum(z), col="red") ## Actual number occupied
```





## Occupancy estimates

::: {style="font-size: 80%;"}

The posterior median (or mean or mode) can be used as a point estimate:

```{r}
#| echo: true
#| label: post-mean-eb
psi.realized.estimate <- median(sites.occupied.post)
psi.realized.estimate
```

<br>

::: {.fragment}

Quantiles can be used for the 95\% confidence interval:

```{r}
#| echo: true
#| label: post-ci-eb
quantile(sites.occupied.post, prob=c(0.025, 0.975))
```

:::

:::

## Bayesian methods

::: {style="font-size: 80%;"}

Bayesian methods incorporate prior distributions into inferences.

<br>
  
We often have prior information, and it's easy to use.
  
<br>

If we don't have prior information, we can use vague priors and obtain
similar inferences as likelihood-based methods.
  
<br>

Ultimately, we want the posterior distribution of the model parameters. We'll use MCMC to obtain samples from this distribution. 

:::


## JAGS and the jagsUI package

<!-- A reminder... -->
  
The basic steps of a Bayesian analysis with JAGS are:

* Write a model in the BUGS language in a text file
* Put data in a named list
* Create a function to generate random initial values
* Specify the parameters you want to monitor
* Use MCMC to draw posterior samples
* Use posterior samples for inference and prediction



## The BUGS language <!-- {background="beige"} -->



The BUGS model file: `occupancy-model.jag`.

<br>

::: {style="font-size: 90%;"}

```{r}
#| label: bugs
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("occupancy-model.jag"))
``` 

:::


## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data
#| echo: true
jags.data <- list(y=y, nSites=nSites, nOccasions=nVisits)
```

<br>

Initial values
  
```{r}
#| label: bugs-inits
#| echo: true
jags.inits <- function() {
    list(psi=runif(1), p=runif(1), z=rep(1, nSites))
}
``` 

<br>

Parameters to monitor
  
```{r}
#| label: bugs-pars
#| echo: true
jags.pars <- c("psi", "p", "sitesOccupied")
``` 



## MCMC

::: {style="font-size: 90%;"}

This function will compile the model, run adaptive MCMC, and then draw
posterior samples using 3 Markov chains run in parallel.

:::

<br>

```{r}
#| label: bugs-mcmc
#| echo: true
#| warning: false
#| results: 'hide'
#| cache: true
## install.packages("jagsUI")
library(jagsUI)
library(coda)
jags.post.samples <- jags.basic(data=jags.data, inits=jags.inits,
                                parameters.to.save=jags.pars,
                                model.file="occupancy-model.jag",
                                n.chains=3, n.adapt=100, n.burnin=0,
                                n.iter=2000, parallel=TRUE)
``` 


## Summarize output

```{r}
#| label: bugs-sum
#| echo: true
summary(jags.post.samples)
``` 



## Traceplots and density plots

```{r}
#| label: bugs-plot
#| echo: true
#| out.width: "70%"
#| fig.align: 'center'
plot(jags.post.samples)
``` 



# Assignment {visibility="hidden"}


## Assignment


::: {style="font-size: 75%;"}

Create a self-contained R script or Rmarkdown file to do the
following: 
  
1. Import the Canada Warbler data (`"cawa_data_2017_occu.csv"`) and compute basic summary stats like we did earlier. 
  * Data are "point count data" from North Carolina
  * Each 10-min survey was divided into 4, 2.5-min intervals 
    
2. Fit the single-season (static) occupancy model using 'unmarked' and 'JAGS'. 

3. Create a `data.frame` to compare the estimates of $\psi$, $p$, and "nSitesOccupied" from the two analyses. Compare point estimates and 95\% CIs. Describe any differences you see in the estimates. 

<br>
  
<center>
  Upload your `.R`, `.Rmd`, or `.qmd` file to ELC by 5:00pm on Tuesday. 
</center>

:::







