---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Binomial *N*-mixture models:<br>simulation, fitting, and prediction'
    slide-number: true
    css: ../../custom.css
lang: en
---


## {content-align="center" style="font-size: 160%;"}


Binomial $N$-mixture models: <br> simulation, fitting, and prediction

  
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
  



# Overview






## Overview

We're often more interested in abundance than occupancy. <br>
  
  
  Even if you're interested in occupancy, you get it for free by modeling abundance. <br>
  
  
  Occupancy isn't useful if all sites are occupied. <br>
  
  
  Binomial $N$-mixture models are useful for studying spatial
  variation in abundance if we have repeated count data, instead of
  repeated detection data.






## Binomial $N$-mixture model
  
  State model (with Poisson assumption)
  \begin{gather*}
    \mathrm{log}(\lambda_i) = \beta_0 + \beta_1 {\color{blue} x_{i1}} +
    \beta_2 {\color{blue} x_{i2}} + \cdots <br>
    N_i \sim \mathrm{Poisson}(\lambda_i)
  \end{gather*}
  
  
  Observation model
  \begin{gather*}
    \mathrm{logit}(p_{ij}) = \alpha_0 + \alpha_1 {\color{blue} x_{i1}}
    + \alpha_2 {\color{Purple} w_{ij}} + \cdots <br>
    y_{ij} \sim \mathrm{Binomial}(N_i, p_{ij})
  \end{gather*}
  
  
  
  Definitions <br>
  $\lambda_i$ -- Expected value of abundance at site $i$ <br>
  $N_i$ -- Realized value of abundance at site $i$ <br>
  $p_{ij}$ -- Probability of detecting \alert{an individual} at site $i$ on occasion $j$ <br>
  $y_{ij}$ -- Count data <br>
  $\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates <br>
  $\color{Purple} w$ -- observation covariate



# Simulation




## Simulation -- No covariates
  
```{r}
#| label: sim-nocov1
#| size: 'scriptsize'
nSites <- 100
nVisits <- 4
set.seed(3439)                         # Make it reproducible
lambda1 <- 2.6                         # Expected value of N at each site
N1 <- rpois(n=nSites, lambda=lambda1)  # Realized values
```
  
  Detection probability and data
```{r}
#| label: sim-nocov2
#| size: 'scriptsize'
p1 <- 0.3
y1 <- matrix(NA, nrow=nSites, ncol=nVisits)
for(i in 1:nSites) {
    y1[i,] <- rbinom(nVisits, size=N1[i], prob=p1)
}
```
  
  
Data and latent abundance
```{r}
#| label: N1y1
#| size: 'scriptsize'
cbind(y1, N1)[1:5,]
```




## Simulation -- Covariates
  
```{r}
#| label: sim-cov1
#| size: 'scriptsize'
forest <- factor(sample(c("Hardwood", "Mixed", "Pine"), nSites, replace=TRUE))
forestMixed <- ifelse(forest=="Mixed", 1, 0)        ## Dummy
forestPine <- ifelse(forest=="Pine", 1, 0)          ## Dummy
temp <- matrix(rnorm(nSites*nVisits), nrow=nSites, ncol=nVisits)
```

  Coefficients, $\lambda$, and $p$
  
```{r}
#| label: nsim-cov2
#| size: 'scriptsize'
beta0 <- 0; beta1 <- -1; beta2 <- 1
lambda2 <- exp(beta0 + beta1*forestMixed + beta2*forestPine)
alpha0 <- -2; alpha1 <- 1
p2 <- plogis(alpha0 + alpha1*temp)
```

  Simulate occupancy and detection data
  
```{r}
#| label: sim-cov3
#| size: 'scriptsize'
N2 <- rpois(nSites, lambda=lambda2)         ## local abundance 
y2 <- matrix(NA, nrow=nSites, ncol=nVisits)
for(i in 1:nSites) {
    y2[i,] <- rbinom(nVisits, size=N2[i], prob=p2[i,])
}
```



## Simulated data
  ::: columns
    ::: {.column width='50%'}
      
      Observations

```{r}
#| label: sim-nocov-dat
#| size: 'scriptsize'
y2[1:19,]
```
  :::
  ::: {.column width='50%'}
    

    { Summary stats <br>}
    
  Detections at each site <br>
```{r}
#| label: sim-nocov-ss1
#| size: 'scriptsize'
# Max count at each site
maxCounts <- apply(y2, 1, max) 
table(maxCounts)              
```

```{r}
#| label: sim-nocov-ss2
#| size: 'scriptsize'
naiveOccupancy <- sum(maxCounts>0)/nSites
naiveOccupancy 
```


Observed abundance
```{r}
#| label: sim-nocov-ss3
#| size: 'scriptsize'
naiveAbund <- sum(maxCounts)
naiveAbund
```
```{r}
#| label: un
#| include: false
library(unmarked)
```
:::
:::




# Prediction




## Prepare data in 'unmarked'
  
```{r}
#| label: un-umf
#| size: 'tiny'
umf <- unmarkedFramePCount(y=y2, siteCovs=data.frame(forest), obsCovs=list(temp=temp))
```

```{r}
#| label: wfac
#| size: 'tiny'
summary(umf)
```




## Fit the model -- without covariates

`pcount` is similar to `occu`, but there's a new argument
  (`K`) that should be an integer much 
  greater than the highest possible value of local abundance. 
```{r}
#| label: un-fit0
#| size: 'tiny'
fm0 <- pcount(~1 ~1, umf, K=100)    
fm0
```


We can use `backTransform` to obtain estimate of average abundance
at each site because there aren't covariates:
```{r}
#| label: back
#| size: 'tiny'
backTransform(fm0, type="state")
```






## Fit the model -- with covariates

Detection depends on `temp`. Abundance depends on `forest`.
```{r}
#| label: un-fit
#| size: 'tiny'
fm <- pcount(~temp ~forest, umf, K=100)    
fm
```


Compare to actual parameter values:
```{r}
#| label: un-compare
#| size: 'tiny'
c(beta0=beta0, beta1=beta1, beta2=beta2); c(alpha0=alpha0, alpha1=alpha1)
```






## Make sure $K$ is high enough

```{r}
#| label: width90
#| include: false
oo <- options(width=90)
```

Estimates should not change when you increase $K$.
  
```{r}
#| label: coef
#| size: 'tiny'
round(coef(fm), digits=4)
```


Looks good:
  
```{r}
#| label: upK
#| size: 'tiny'
fm.test <- pcount(~temp ~forest, umf, K=150)    
round(coef(fm.test), digits=4)
```


If the estimates do change, increase $K$ until they stabilize.



##  Empirical Bayes -- Site-level abundance

```{r}
#| label: ranef
#| size: 'scriptsize'
#| out.width: '80%'
#| fig.align: 'center'
#| fig.width: 9
re <- ranef(fm)
plot(re, layout=c(4,3), subset=site%in%1:12, xlim=c(-1, 11), lwd=5)
```





## Total abundance (in surveyed region)

```{r}
#| label: Ntotal
#| size: 'tiny'
#| out.width: '50%'
#| fig.align: 'center'
N.total.post <- predict(re, func=sum, nsim=1000)
hist(N.total.post, freq=FALSE, main="", xlab="N total", ylab="Probability")
```
Estimate of total abundance in sampled area, with 95\% CI:
```{r}
#| label: Ntotal-stats
#| size: 'tiny'
c(Estimate=mean(N.total.post), quantile(N.total.post, prob=c(0.025, 0.975)))
```








## Prediction in `unmarked'
  
Create `data.frame` with prediction covariates. 
  
```{r}
#| label: preddat
#| size: 'footnotesize'
pred.data <- data.frame(forest=c("Hardwood", "Mixed", "Pine"),
                        temp=0) 
```


Get predictions of $\lambda$ for each row of prediction data.
  
```{r}
#| label: predpsi
#| size: 'footnotesize'
lambda.pred <- predict(fm, newdata=pred.data,
                       type='state', append=TRUE)
```


View $\lambda$ predictions
  
```{r}
#| label: psi-head
#| size: 'footnotesize'
print(head(lambda.pred), digits=2)
```






## Prediction in `unmarked'

```{r}
#| label: pred-psi1
#| fig.width: 7
#| fig.height: 5.5
#| size: 'tiny'
#| out.width: '80%'
#| fig.align: 'center'
bpx <- barplot(lambda.pred$Predicted, ylab="Expected value of abundance", #col="blue",
               ylim=c(0,3.5), names=lambda.pred$forest, xlab="Forest type"); box()
arrows(bpx, lambda.pred$Predicted, bpx, lambda.pred$Predicted+lambda.pred$SE,
       angle=90, length=0.1)
```



## The BUGS model -- without covariates

```{r}
#| label: bugs0
#| size: 'scriptsize'
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model.jag"))
```

```{r}
#| label: jagsUI0
#| include: false
library(jagsUI)
```

## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data0
#| size: 'scriptsize'
jags.data0 <- list(y=y2, nSites=nSites, nOccasions=nVisits)
```
  
Initial values
  
```{r}
#| label: bugs-inits0
#| size: 'scriptsize'
jags.inits0 <- function() list(lambda=runif(1), p=runif(1), N=maxCounts)
```
  
  
Run it
  
```{r}
#| label: bugs-pars0
#| results: 'hide'
#| message: false
#| cache: false
jags0.post.samples <- jags.basic(data=jags.data0, inits=jags.inits0,
                                 parameters.to.save=c("lambda", "p"), 
                                 model.file="Nmix-model.jag",
                                 n.chains=3, n.adapt=100, n.burnin=0,
                                 n.iter=2000, parallel=TRUE)
```


## The BUGS model -- with covariates
```{r}
#| label: bugs
#| size: 'scriptsize'
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("Nmix-model-covs.jag"))
```
```{r}
#| label: jagsUI
#| include: false
library(jagsUI)
```


## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data
#| size: 'small'
jags.data <- list(y=y2, temp=temp,
                  forestMixed=forestMixed,
                  forestPine=forestPine,
                  nSites=nSites, nOccasions=nVisits)
```



Initial values
  
```{r}
#| label: bugs-inits
#| size: 'small'
jags.inits <- function() {
    list(beta0=rnorm(1), alpha0=rnorm(1), N=maxCounts)
}
```

Parameters to monitor
  
```{r}
#| label: bugs-pars
#| size: 'small'
jags.pars <- c("beta0", "beta1", "beta2",
               "alpha0", "alpha1", "totalAbundance")
```





## MCMC
  
```{r}
#| label: bugs-mcmc
#| size: 'scriptsize'
#| message: false
#| results: 'hide'
#| cache: true
library(jagsUI)
jags.post.samples <- jags.basic(data=jags.data, inits=jags.inits,
                                ## Monitor each "N[i]" 
                                parameters.to.save=c(jags.pars, "N"), 
                                model.file="Nmix-model-covs.jag",
                                n.chains=3, n.adapt=100, n.burnin=0,
                                n.iter=2000, parallel=TRUE)
```



## Summarize output
```{r}
#| label: bugs-sum
#| size: 'tiny'
summary(jags.post.samples[,jags.pars])
```



## Local abundance
```{r}
#| label: localN
#| out.width: '70%'
#| fig.align: 'center'
#| size: 'scriptsize'
plot(jags.post.samples[,paste0("N[", 1:4, "]")])
```


## Traceplots and density plots

```{r}
#| label: bugs-plot1
#| size: 'footnotesize'
#| out.width: "0.7<br>textwidth"
#| fig.align: 'center'
plot(jags.post.samples[,jags.pars[1:3]])
```


## Traceplots and density plots

```{r}
#| label: bugs-plot2
#| size: 'footnotesize'
#| out.width: "0.7<br>textwidth"
#| fig.align: 'center'
plot(jags.post.samples[,jags.pars[4:5]])
```

## Bayesian prediction
  
First, extract the $p$ coefficients
  
```{r}
#| label: psi-coefs
#| size: 'scriptsize'
p.coef.post <- as.matrix(jags.post.samples[,c("alpha0","alpha1")])
head(p.coef.post, n=4)
```
  
Create prediction matrix, one row for each MCMC iteration.
  
```{r}
#| label: p-predmat
#| size: 'scriptsize'
n.iter <- nrow(p.coef.post)
temp.pred <- seq(-3, 3, length=50)
p.post.pred <- matrix(NA, nrow=n.iter, ncol=length(temp.pred))
```
  
Predict $p$ for each MCMC iteration.
  

```{r}
#| label: psi-pred-bayes
#| size: 'scriptsize'
for(i in 1:n.iter) {
    p.post.pred[i,] <- plogis(p.coef.post[i,"alpha0"] +
                              p.coef.post[i,"alpha1"]*temp.pred)
}
```



## Bayesian prediction

```{r, echo=-1}
#| label: psi-pred-post-meanCI
#| fig.align: 'center'
#| out.width: '70%'
#| fig.height: 5
#| cache: false
#| dpi: 200
par(mai=c(0.9,0.9,0.1,0.1))  
plot(temp.pred, p.post.pred[1,], type="l", xlab="Temperature (standardized)",
     ylab="Detection probability", ylim=c(0, 1), col=gray(0.8))
for(i in seq(1, n.iter, by=10)) {  ## Thin by 10
    lines(temp.pred, p.post.pred[i,], col=gray(0.8))  }
pred.post.mean <- colMeans(p.post.pred)
pred.post.lower <- apply(p.post.pred, 2, quantile, prob=0.025)
pred.post.upper <- apply(p.post.pred, 2, quantile, prob=0.975)
lines(temp.pred, pred.post.mean, col="blue")
lines(temp.pred, pred.post.lower, col="blue", lty=2)
lines(temp.pred, pred.post.upper, col="blue", lty=2)
```


## Prior sensitivity

Questions
  * What are the implied priors on $\lambda$ and $p$?
  * What would be the consequence of changing the priors?
  
  [Prior predictive checks] can help us answer these questions. <br>

  Prior prediction can be acheived by replacing the data with missing values <br>


## Prior sampling

Replace data with missing values
  
```{r}
#| label: bugs-data2
#| size: 'small'
jags.data2 <- jags.data
jags.data2$y[] <- NA     # Replace data with missing values
```
  
Fit model (Really, we're just sampling from the prior)

```{r}
#| label: bugs-mcmc-prior
#| size: 'scriptsize'
#| message: false
#| results: 'hide'
#| cache: true
jags.prior.samples <- jags.basic(data=jags.data2, inits=jags.inits,
                                 parameters.to.save=jags.pars,
                                 model.file="Nmix-model-covs.jag",
                                 n.chains=3, n.adapt=100, n.burnin=0,
                                 n.iter=2000, parallel=TRUE)
```


## Prior summary

```{r}
#| label: summary-prior
#| size: 'tiny'
summary(jags.prior.samples)
```



## Prior prediction

Push the prior samples through the model to predict $p$

```{r}
#| label: psi-coefs-prior
#| size: 'scriptsize'
p.coef.prior <- as.matrix(jags.prior.samples[,c("alpha0","alpha1")])
p.prior.pred <- matrix(NA, nrow=n.iter, ncol=length(temp.pred))
for(i in 1:n.iter) {
    p.prior.pred[i,] <- plogis(p.coef.prior[i,"alpha0"] +
                               p.coef.prior[i,"alpha1"]*temp.pred)
}
```

## Bayesian prediction

Compute prior mean and 95\% CI for $p$ predictions
```{r}
#| label: prior-post-pred-1
#| size: 'scriptsize'
pred.prior.mean <- colMeans(p.prior.pred)
pred.prior.lower <- apply(p.prior.pred, 2, quantile, prob=0.025)
pred.prior.upper <- apply(p.prior.pred, 2, quantile, prob=0.975)
```
  
Show credible regions using shaded polygons. Include a few prior prediction lines.

```{r}
#| label: prior-post-pred-2
#| size: 'scriptsize'
#| fig.height: 6
#| dev: 'png'
#| dpi: 200
#| fig.show: 'hide'
plot(temp.pred, p.post.pred[1,], type="n", xlab="Temperature (standardized)",
     ylab="Detection probability", ylim=c(0, 1.3), col=gray(0.8))
polygon(x=c(temp.pred, rev(temp.pred)),
        y=c(pred.post.lower, rev(pred.post.upper)),
        col=rgb(0,0,1,0.5), border=NA)                   # Post CI
polygon(x=c(temp.pred, rev(temp.pred)),
        y=c(pred.prior.lower, rev(pred.prior.upper)),
        col=rgb(0,1,0,0.5), border=NA)                   # Prior CI
for(i in seq(1, n.iter, by=100)) {  ## Thin by 100
    lines(temp.pred, p.prior.pred[i,], col=gray(0.8))  } # Prior preds
lines(temp.pred, pred.post.mean, col="blue", lwd=2)
lines(temp.pred, pred.prior.mean, col="darkgreen", lwd=2)
legend(-3, 1.3, c("Prior mean", "Prior samples", "Posterior mean"),
       col=c("darkgreen", "grey", "blue"), lwd=2)
```


## Prior and posterior prediction
  
![]{prior-post-pred-2-1}


## Prior predictive recap

There are two common strategies for specifying priors:
  * Use uniform/flat priors to obtain results that will often be similar to likelihood-based methods.
  * Use priors that reflect available information.
  
  
The second approach can be used even if there isn't much prior
information available. Priors can allow for the
possibility of being ``surprised'' by the data, but without
putting most of the prior weight on extreme values. <br>
  
  
Often, highly diffuse priors on the link scale result in most of the
prior weight being on extreme values on the natural scale. <br>
  
  
Prior predictive checks can be used to assess this possibility, and
should be a standard component of Bayesian analysis. <br>




# Assignment


## Assignment

  Create a self-contained R script or Rmarkdown/Quarto file
  to do the following:
  
Using the simulated data, compare the prior and posterior
      predictive distributions of $\lambda$ for each of the 3 forest
      types. Hint: because forest type is categorical (not
      continuous), you can visualize the prior and posterior
      predictive distributions as histograms for each forest type.  <br>
  
  Upload your `.R`, `.Rmd`, or `.qmd` file to ELC by 5:00 PM on Tuesday.







