---
title: "Lecture  13 - Dynamic N-mixture models"
author: "Heather Gaya"
date: "November 2021"
output:
  beamer_presentation: 
    fig_width: 5
    fig_height: 4
    highlight: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(tidy = 'styler')
knitr::opts_chunk$set(dev = 'pdf')
library(unmarked)
library(coda)
library(jagsUI)
library(knitr)
library(raster)
library(ggplot2)
```
# Overview 

  We've previously seen N-mixture models across only one year or season 
  
  \pause
  \vfill
  But often we want to know how populations change over time. 
  
  \pause
  \vfill
  Dynamic $N$-mixture models are useful if we have repeated count data without uniquely marked individuals. 
  
  \pause
  \vfill
 Dynamic models are also very flexible. We can consider how detection parameters change between years, add temporal autocorrelation, account for spatial variation in density, and/or account for treatment effects.  
 

# Dynamic N-Mixtures 
State model (with Poisson assumption)
  \begin{gather*}
    \mathrm{log}(\lambda_{it}) = \beta_0 + \beta_1 {\color{blue} x_{it1}} +
    \beta_2 {\color{blue} x_{it2}} + \cdots \\
    N_{it} \sim \mathrm{Poisson}(\lambda_{it})
  \end{gather*}
  \pause
  Observation model
  \begin{gather*}
    \mathrm{logit}(p_{ijt}) = \alpha_0 + \alpha_1 {\color{blue} w_{it1}}
    + \alpha_2 {\color{purple} w_{ijt2}} + \cdots \\
    y_{ijt} \sim \mathrm{Binomial}(N_{it}, p_{ijt})
  \end{gather*}
  \pause
  \small
  $\lambda_{it}$ -- Expected value of abundance at site $i$ in year $t$
  
  $N_{it}$ -- Realized value of abundance at site $i$ in year $t$
  
  $p_{ijt}$ -- Probability of detecting \alert{an individual} at site $i$ on occasion $j$ in year $t$
  
  $y_{ijt}$ -- Count data
  
  $\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates 
  
  $\color{purple} w_1$ and $\color{purple} w_2$ -- observation covariates


# Temporal Autocorrelation
The previous model allows for independence between years.

\pause
\vfill

But we might expect that abundance at year $t$ is related to abundance at year $t-1$.

\pause
\vfill
Dail and Madsen (2011) suggested an expansion to the $N$-mixture framework that allows us to consider populations from this B.I.D.E model perspective.

# Temporal Autocorrelation (Cont')
\small
Year $k = 1$:
\begin{gather*}
    \mathrm{log}(\lambda_{i1}) = \beta_0 + \beta_1 {\color{blue} x_{i1}} \cdots \\
    N_{i1} \sim \mathrm{Poisson}(\lambda_{i1})
  \end{gather*}
  \pause
  \small
Years $k > 1$:
  \begin{gather*}
    S_{it} \sim \mathrm{Binomial}(N_{i(t-1)}, \phi_{it}) \\
    G_{it} \sim \mathrm{Poisson}(N_{i(t-1)}*\gamma_{it}) \\
    N_{it} = S_{it} + G_{it}
  \end{gather*}
  \small
  \pause
$\lambda_{it}$ -- Expected value of abundance at site $i$ in year 1

$N_{it}$ -- Realized value of abundance at site $i$ in year $t$

$\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates

$\phi_{it}$ -- Apparent survival at site $i$ from year $t-1$ to $t$

$\gamma_{it}$ -- Apparent recuritment at site $i$ from year $t-1$ to $t$

$S_{it}$ -- Realized number of individuals that survived/didn't emmigrate

$G_{it}$ -- Realized number of new individuals

# Simulation

First lets simulate some abundance data from a landscape.

```{r, unnecessarily fun rasters, eval = TRUE, echo = F}
set.seed(100)
r <- raster(xmn=0, ymn=0, xmx=1, ymx=1, ncols = 16, nrows = 8)
r[] <- 0
r[c(1:3, 17,19,5:7, 9:12,14:15, 
    22,25,28,30,32)] <- 1
r[c(33,35,38,41,44,46,48, 49:51,54,57,59:60, 62,64)] <- 1.25
r[c(65:66,70,73:74,78,80, 81,83, 86,89,91,94,96)] <- 2
r[c(97,99,102,105,108,110,112)] <- 2
r[c(113:115,117:119,121,124,126:128)] <-1.2
#writeRaster(r, "Elevation.tif", overwrite = T)

p <- raster(xmn=0, ymn=0, xmx=1, ymx=1, ncols = 16, nrows = 8)
p[] <- runif(128)
p[c(4,8,13,16,18,20,21,23,24,26,27,29,31,34,36,37,39,40,42,
  43,45,47,52,53,55,56,58,61,63,67,68,69,71,72,75,76,
  77,79,82,84,85,87,88,90,92,93,95,98,100, 101, 103, 104,106,107, 109,
  111,116,120,122,123,125)] <- 0
#writeRaster(p, "Precipitation1.tif")
p[] <- runif(128)
p[c(4,8,13,16,18,20,21,23,24,26,27,29,31,34,36,37,39,40,42,
  43,45,47,52,53,55,56,58,61,63,67,68,69,71,72,75,76,
  77,79,82,84,85,87,88,90,92,93,95,98,100, 101, 103, 104,106,107, 109,
  111,116,120,122,123,125)] <- 0
#writeRaster(p, "Precipitation2.tif")
p[] <- runif(128)
p[c(4,8,13,16,18,20,21,23,24,26,27,29,31,34,36,37,39,40,42,
  43,45,47,52,53,55,56,58,61,63,67,68,69,71,72,75,76,
  77,79,82,84,85,87,88,90,92,93,95,98,100, 101, 103, 104,106,107, 109,
  111,116,120,122,123,125)] <- 0
#writeRaster(p, "Precipitation3.tif")
p[] <- runif(128)
p[c(4,8,13,16,18,20,21,23,24,26,27,29,31,34,36,37,39,40,42,
  43,45,47,52,53,55,56,58,61,63,67,68,69,71,72,75,76,
  77,79,82,84,85,87,88,90,92,93,95,98,100, 101, 103, 104,106,107, 109,
  111,116,120,122,123,125)] <- 0
#writeRaster(p, "Precipitation4.tif")
p[] <- runif(128)
p[c(4,8,13,16,18,20,21,23,24,26,27,29,31,34,36,37,39,40,42,
  43,45,47,52,53,55,56,58,61,63,67,68,69,71,72,75,76,
  77,79,82,84,85,87,88,90,92,93,95,98,100, 101, 103, 104,106,107, 109,
  111,116,120,122,123,125)] <- 0
#writeRaster(p, "Precipitation5.tif")

```
Elevation stays the same but precipitation will change each year

```{r Simulate Dynamic,eval= TRUE, echo = TRUE}
set.seed(100)
elev <- raster("Elevation.tif")
precip1 <- raster("Precipitation1.tif")
precip2 <- raster("Precipitation2.tif")
precip3 <- raster("Precipitation3.tif")
precip4 <- raster("Precipitation4.tif")
precip5 <- raster("Precipitation5.tif")
```

# Simulation (cont.)
\small
```{r, eval = T, echo = TRUE,tidy.opts=list(width.cutoff=60),tidy=TRUE}
n.years <- 5
sites <- data.frame(x = rep(seq(.025, .95, length.out = 5),5), 
                    y = rep(seq(.025,.95, length.out = 5), each = 5))
n.sites <- nrow(sites)
head(sites, n = 5)
```

# Simulation (cont.)
\small
```{r, eval = T, echo = T}
plot(elev)
points(sites, pch = 3)
```

# Get the Elevation and Precipitation Covariates For Each Site
\small
```{r, covariates, eval = T, echo = TRUE}
sites$elev <- extract(elev, sites[,c("x", "y")])
sites$precip1 <- extract(precip1, sites[,c("x", "y")])
sites$precip2 <- extract(precip2, sites[,c("x", "y")])
sites$precip3 <- extract(precip3, sites[,c("x", "y")])
sites$precip4 <- extract(precip4, sites[,c("x", "y")])
sites$precip5 <- extract(precip5, sites[,c("x", "y")])
precip <- data.frame(p1 = sites$precip1, 
                     p2 = sites$precip2, p3 = sites$precip3, 
                     p4 = sites$precip4, p5 = sites$precip5)
precip <- as.matrix(precip)
elevation <- matrix(sites$elev)
```



# Simulate Site-Specific Parameters
Simulate initial N, $\phi$ and $\gamma$
\small
```{r, eval  = TRUE, echo = TRUE}
psi0 <- 3
psi1 <- -.25
phi0 <-.8
phi1 <- -3
gamma0 <- -3
gamma1 <- .2
gamma2 <- .9
phi <- gamma <- matrix(NA, nrow = n.sites, ncol = n.years)
for(i in 1:n.years){
phi[,i] <- plogis(phi0 + 
          phi1*(precip[,i]-mean(precip))/sd(precip))
gamma[,i] <- exp(gamma0 + 
          gamma1*(precip[,i]-mean(precip))/sd(precip) +
          gamma2*(elevation-mean(elevation))/sd(elevation))
}
```

# Simulate Abundance
\small
```{r, echo = TRUE}
n.sites <- nrow(sites)
N <- S <- G <- array(NA, dim = c(n.sites, n.years))
lambda <- array(NA, dim = n.sites)
for (i in 1:n.sites){
  lambda[i] <- exp(psi0 + 
               psi1*(elevation[i]-mean(elevation))/sd(elevation))
  N[i,1] <- rpois(1, lambda[i])
 for (t in 2:n.years){
   S[i,t] <- rbinom(1, N[i,t-1], phi[i,t])
   G[i,t] <- rpois(1, N[i,t-1]*gamma[i])
   N[i,t] <- S[i,t] + G[i,t]
 }
}
```
# Simulate
Let's see what N looks like for the first 3 sites
```{r, echo = T, include=T}
N[1:3,]
```

# Simulate Detection
We'll use a time-varying detection for simulation but it could be anything. 


```{r, echo = T}
n.visit <- 4
p <- rnorm(n.years, mean = .3, .1)
y1 <- array(NA, dim = c(n.sites, n.visit, n.years))
for (i in 1:n.sites){
  for (t in 1:n.years){
    y1[i,,t] <- rbinom(n.visit, size = N[i,t], 
                       prob = p[t])
  }
}
```

# Inspect Our Simulation
Grey = observed data, black = simulated abundance
```{r, echo = F, include = T}
max.seen <- array(NA, dim = c(n.sites, n.years))
for(i in 1:n.sites){ 
  for(t in 1:n.years){
    max.seen[i,t] <- max(y1[i,,t])
  }}
tot <- data.frame(obs = c(max.seen[,1], max.seen[,2], max.seen[,3], max.seen[,4], max.seen[,5]), abundance = c(N[,1], N[,2], N[,3], N[,4], N[,5]), year = rep(1:5, each = n.sites), site = rep(1:n.sites, n.years))
ggplot(tot, aes(x =year, group = site))+
  geom_line(aes(y = obs), col = "grey")+
    geom_line(aes(y = abundance))+
  facet_wrap(~site)+
  theme_classic()
```
# Unmarked 

Remember in real life we'd only have precipitation, elevation and the detection matrix (y1). We wouldn't know anything else. 

\pause

Unmarked wants the observation data in a M x (J*T) matrix. M = sites, J = Surveys within a year (or study period), T = number of years.
\small
```{r unmarked, echo = T}
y_matrix <- cbind(y1[,,1], y1[,,2], y1[,,3], 
                  y1[,,4], y1[,,5])
```
\tiny
```{r, echo = T}
head(y_matrix, n = 2)
```

# Setup in Unmarked
\small
```{r, echo = T}
umf <- unmarkedFramePCO(y = y_matrix, 
        siteCovs = data.frame(elevation),
        yearlySiteCovs = list(precip=precip),
        numPrimary = 5)
```

# Run in Unmarked
This may take awhile. 
\small
```{r, echo = T, eval = T}
m1 <- pcountOpen(~scale(elevation), 
      ~scale(elevation)+scale(precip), 
      ~scale(precip), 
      ~1, umf, K=40, mixture = "P",
      dynamics = "autoreg")
```
Order of equations is lambda, apparent recruitment (gamma), apparent survival, detection

# Unmarked Output
\tiny
```{r, echo = T}
m1
```

# Extract real estimates for sites
```{r, echo = T}
re <- ranef(m1)
```
Let's check site 1 each year and compare with our simulation
\small

```{r, echo = T,}
round(bup(re, stat="mean")[1,], digits = 2)
confint(re, level=0.95)[1,,] # 95% CI
N[1,]
```

# JAGS
\tiny
```{r, eval = F, include = T, echo = T,}
model {
for (i in 1:n.sites) {
  log(lambda[i]) <- psi0 + psi1*elevation[i]
  
  N[i,1] ~dpois(lambda[i]) 
  
  for (t in 2:n.years){
    logit(phi[i,t]) <- phi0 + phi1*precip[i,t]
    log(gamma[i,t]) <- gamma0 + gamma1*elevation[i]+gamma2*precip[i,t]
    
    S[i,t] ~ dbin(phi[i,t], N[i,t-1])
    G[i,t] ~ dpois(N[i,t-1]*gamma[i,t])
             
    N[i,t] <- S[i,t] + G[i,t]
  } #end t
  for (t in 1:n.years){
  for (j in 1:n.visit){
    y1[i,j,t] ~ dbin(p[t], N[i,t])
    } #end j
  }#end t again
  } #end i
  
  for(t in 1:n.years){
     p[t] ~ dunif(0,1)
  }
  gamma0 ~ dunif(-5,5)
  gamma1 ~ dunif(-5,5)
  gamma2 ~ dunif(-5,5)
  psi0 ~ dunif(-5,5)
  psi1 ~ dunif(-5,5)
  phi0 ~ dnorm(0,.3)
  phi1 ~ dnorm(0,.3)
}
```

# Send to JAGS
\small
```{r, eval = T, echo = T, include = T}
jd <- list(n.sites =n.sites, n.visit = n.visit,
           n.years = n.years, y1 = y1,
           elevation = as.vector(scale(elevation)),
           precip = (precip - mean(precip))/sd(precip))
params = c("p", "psi0", "psi1", 
           "phi0", "phi1", 
           "gamma0", "gamma1", "gamma2")
```
# What do we do about initial values? 
Initial values are tricky because the model is recursive. Luckily, we have a fun loop we can use to help. 

\small
```{r, echo = T}
getInits <- function(counts, sites, years) {
  nSites <- sites
  nYears <- years
  N <- array(NA_integer_, c(nSites, nYears))
  G <- S <- array(NA_integer_, c(nSites,nYears))
  for(i in 1:nSites){
    N[i,1:nYears] <- max(counts[i,,], na.rm=TRUE)+2
  }
    S[,1] <- G[,1] <- NA
    for(t in 2:nYears) {
      S[,t] <- rbinom(nSites, size=N[,t-1], 0.6)
      G[,t] <- N[,t]-S[,t]
    }
  N.r <- N
  N.r[,2:nYears] <- NA
  return(list(S=S, G=G, N = N.r))
}
```
# Create Initial Values
```{r, eval = T, echo = T}
inits <- getInits(y1, n.sites, n.years)
ji <- function() {
  list(psi0 = runif(1),psi1 = runif(1), 
    phi0 = runif(1), phi1 = runif(1),
    gamma0 = runif(1), gamma1 = runif(1), 
    gamma2 = runif(1), S = inits$S, 
    G= inits$G, N = inits$N
    )
}

```

# Run JAGS
```{r, eval = T, echo = T,results = 'hide'}
jags.post <- jags.basic(data=jd, inits = ji,
             parameters.to.save=params,
             model.file="dynamic_n.txt",
             n.chains=3, n.adapt=100, n.burnin=2000,
             n.iter=3000, parallel=TRUE)
```

# JAGS Output 
\small
```{r, echo = T, eval = F}
summary(jags.post)$quantiles[-1,]
```
```{r, echo = F}
kable(summary(jags.post)$quantiles[-1,], 
      digits = 2,align = "c")
```

# Trace Plot
\small
```{r, echo = T}
plot(jags.post[,c("psi0","gamma0"),])
```
Need to run for longer to get more reliable values. 

# Compare with Truth
Pretty close to the simulated values
```{r, echo =F}
true <- c(gamma0, gamma1, gamma2, p, phi0, phi1, psi0, psi1)
est <- summary(jags.post)$quantiles[-1,3]
up <- summary(jags.post)$quantiles[-1,5]
low <- summary(jags.post)$quantiles[-1,1]
outs <- data.frame(Simulated = true, JAGS = est, lower = low, upper = up)
kable(outs, digits = 2, align = "c")
```

# Assignment 


1. Fit the unmarked model but allow p to vary by year. Hint: Look at the unmarked help page. Compare the AIC score for the model with time varying detection and the one we ran in class. Did unmarked correctly identify the true model? 

\vfill
2. Using the JAGS model, compare the true abundance and estimated abundance of sites 1, 2, and 3. Did the JAGS model capture the true value? Be sure to plot means and CIs for your JAGS model output. 


