---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Non-spatial mark-recapture for <br> estimating population size'
    slide-number: true
	echo: true
lang: en
---





## {content-align="center" style="font-size: 140%;"}

<br>

<center>
  
Non-spatial mark-recapture for <br> estimating
population size

<br> 
  
::: {style='font-size: 80%;'}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
:::
  
</center>





# Overview {visibility="hidden"}

## Outline
  

## Mark-recapture overview

Until now, we've focused on models for data on aggregated quantities
like abundance and occupancy. 

<br> 
  
Now we will talk about data on uniquely identifiable individuals. 

<br>
  
This is nice because all individuals are different, and we're often interested in these differences...
  
...even when our primary goal is inference on population-level parameters like abundance and density.




## Closed populations

For this lecture, we will assume population closure:
  
- No births or deaths during the time of sampling (demographic closure)
- No permanent immigration or emigration (geographic closure)
  
::: {.fragment}
There are several other assumptions that depend on the specific model being employed, including:
  
- No misidentification
- No band/tag loss
- etc...
:::
  
::: {.fragment}
Later, we will relax the closure assumption so that we can study survival, recruitment, and movement.
:::



## Mark-recapture overview

The simplest estimator of abundance is 
$$
    \hat{N} = \frac{n}{\hat{p}}
$$
where $n$ is the number of individuals detected, $p$ is detection probability, and $E(n)=Np$.
  
In distance sampling, we modeled detection probability as a function of distance, and we replaced $p$ with average detection probability. 
  
For the simplest mark-recapture problems, $p$ is a constant (scalar) and we estimate it by trapping on multiple occasions and marking and recapturing as we go.



## Classical approaches

Lincoln-Peterson

- $J=2$ occasions
- Data are:
      
  * $n_1$ -- number of individuals captured on first occasion
  * $n_2$ -- number of individuals captured on second occasion
  * $m_2$ -- number of recaptures

Estimator is $\hat{N} = n_1n_2/m_2$
  
  
::: {.fragment}
Closed population models
  
- Usually better to have $J>2$ occasions
- Allows for assessment of behavioral effects, individual heterogeneity, etc...
:::
  



## Mark-recapture data
  
Mark-recapture data are often called "capture histories." Here's an example with $n=6$ animals captured on $J=3$ occasions. 
  
------  | -- | -- | --
Individual & 1 & 2 & 3
------  | -- | -- | --
1 | 0 | 0 | 1
2 | 1 | 1 | 1
3 | 0 | 1 | 0
4 | 0 | 1 | 1
5 | 1 | 0 | 0
6 | 0 | 0 | 1
  
  
How do we estimate $p$ (and $N$) from these data?

You can get a sense for $p$ (and $N$) by looking at the data.
  
- If $p\approx 1$, most individuals would be captured on all 3 occasions, and few would go undetected
- If $p\approx 0$, few individuals would be captured  more than once, and many would go undetected


##  Closed population model (if we knew *N*) 
  
State model
  
There isn't one! $N$ is often treated as a constant, although we sometimes put a Poisson or binomial prior on it.
  
  
Observation model (supposing $N$ was known)

$$
    y_{ij} \sim \mathrm{Bernoulli}(p) \;\; \mathrm{for}\, i=1,\dots,N 
$$
  
  
Definitions
- $N$ -- Population size
- y_{ij}$ -- Capture histories, indicating if individual $i$ was captured on occasion $j$
- $p$ -- Capture probability
  
The problem with this formulation is that we don't observe the "all zero" encounter histories (and thus we don't know $N$).  As a result, the Bernoulli assumption isn't valid, and we need to develop a model that describes how many individuals we failed to capture.



## Closed population estimation options

Conditional likelihood
  
- Estimate $p$, and then estimate $N$ as a derived quantity. 
- Most software is based on the work by Huggins, which allows for continuous covariates. 
  
  
Joint likelihood
  
- Estimate $N$ and $p$ jointly
   
Data augmentation
  
- Tack on many "all zero" encounter histories and estimate how many of them actually occurred
- Usually, but not necessarily, used in Bayesian inference
  


## Joint likelihood

  
The joint likelihood is based on the multinomial distribution, with
the "all-zero" encounter history being unobserved.
$$
    \{f_1, \dots, f_{K+1}\} \sim \mathrm{Multinomial}(N, \pi_1, \dots, \pi_{K+1}) \;\;
$$
where $\{f_1, \dots, f_K\}$ are the frequencies of the observable capture
histories.
  
Here's an example with $J=3$ capture occasions.  <br>
  
----------------- | -----------------------------
Capture history   |     Multinomial cell probability
----------------- | -----------------------------
      100                   | $\pi_1=p_1(1-p_2)(1-p_3)$
      010                   | $\pi_2=(1-p_1)p_2(1-p_3)$
      001                   | $\pi_3=(1-p_1)(1-p_2)p_3$
      110                   | $\pi_4=p_1p_2(1-p_3)$
      101                   | $\pi_5=p_1(1-p_2)p_3$
      011                   | $\pi_6=(1-p_1)p_2p_3$
      111                   | $\pi_7 = p_1p_2p_3$
      000                   | $\pi_0 = \prod_j 1-p_j$
----------------- | -----------------------------


## Model variations
  
More important than the estimation approach, is identification of the sources of variation in capture probability ($p$).
  
Otis et al. (1978) identified several model variations: 
  
- $M_0$ -- $p$ is constant
- $M_t$ -- unique $p$ for each capture occasion
- $M_b$ -- behavioral response with $p$ different than recapture probability $c$
- $M_h$ -- individual heterogeneity in $p$
- Finite mixture models
- Continuous mixture models (not available in RMark)
   
These can be combined, but beware of identifiability issues. See Otis et al. (1978) for details.
   
Later we'll talk about "individual covariate" models.


## Software options
  
Program MARK

- Has the most model types
- Written for Windows, but can be executed on other platforms
  
R package 'RMark'
  
- Function `mark` runs MARK from R.

R package 'marked'
  
- Includes CJS and JS models
  
R package 'mra'
  
- Includes closed-population, CJS, and JS models
  
R package 'Rcapture'
  
- Includes closed-population
  
R package 'unmarked'
  
- Likelihood-based data augmentation with \inr{occu}
  
R package 'secr'
  
- Designed for spatially explicit models, but see \inr{closedN}
  


# Simulation


```{r}
#| include: false
#| echo: false
set.seed(34889243)
```


## Model *M~0~*
  
Parameters
  
```{r}
#| label: sim-M0-pars
N <- 100
p <- 0.2
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-M0-ch
J <- 4  ## Occasions
y.all <- matrix(NA, N, J)
for(i in 1:N) {
    y.all[i,] <- rbinom(J, 1, p)
}
```
  
Observed capture histories (data)
  
```{r}
#| label: sim-M0-y1
captured <- rowSums(y.all)>0
(n <- sum(captured))
y <- y.all[captured,]
y[1:3,]
```




## Summary stats

Capture history frequencies
```{r}
#| label: M0-hist
histories <- apply(y, 1, paste, collapse="")
sort(table(histories))
```


Detection frequencies
```{r}
#| label: M0-freq
y.tilde <- rowSums(y)
sort(table(y.tilde))
```


## Model *M~t~* -- Temporal variation

Capture probability for each occasion
```{r}
#| label: sim-Mt-pars
p.t <- c(0.3, 0.5, 0.2, 0.4)
```
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mt-ch
y.all.Mt <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mt[i,] <- rbinom(J, 1, p.t) }
```
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mt-y1
captured.Mt <- rowSums(y.all.Mt)>0
n.Mt <- sum(captured.Mt)
y.Mt <- y.all.Mt[captured.Mt,]
y.Mt[1:3,]
colSums(y.Mt)
```

## Model *M~b~* -- Behavioral response

Capture probability for each occasion
```{r}
#| label: sim-Mb-pars
p.b <- 0.3
c <- 0.5  ## Trap happy
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mb-ch
y.all.Mb <- matrix(NA, N, J)
prevcap <- matrix(FALSE, N, J)
for(i in 1:N) {
    y.all.Mb[i,1] <- rbinom(1, 1, p.b)
    for(j in 2:J) {
        prevcap[i,j] <- any(y.all.Mb[i,1:(j-1)]>0)
        prob <- ifelse(prevcap[i,j], c, p.b)
        y.all.Mb[i,j] <- rbinom(1, 1, prob)
    }
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mb-y1
captured.Mb <- rowSums(y.all.Mb)>0
n.Mb <- sum(captured.Mb)
y.Mb <- y.all.Mb[captured.Mb,]
## prevcap[1:3,]
```


## Model *M~h~* -- Finite mixture

There are many flavors of $M_h$. This one assumes there are two (unknown) groups of individuals with different capture probs.

```{r}
#| label: sim-Mhm-pars
mixture.prob <- 0.6
group <- rbinom(N, 1, mixture.prob)    ## Two groups
p.Mh.mix <- ifelse(group==0, 0.2, 0.7) ## Two-point mixture
```
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mhm-ch
y.all.Mh.mix <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mh.mix[i,] <- rbinom(J, 1, p.Mh.mix[i])
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mhm-y1
captured.Mh.mix <- rowSums(y.all.Mh.mix)>0
n.Mh.mix <- sum(captured.Mh.mix)
y.Mh.mix <- y.all.Mh.mix[captured.Mh.mix,]
```

## Model *M~h~* -- Individual heterogeneity

Here's another flavor of $M_h$: logit-normal distribution for $p$.
```{r}
#| label: sim-Mh-pars
logit.p.bar <- -1  ## Mean p on logit scale
logit.p.var <- 1   ## SD of p on logit scale
p.h <- plogis(rnorm(N, logit.p.bar, logit.p.var))
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mh-ch
y.all.Mh <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mh[i,] <- rbinom(J, 1, p.h[i])
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mh-y1
captured.Mh <- rowSums(y.all.Mh)>0
n.Mh <- sum(captured.Mh)
y.Mh <- y.all.Mh[captured.Mh,]
#y.Mh[1:3,]
```






# Joint likelihood


## R package `RMark'

Before you can use 'RMark', you have to install MARK.
  
The main MARK page is [here]( http://www.phidot.org/software/mark/index.html)
  
[Instructions for Linux and Mac users](http://www.phidot.org/software/mark/rmark/linux/)
    
[Heather Gaya's tutorial](https://github.com/heathergaya/JAGS-NIMBLE-Tutorials/blob/master/Installing\%20RMark\%20on\%20Mac\%202022)
  
Lots of [good information about closed population mark-recapture](http://www.phidot.org/software/mark/docs/book/pdf/chap14.pdf)


## R package 'RMark' -- Model *M~0~*
  
Fit the model and set capture probability equal to recapture probability
($p=c$) using the `share` argument.

```{r}
#| label: RMark
#| warning: false
#| results: 'hide'
#| cache: false
#| warning: false
#| message: false
## install.packages("RMark") ## Must install MARK too!!
library(RMark)
y.ch <- data.frame(ch=apply(y, 1, paste, collapse=""))
mark.M0 <- mark(data=y.ch, model="Closed", silent=TRUE,
                model.parameters=list(p=list(formula=~1,share=TRUE)))
```

Estimates of $p$ and $n_0$ (called $f_0$ in MARK)
```{r}
#| label: RMark-pn0
mark.M0$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N
mark.M0$results$derived   
```


## R package `RMark' -- Model *M~t~*

```{r}
#| label: RMark-Mt
#| warning: false
#| results: 'hide'
#| cache: true
yt.ch <- data.frame(ch=apply(y.Mt, 1, paste, collapse=""))
mark.Mt <- mark(data=yt.ch, model="Closed", silent=TRUE,
                model.parameters=list(p=list(formula=~time,share=TRUE)))
```


Estimates of $p_j$ and $n_0$ (called $f_0$ in MARK)

```{r}
#| label: RMark-pn0-Mt
mark.Mt$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mt
mark.Mt$results$derived   
```




## R package `RMark' -- Model *M~b~*

```{r}
#| label: RMark-Mb
#| warning: false
#| results: 'hide'
#| cache: true
yb.ch <- data.frame(ch=apply(y.Mb, 1, paste, collapse=""))
mark.Mb <- mark(data=yb.ch, model="Closed", silent=TRUE) ## Default model
```


Estimates of $p$, $c$, and $n_0$ (called $f_0$ in MARK)
```{r}
#| label: RMark-pn0-Mb
#| size: 'scriptsize'
mark.Mb$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mb
#| size: 'scriptsize'
mark.Mb$results$derived   
```




## 'RMark' -- Model *M~h~*

This is a two-point mixture model. 

```{r}
#| label: RMark-Mh
#| warning: false
#| results: 'hide'
#| cache: false
yh.ch <- data.frame(ch=apply(y.Mh.mix, 1, paste, collapse=""))
mark.Mh <- mark(data=yh.ch, silent=TRUE, model="HetClosed", # "HugHet", 
                model.parameters=list(p=list(formula=~mixture,share=TRUE)))
```


Estimates of $p$ and $\pi$, the proportion of individuals in group 2.
```{r}
#| label: RMark-pn0-Mh
mark.Mh$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mh
mark.Mh$results$derived   
```






# Data augmentation


## Data augmentation

Data augmentation is strange.
  
We add "all zero" capture histories to our data, then we try to estimate how many of them are real.
  
The total number of observed and augmented encounter histories is $M$.
  
We define the model in terms of a binary indicator variable $z_i$, such that $N=\sum_{i=1}^M z_i$.
  
We know $z_i=1$ for the $n$ captured individuals.



## Data augmentation

We can think about the problem like this:
  
  
  \begin{tabular}{lcccc}
    \hline
               & \multicolumn{3}{c}{Occasion} <br>
    \cline{2-4}
    Individual & 1     & 2     & 3     & z    <br>
    \hline
    1          & 0     & 0     & 1     & 1    <br>
    2          & 1     & 1     & 1     & 1            <br>
    3          & 0     & 1     & 0     & 1            <br>
    4          & 0     & 1     & 1     & 1            <br>
    5          & 1     & 0     & 0     & 1            <br>
    \dots      & \dots & \dots & \dots      & 1        <br>
    $n$        & 0     & 0     & 1     & 1            <br>
    \hline
    $n+1$      & 0     & 0     & 0     & ?            <br>
    $n+2$      & 0     & 0     & 0     & ?            <br>
    \dots      & \dots & \dots & \dots     & ?        <br>
    $N$        & 0     & 0     & 0     & ?            <br>
    \hline
    $N+1$      & 0     & 0     & 0     & ?            <br>
    $N+2$      & 0     & 0     & 0     & ?            <br>
    \dots      & \dots & \dots & \dots     & ?        <br>
    $M$      & 0     & 0     & 0     & ?            <br>
    \hline
  \end{tabular}





## Data augmentation

The model:
$$
  z_i \sim \mathrm{Bern}(\psi) \\
  y_{ij} \sim \mathrm{Bern}(z_i \times p_{ij}) \\
  N=\sum_{i=1}^M z_i
$$
   
Note that this looks exactly like an occupancy model. <br>
   
But why bother with augmentation?
  
- DA works for [all] varieties of mark-recapture models
- Makes it easy to incorporate individual-covariates... including distance and location.
  


## Data augmentation


First, let's augment the data simulated under model $M_b$, with $M=200$.

```{r}
#| label: aug-y
M <- 200
y.aug <- matrix(0, M, J)
y.aug[1:nrow(y.Mb),] <- y.Mb
```
We also need to augment the `prevcap` object.

```{r}
#| label: un-prevcap
prevcap.aug <- matrix(FALSE, M, J)
prevcap.aug[1:nrow(y.Mb),] <- prevcap[captured.Mb,]
prevcap.aug <- ifelse(prevcap.aug, "Yes", "No")
```


## Bayesian data augmentation

For Bayesian inference, we need priors on $\psi$ and $p$. 
$$
  z_i \sim \mathrm{Bern}(\psi) \\
  y_{ij} \sim \mathrm{Bern}(z_i \times p) \\
  N = \sum_{i=1}^M z_i
$$
  
A uniform prior on $\psi$ results in a discrete uniform prior on $N$. We can change the prior for $N$ by changing the prior on $\psi$, recognizing:
$$
    E(N)=M\psi
$$


## Model *M~0~* -- data augmentation

```{r}
#| label: bugs-M0-aug
#| comment: ''
#| echo: -1
#| background: 'beige'
writeLines(readLines("M0-aug.jag"))
```




## Model *M~0~* -- data augmentation

Data
```{r}
#| label: jd-M0-aug
y.aug <- matrix(0, M, J)
y.aug[1:n,] <- y
jags.data.M0 <- list(y=y.aug, M=M, J=J)
```


Inits and parameters
```{r}
#| label: ji-M0-aug
ji.M0 <- function() list(z=rep(1,M), psi=runif(1), p=runif(1))
jp.M0 <- c("p", "psi", "N")
```
MCMC
```{r}
#| label: mcmc-M0-aug
#| results: 'hide'
#| warning: false
#| message: false
library(jagsUI)
jags.post.M0 <- jags.basic(data=jags.data.M0, inits=ji.M0,
                           parameters.to.save=jp.M0,
                           model.file="M0-aug.jag",
                           n.chains=3, n.adapt=100, n.burnin=0,
                           n.iter=2000, parallel=TRUE)
```


## Traceplots and density plots

```{r}
#| label: plot-mcmc-M0
#| out.width: "0.7<br>textwidth"
#| fig.align: 'center'
#| cache: true
plot(jags.post.M0[,jp.M0])
```



## Model *M~b~* -- data augmentation

```{r, echo=-1}
#| label: bugs-Mb-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("Mb-aug.jag"))
```

## Model *M~b~* -- data augmentation

Data, inits, parameters to monitor
```{r}
#| label: ji-Mb-aug
jags.data.Mb <- jags.data.M0
jags.data.Mb$y[1:nrow(y.Mb),] <- y.Mb
jags.data.Mb$prevcap <- ifelse(prevcap.aug=="Yes", 1, 0)
ji.Mb <- function() list(z=rep(1,M), psi=runif(1), p=runif(1), c=runif(1))
jp.Mb <- c("p", "c", "psi", "N")
```


MCMC
```{r}
#| label: mcmc-Mb-aug
#| results: 'hide'
jags.post.Mb <- jags.basic(data=jags.data.Mb, inits=ji.Mb,
                           parameters.to.save=jp.Mb,
                           model.file="Mb-aug.jag",
                           n.chains=3, n.adapt=100, n.burnin=0,
                           n.iter=2000, parallel=TRUE)
```



## Traceplots and density plots

```{r}
#| label: plot-mcmc-Mb
#| out.width: "0.7<br>textwidth"
#| fig.align: 'center'
#| cache: true
plot(jags.post.Mb[,jp.Mb])
```



## Model *M~h~* -- finite mixture

```{r, echo=-1}
#| label: bugs-Mhm-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("Mhm-aug.jag"))
```


## Model *M~h~* -- logit normal

```{r}
#| label: bugs-Mh-aug
#| comment: ''
#| echo: -1
#| background: 'beige'
writeLines(readLines("Mh-aug.jag"))
```


## Mark-recapture summary

We assume that variation in $p$ arises from changes over time, behavioral effects, or random individual heterogeneity.
   
Individual heterogeneity models pose estimation problems.
   
A solution is to use individual covariates to explain the variation, rather than random effects.  But what covariates are important? And how do we know the covariates values for uncaptured individuals?
   
With spatial capture-recapture (SCR) models we will use [the unobserved distance between animals and traps] as a covariate, and we'll use a spatial point process model for animal locations.



# Assignment



## Assignment
  
Create a self-contained R script or Rmarkdown file to do the following: 
  
- Import the Canada warbler mark-recapture data `CAWAcaps_2018.csv` and fit the following models in RMark and JAGS^[For JAGS, use $M=800$ but note if the posterior of $N$ pushes up against $M$.]:
  * Model $M_0$
  * Model $M_t$ 
  * Model $M_h$ -- Finite mixture
  * Model $M_h$ -- Logit-normal (JAGS only)
      
- Create a table reporting point estimates and 95% CIs for $N$ for each model fit in RMark and JAGS. Which model has the lowest AICc (RMark only)? 
  
Upload your `.R` or `.Rmd` file to ELC by 5:00pm on Tuesday.
