---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Non-spatial mark-recapture for <br> estimating population size'
    slide-number: true
    echo: true
lang: en
---



## {content-align="center" style="font-size: 140%;"}

<br>

<center>
  
**Non-spatial mark-recapture for <br> estimating population size**

<br> 
  
::: {style='font-size: 80%;'}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
:::
  
</center>





# Overview {visibility="hidden"}

<!-- ## Outline -->
  

## Mark-recapture overview

Until now, we've focused on models for data on aggregated quantities
like abundance and occupancy. 

<br> 
  
Now we will talk about data on uniquely identifiable individuals. 

<br>
  
::: {.fragment}
This is nice because all individuals are different, and we're often interested in these differences, even when our primary goal is inference on population-level parameters like abundance and density.
:::




## Closed populations

::: {style='font-size: 80%;'}
For this lecture, we will assume population closure:
  
- No births or deaths during the time of sampling (demographic closure)
- No permanent immigration or emigration (geographic closure)
  
::: {.fragment}
There are several other assumptions that depend on the specific model being employed, including:
  
- No misidentification
- No band/tag loss
- etc...
:::
  
::: {.fragment}
Later, we will relax the closure assumption so that we can study survival, recruitment, and movement.
:::
:::



## Mark-recapture overview

::: {style='font-size: 75%;'}
The simplest estimator of abundance is 
$$
    \hat{N} = \frac{n}{\hat{p}}
$$
where $n$ is the number of individuals detected, $p$ is detection probability, and $E(n)=Np$.
  
<br>

In distance sampling, we modeled detection probability as a function of distance, and we replaced $p$ with average detection probability. 
  
<br>

::: {.fragment}
For the simplest mark-recapture problems, $p$ is a constant (scalar) and we estimate it by trapping on multiple occasions and marking and recapturing as we go.
:::
:::



## Classical approaches

::: {style='font-size: 75%;'}
Lincoln-Peterson

- $J=2$ occasions
- Data are:
      
  * $n_1$ -- number of individuals captured on first occasion
  * $n_2$ -- number of individuals captured on second occasion
  * $m_2$ -- number of recaptures

Estimator is $\hat{N} = n_1n_2/m_2$

::: {.fragment}
Closed population models
  
- Usually better to have $J>2$ occasions
- Allows for assessment of behavioral effects, individual heterogeneity, etc...
:::
:::
  



## Mark-recapture data
  
::: {style='font-size: 80%;'}
Mark-recapture data are often called "capture histories." Here's an example with $n=6$ animals captured on $J=3$ occasions. 

::: {style='font-size: 60%;'}
<center>
Individual | 1 | 2 | 3
------  | -- | -- | --
1 | 0 | 0 | 1
2 | 1 | 1 | 1
3 | 0 | 1 | 0
4 | 0 | 1 | 1
5 | 1 | 0 | 0
6 | 0 | 0 | 1
</center>
:::
  
  
How do we estimate $p$ (and $N$) from these data? 
:::

<br>


<!-- You can get a sense for $p$ (and $N$) by looking at the data. -->


::: {.fragment}
::: {style='font-size: 60%;'}
Conceptually...
- If $p\approx 1$, most individuals would be captured on all 3 occasions, and few would go undetected
- If $p\approx 0$, few individuals would be captured  more than once, and many would go undetected
:::
:::


## [Closed population model (if we knew *N*)]{style="font-size: 85%;"}
  
::: {style='font-size: 75%;'}
**State model**
  
There isn't one! $N$ is often treated as a constant, although we sometimes put a Poisson or binomial prior on it.
  
  
::: {.fragment}
**Observation model** (supposing $N$ was known)
$$
    y_{ij} \sim \mathrm{Bernoulli}(p) \;\; \mathrm{for}\, i=1,\dots,N 
$$

- $N$ -- Population size
- $y_{ij}$ -- Capture histories, indicating if individual $i$ was captured on occasion $j$
- $p$ -- Capture probability
:::
  
::: {.fragment}
The problem with this formulation is that we don't observe the "all zero" encounter histories (and thus we don't know $N$).  As a result, the Bernoulli assumption isn't valid, and we need to develop a model that describes how many individuals we failed to capture.
:::
:::



## Closed population estimation options

::: {style='font-size: 75%;'}
**Conditional likelihood**
  
- Estimate $p$, and then estimate $N$ as a derived quantity. 
- Most software is based on the work by Huggins, which allows for continuous covariates. 
  
  
**Joint likelihood**
  
- Estimate $N$ and $p$ jointly
   
**Data augmentation**
  
- Tack on many "all zero" encounter histories and estimate how many of them actually occurred
- Usually, but not necessarily, used in Bayesian inference
:::
  


## Joint likelihood

  
::: {style='font-size: 75%;'}
The joint likelihood is based on the multinomial distribution, with
the "all-zero" encounter history being unobserved.
$$
    \{f_1, \dots, f_{K+1}\} \sim \mathrm{Multinomial}(N, \pi_1, \dots, \pi_{K+1}) \;\;
$$
where $\{f_1, \dots, f_K\}$ are the frequencies of the observable capture
histories.
:::
  
::: {.fragment}
::: {style='font-size: 60%;'}
Here's an example with $J=3$ capture occasions.  <br>

<center>
Capture history   |     Multinomial cell probability
----------------- | -----------------------------
      100                   | $\pi_1=p_1(1-p_2)(1-p_3)$
      010                   | $\pi_2=(1-p_1)p_2(1-p_3)$
      001                   | $\pi_3=(1-p_1)(1-p_2)p_3$
      110                   | $\pi_4=p_1p_2(1-p_3)$
      101                   | $\pi_5=p_1(1-p_2)p_3$
      011                   | $\pi_6=(1-p_1)p_2p_3$
      111                   | $\pi_7 = p_1p_2p_3$
      000                   | $\pi_0 = \prod_j 1-p_j$
</center>

:::
:::


## Model variations
  
::: {style='font-size: 75%;'}
More important than the estimation approach, is identification of the sources of variation in capture probability ($p$).
  
Otis et al. (1978) identified several model variations: 
  
- $M_0$ -- $p$ is constant
- $M_t$ -- unique $p$ for each capture occasion
- $M_b$ -- behavioral response with $p$ different than recapture probability $c$
- $M_h$ -- individual heterogeneity in $p$
- Finite mixture models
- Continuous mixture models (not available in RMark)
   
These can be combined, but beware of identifiability issues. See Otis et al. (1978) for details.
   
Later we'll talk about "individual covariate" models.
:::


## Software options
  
::: {style='font-size: 80%;'}
Program MARK

- Has the most model types
- Written for Windows, but can be executed on other platforms
  
R package 'RMark'
  
- Function `mark` runs MARK from R.

R package 'marked'
  
- Includes CJS and JS models
  
R package 'mra'
  
- Includes closed-population, CJS, and JS models
  
R package 'Rcapture'
  
- Includes closed-population
  
R package 'unmarked'
  
- Likelihood-based data augmentation with \inr{occu}
  
R package 'secr'
  
- Designed for spatially explicit models, but see \inr{closedN}
:::
  


# Simulation

```{r}
#| include: false
#| echo: false
set.seed(34889243)
```

## Model *M~0~*
  
::: {style='font-size: 80%;'}
Parameters
  
```{r}
#| label: sim-M0-pars
N <- 100
p <- 0.2
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-M0-ch
J <- 4  ## Occasions
y.all <- matrix(NA, N, J)
for(i in 1:N) {
    y.all[i,] <- rbinom(J, 1, p)
}
```
  
Observed capture histories (data)
  
```{r}
#| label: sim-M0-y1
captured <- rowSums(y.all)>0
(n <- sum(captured))
y <- y.all[captured,]
y[1:3,]
```
:::




## Summary stats

Capture history frequencies
```{r}
#| label: M0-hist
histories <- apply(y, 1, paste, collapse="")
sort(table(histories))
```

<br>

Detection frequencies
```{r}
#| label: M0-freq
y.tilde <- rowSums(y)
sort(table(y.tilde))
```


## Model *M~t~* -- Temporal variation

::: {style='font-size: 80%;'}
Capture probability for each occasion
```{r}
#| label: sim-Mt-pars
p.t <- c(0.3, 0.5, 0.2, 0.4)
```
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mt-ch
y.all.Mt <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mt[i,] <- rbinom(J, 1, p.t) }
```
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mt-y1
captured.Mt <- rowSums(y.all.Mt)>0
n.Mt <- sum(captured.Mt)
y.Mt <- y.all.Mt[captured.Mt,]
y.Mt[1:3,]
colSums(y.Mt)
```
:::



## Model *M~b~* -- Behavioral response

::: {style='font-size: 80%;'}
Capture probability for each occasion
```{r}
#| label: sim-Mb-pars
p.b <- 0.3
c <- 0.5  ## Trap happy
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mb-ch
y.all.Mb <- matrix(NA, N, J)
prevcap <- matrix(FALSE, N, J)
for(i in 1:N) {
    y.all.Mb[i,1] <- rbinom(1, 1, p.b)
    for(j in 2:J) {
        prevcap[i,j] <- any(y.all.Mb[i,1:(j-1)]>0)
        prob <- ifelse(prevcap[i,j], c, p.b)
        y.all.Mb[i,j] <- rbinom(1, 1, prob)
    }
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mb-y1
captured.Mb <- rowSums(y.all.Mb)>0
n.Mb <- sum(captured.Mb)
y.Mb <- y.all.Mb[captured.Mb,]
## prevcap[1:3,]
```
:::


## Model *M~h~* -- Finite mixture

::: {style='font-size: 80%;'}
There are many flavors of $M_h$. This one assumes there are two (unknown) groups of individuals with different capture probs.

```{r}
#| label: sim-Mhm-pars
mixture.prob <- 0.6
group <- rbinom(N, 1, mixture.prob)    ## Two groups
p.Mh.mix <- ifelse(group==0, 0.2, 0.7) ## Two-point mixture
```
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mhm-ch
y.all.Mh.mix <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mh.mix[i,] <- rbinom(J, 1, p.Mh.mix[i])
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mhm-y1
captured.Mh.mix <- rowSums(y.all.Mh.mix)>0
n.Mh.mix <- sum(captured.Mh.mix)
y.Mh.mix <- y.all.Mh.mix[captured.Mh.mix,]
```
:::


## Model *M~h~* -- Individual heterogeneity

::: {style='font-size: 80%;'}
Here's another flavor of $M_h$: logit-normal distribution for $p$.
```{r}
#| label: sim-Mh-pars
logit.p.bar <- -1  ## Mean p on logit scale
logit.p.var <- 1   ## SD of p on logit scale
p.h <- plogis(rnorm(N, logit.p.bar, logit.p.var))
```
  
  
All capture histories (for captured and uncaptured individuals)
  
```{r}
#| label: sim-Mh-ch
y.all.Mh <- matrix(NA, N, J)
for(i in 1:N) {
    y.all.Mh[i,] <- rbinom(J, 1, p.h[i])
}
```
  
  
Observed capture histories (data)
  
```{r}
#| label: sim-Mh-y1
captured.Mh <- rowSums(y.all.Mh)>0
n.Mh <- sum(captured.Mh)
y.Mh <- y.all.Mh[captured.Mh,]
#y.Mh[1:3,]
```
:::






# Joint likelihood


## R package `RMark'

Before you can use 'RMark', you have to install MARK.

<br>

The main MARK page is [here]( http://www.phidot.org/software/mark/index.html).

<br>

[Instructions for Linux and Mac users](http://www.phidot.org/software/mark/rmark/linux/)
    
<br>

[Heather Gaya's tutorial](https://github.com/heathergaya/JAGS-NIMBLE-Tutorials/blob/master/Installing\%20RMark\%20on\%20Mac\%202022)
  
<br>

Lots of [good information about closed population mark-recapture](http://www.phidot.org/software/mark/docs/book/pdf/chap14.pdf)


## R package 'RMark' -- Model *M~0~*
  
::: {style='font-size: 80%;'}
Fit the model and set capture probability equal to recapture probability
($p=c$) using the `share` argument.

```{r}
#| label: RMark
#| warning: false
#| results: 'hide'
#| cache: false
#| message: false
## install.packages("RMark") ## Must install MARK too!!
library(RMark)
y.ch <- data.frame(ch=apply(y, 1, paste, collapse=""))
mark.M0 <- mark(data=y.ch, model="Closed", silent=TRUE,
                model.parameters=list(p=list(formula=~1,share=TRUE)))
```

Estimates of $p$ and $n_0$ (called $f_0$ in MARK)
```{r}
#| label: RMark-pn0
mark.M0$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N
mark.M0$results$derived   
```
:::


## R package `RMark' -- Model *M~t~*

```{r}
#| label: RMark-Mt
#| warning: false
#| results: 'hide'
#| cache: true
yt.ch <- data.frame(ch=apply(y.Mt, 1, paste, collapse=""))
mark.Mt <- mark(data=yt.ch, model="Closed", silent=TRUE,
                model.parameters=list(p=list(formula=~time,share=TRUE)))
```


Estimates of $p_j$ and $n_0$ (called $f_0$ in MARK)

```{r}
#| label: RMark-pn0-Mt
mark.Mt$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mt
mark.Mt$results$derived   
```




## R package `RMark' -- Model *M~b~*

::: {style='font-size: 80%;'}
```{r}
#| label: RMark-Mb
#| warning: false
#| results: 'hide'
#| cache: true
yb.ch <- data.frame(ch=apply(y.Mb, 1, paste, collapse=""))
mark.Mb <- mark(data=yb.ch, model="Closed", silent=TRUE) ## Default model
```


Estimates of $p$, $c$, and $n_0$ (called $f_0$ in MARK)
```{r}
#| label: RMark-pn0-Mb
#| size: 'scriptsize'
mark.Mb$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mb
#| size: 'scriptsize'
mark.Mb$results$derived   
```
:::




## 'RMark' -- Model *M~h~*

::: {style='font-size: 80%;'}
This is a two-point mixture model. 

```{r}
#| label: RMark-Mh
#| warning: false
#| results: 'hide'
#| cache: false
yh.ch <- data.frame(ch=apply(y.Mh.mix, 1, paste, collapse=""))
mark.Mh <- mark(data=yh.ch, silent=TRUE, model="HetClosed", # "HugHet", 
                model.parameters=list(p=list(formula=~mixture,share=TRUE)))
```


Estimates of $p$ and $\pi$, the proportion of individuals in group 2.
```{r}
#| label: RMark-pn0-Mh
mark.Mh$results$real      
```


Estimate of $N$ 
```{r}
#| label: RMark-N-Mh
mark.Mh$results$derived   
```
:::



# Data augmentation


## Data augmentation

::: {style='font-size: 80%;'}
Data augmentation is strange.

<br>

We add "all zero" capture histories to our data, then we try to estimate how many of them are real.

<br>

::: {.fragment}
The total number of observed and augmented encounter histories is $M$. We define the model in terms of a binary indicator variable $z_i$, such that $N=\sum_{i=1}^M z_i$.

<br>

We know $z_i=1$ for the $n$ captured individuals.
:::
:::



## Data augmentation

::: {style='font-size: 80%;'}
We can think about the problem like this:
:::

<center>
::: {style='font-size: 40%;'}
Individual | 1     | 2     | 3     | z
--------- | ----- | ------ | ----- | -----
1          | 0     | 0     | 1     | 1
2          | 1     | 1     | 1     | 1
3          | 0     | 1     | 0     | 1
4          | 0     | 1     | 1     | 1
5          | 1     | 0     | 0     | 1
$\dots$    | $\dots$ | $\dots$ | $\dots$      | 1
$n$        | 0     | 0     | 1     | 1
$n+1$      | 0     | 0     | 0     | ?
$n+2$      | 0     | 0     | 0     | ?
$\dots$      | $\dots$ | $\dots$ | $\dots$     | ?
$N$        | 0     | 0     | 0     | ?
$N+1$      | 0     | 0     | 0     | ?
$N+2$      | 0     | 0     | 0     | ?
$\dots$      | $\dots$ | $\dots$ | $\dots$     | ?
$M$      | 0     | 0     | 0     | ?
:::
</center>




## Data augmentation

::: {style='font-size: 80%;'}
The model:
$$
  z_i \sim \mathrm{Bern}(\psi) \\
  y_{ij} \sim \mathrm{Bern}(z_i \times p_{ij}) \\
  N=\sum_{i=1}^M z_i
$$
   
Note that this looks exactly like an occupancy model.
   
::: {.fragment}
But why bother with augmentation?
  
- DA works for *all* varieties of mark-recapture models
- Makes it easy to incorporate individual-covariates... including distance and location.
:::
:::


## Data augmentation


::: {style='font-size: 90%;'}
First, let's augment the data simulated under model $M_b$, with $M=200$.

```{r}
#| label: aug-y
M <- 200
y.aug <- matrix(0, M, J)
y.aug[1:nrow(y.Mb),] <- y.Mb
```

<br>

We also need to augment the `prevcap` object.

```{r}
#| label: un-prevcap
prevcap.aug <- matrix(FALSE, M, J)
prevcap.aug[1:nrow(y.Mb),] <- prevcap[captured.Mb,]
prevcap.aug <- ifelse(prevcap.aug, "Yes", "No")
```
:::



## Bayesian data augmentation

::: {style='font-size: 80%;'}
For Bayesian inference, we need priors on $\psi$ and $p$. 
$$
  z_i \sim \mathrm{Bern}(\psi) \\
  y_{ij} \sim \mathrm{Bern}(z_i \times p) \\
  N = \sum_{i=1}^M z_i
$$

<br>

::: {.fragment}
A uniform prior on $\psi$ results in a discrete uniform prior on $N$. We can change the prior for $N$ by changing the prior on $\psi$, recognizing:
$$
    E(N)=M\psi
$$
:::

:::

## Model *M~0~* -- data augmentation

[M0-aug.jag](M0-aug.jag)


```{r, echo=-1}
#| label: bugs-M0-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("M0-aug.jag"))
```




## Model *M~0~* -- data augmentation

Data
```{r}
#| label: jd-M0-aug
y.aug <- matrix(0, M, J)
y.aug[1:n,] <- y
jags.data.M0 <- list(y=y.aug, M=M, J=J)
```


Inits and parameters
```{r}
#| label: ji-M0-aug
ji.M0 <- function() list(z=rep(1,M), psi=runif(1), p=runif(1))
jp.M0 <- c("p", "psi", "N")
```
MCMC
```{r}
#| label: mcmc-M0-aug
#| results: 'hide'
#| warning: false
#| message: false
library(jagsUI)
jags.post.M0 <- jags.basic(data=jags.data.M0, inits=ji.M0,
                           parameters.to.save=jp.M0,
                           model.file="M0-aug.jag",
                           n.chains=3, n.adapt=100, n.burnin=0,
                           n.iter=2000, parallel=TRUE)
```


## Traceplots and density plots

```{r}
#| label: plot-mcmc-M0
#| out-width: "50%"
#| fig-align: 'center'
#| cache: false
plot(jags.post.M0[,jp.M0])
```



## Model *M~b~* -- data augmentation

[Mb-aug.jag](Mb-aug.jag)

```{r, echo=-1}
#| label: bugs-Mb-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("Mb-aug.jag"))
```

## Model *M~b~* -- data augmentation

Data, inits, parameters to monitor
```{r}
#| label: ji-Mb-aug
jags.data.Mb <- jags.data.M0
jags.data.Mb$y[1:nrow(y.Mb),] <- y.Mb
jags.data.Mb$prevcap <- ifelse(prevcap.aug=="Yes", 1, 0)
ji.Mb <- function() list(z=rep(1,M), psi=runif(1), p=runif(1), c=runif(1))
jp.Mb <- c("p", "c", "psi", "N")
```


MCMC
```{r}
#| label: mcmc-Mb-aug
#| results: 'hide'
jags.post.Mb <- jags.basic(data=jags.data.Mb, inits=ji.Mb,
                           parameters.to.save=jp.Mb,
                           model.file="Mb-aug.jag",
                           n.chains=3, n.adapt=100, n.burnin=0,
                           n.iter=2000, parallel=TRUE)
```



## Traceplots and density plots

```{r}
#| label: plot-mcmc-Mb
#| out-width: "50%"
#| fig-align: 'center'
#| cache: false
plot(jags.post.Mb[,jp.Mb])
```



## Model *M~h~* -- finite mixture

```{r, echo=-1}
#| label: bugs-Mhm-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("Mhm-aug.jag"))
```


## Model *M~h~* -- logit normal

```{r, echo=-1}
#| label: bugs-Mh-aug
#| comment: ''
#| background: 'beige'
writeLines(readLines("Mh-aug.jag"))
```


## Mark-recapture summary

::: {style='font-size: 80%;'}
We assume that variation in $p$ arises from changes over time, behavioral effects, or random individual heterogeneity.

<br>

::: {.fragment}
Individual heterogeneity models pose estimation problems.
:::
   
<br>

::: {.fragment}
A solution is to use individual covariates to explain the variation, rather than random effects.  But what covariates are important? And how do we know the covariates values for uncaptured individuals?
:::
   
<br>

::: {.fragment}
With spatial capture-recapture (SCR) models we will use *the unobserved distance between animals and traps* as a covariate, and we'll use a spatial point process model for animal locations.
:::
:::



# Assignment

## Assignment
  
::: {style='font-size: 70%;'}
Create a self-contained R script or Rmarkdown file to do the following: 
  
- Import the Canada warbler mark-recapture data `CAWAcaps_2018.csv` and fit the following models in RMark and JAGS^[For JAGS, use $M=800$ but note if the posterior of $N$ pushes up against $M$.]:
  * Model $M_0$
  * Model $M_t$ 
  * Model $M_h$ -- Finite mixture
  * Model $M_h$ -- Logit-normal (JAGS only)
      
- Create a table reporting point estimates and 95% CIs for $N$ for each model fit in RMark and JAGS. Which model has the lowest AICc (RMark only)? 

<center>
Upload your `.R` or `.Rmd` file to ELC by 5:00pm on Tuesday.
</center>
:::
