---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Dog'
    slide-number: true
    css: ../../custom.css
lang: en
---





## {content-align="center" style="font-size: 160%;"}

<center>

Occupancy models:<br>covariates and prediction 

<br>

::: {style="font-size: 80%;"}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics <br>
:::

</center>
  

# Overview {visibility="hidden"}




## Overview

::: {style="font-size: 90%;"}

<!-- We almost always have covariates, which are also called explanatory variables, predictors, independent variables, etc\dots <br>  -->
  
  
Explaining variation in occupancy and detection probability with covariates is central to reliable prediction and scientific investigation. <br>

<br>

::: {.fragment}
In static occupancy models, there are two types of covariates:
  
- [Site covariates]{style="color: blue;"} vary among sites
      but not among occasions.
- [Observation covariates]{style="color: Coral;"} may vary among
      sites and occasions
:::

<br>

::: {.fragment}
In both cases, covariates can be continuous or categorical.
:::
  
::: {.fragment}
Observation covariates [cannot](style="color: red;") be used to describe variation in occupancy because occupancy is assumed to be constant over time. <br> 
:::

:::

## Covariates

::: {style="font-size: 90%;"}

State model
$$
  \small
  \begin{gather*}
    \mathrm{logit}(\psi_i) = \beta_0 + \beta_1 {\color{blue} x_{i1}} +
    \beta_2 {\color{blue} x_{i2}} + \cdots \\
    z_i \sim \mathrm{Bern}(\psi_i)
  \end{gather*}
$$

Observation model
$$
  \small
  \begin{gather*}
    \mathrm{logit}(p_{ij}) = \alpha_0 + \alpha_1 {\color{blue} x_{i1}}
    + \alpha_2 {\color{Coral} w_{ij}} + \cdots \\
    y_{ij} \sim \mathrm{Bern}(z_i\times p_{ij})
  \end{gather*}
$$

$\color{blue} x_1$ and $\color{blue} x_2$ are site covariates. <br>
Examples: elevation, forest cover, distance to water.
  
  
$\color{Coral} w$ is an observation covariate. <br>
Examples: time of day, observer, wind.

:::

# Simulation {visibility="hidden"}



## Simulation

::: {style="font-size: 80%;"}

Two continuous covariates and one categorical covariate with 2 levels.
  
```{r}
#| label: sim-cov1
#| echo: true
nSites <- 200; nVisits <- 4; set.seed(83) # Make it reproducible
elev <- rnorm(nSites,5000,500) # Site covariate (continuous)
forest <- rnorm(nSites,100,10) # Another continuous site covariate
temp <- matrix(sample(c("Cold", "Hot"), size=nSites*nVisits,
                      replace=TRUE),
               nrow=nSites, ncol=nVisits) # Categorical obs covar
tempHot <- ifelse(temp=="Hot", 1, 0)      # Dummy variable
```

::: {.fragment}
Coefficients, $\psi$, and $p$
```{r}
#| label: sim-cov2
#| echo: true
beta0 <- 5; beta1 <- -0.001; beta2 <- 0
psi <- plogis(beta0 + beta1*elev + beta2*forest)
alpha0 <- -6.5; alpha1 <- 0.001; alpha2 <- 2
p <- plogis(alpha0 + alpha1*elev + alpha2*tempHot)
```
:::

::: {.fragment}
Simulate occupancy and detection data
```{r}
#| label: sim-cov3
#| echo: true
z <- rbinom(nSites, size=1, psi)            ## pres/absence
y <- matrix(NA, nrow=nSites, ncol=nVisits)
for(i in 1:nSites) {
    y[i,] <- rbinom(nVisits, size=1, prob=z[i]*p[i,]) }
```
:::

:::






## Prediction in 'unmarked'

```{r}
#| label: sim-sitch
#| fig-width: 7
#| fig-height: 5
#| out-width: '90%'
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))  
plot(function(x) plogis(beta0+beta1*x), from=min(elev), to=max(elev),
     ylab="Probability", col="blue", ylim=0:1, xlab="Elevation", lwd=2)
plot(function(x) plogis(alpha0+alpha1*x), from=min(elev), to=max(elev),
     col="orange", add=TRUE)
legend(5000, 1, c("Occupancy (psi)", "Detection (p)"), lty=c(1, 1),
       col=c("blue", "orange"), lwd=2, cex=1.2)
```



Expect problems if you ignore imperfect detection in this case. <br>




## Simulated data

::: {.columns}
::: {.column width='40%'}
  Observations

```{r}
#| label: sim-nocov-dat
#| echo: true
y[1:20,]
``` 

:::
::: {.column width='5%'}
:::
::: {.column width='55%'}

::: {style="font-size: 80%;"}

Summary stats <br>

Detections at each site <br>

```{r}
#| echo: true
#| label: sim-nocov-ss1
siteDets <- rowSums(y) # Dets at each site
table(siteDets)        # Frequency
```

Proportion of sites known to be occupied <br>
```{r}
#| label: sim-nocov-ss2
#| echo: true
naiveOccupancy <- sum(siteDets>0)/nSites
naiveOccupancy 
```
```{r}
#| label: un
#| include: false
library(unmarked)
``` 
Actual occupancy <br>
```{r}
#| label: sim-z
#| echo: true
sum(z) / nSites
```

:::
:::

:::


# Prediction {visibility="hidden"}




<!-- \subsection{Likelihood-based methods} -->




## Prepare data in 'unmarked'


Notice the two new arguments `siteCovs` and `obsCovs`: 
```{r}
#| label: un-umf
#| echo: true
#| warning: false
umf <- unmarkedFrameOccu(y=y, siteCovs=data.frame(elev,forest),
                         obsCovs=list(temp=temp))
```


```{r}
#| label: wfac
#| echo: true
summary(umf)
```


## Standardizing

::: {style="font-size: 90%;"}


It's almost always a good idea to standardize [continuous]{style="color: red;"} covariates. <br>
  
Standardizing involves subtracting the mean and then dividing by the standard deviation. <br>
  

```{r}
#| label: umf-zcov
#| echo: true
mean.elev <- mean(elev); mean.forest <- mean(forest)
sd.elev <- sd(elev); sd.forest <- sd(forest)
elev_s <- (elev-mean.elev)/sd.elev
forest_s <- (forest-mean.forest)/sd.forest
siteCovs(umf)$elev_s <- elev_s
siteCovs(umf)$forest_s <- forest_s
```

Standardizing makes it easier to find the maximum likelihood estimates (MLEs) and it facilitates comparison of estimates.  <br> 
  
We just have to remember to back-transform covariates when graphing predictions.


:::


## Fit the model

```{r}
#| label: un-fit
#| echo: true
fm <- occu(~elev_s+temp ~elev_s+forest_s, umf) # Standardized covariates
fm
```



## Prediction in 'unmarked'

::: {style="font-size: 90%;"}


Create a data frame with prediction covariates. We'll let elevation
  vary while holding the other two covariates constant. 

Important that we predict using the standardized covariates. We can
still graph on the original scales.

::: {.fragment}
```{r}
#| label: preddat
#| echo: true
pred.data <- data.frame(elev_s=seq(from=-3, to=3, length=50),
                        forest_s=0, temp='Hot') # Standardized
pred.data$elev <- pred.data$elev_s*sd.elev+mean.elev # Orig scale
pred.data$forest <- pred.data$forest_s*sd.forest+mean.forest 
```
:::


::: {.fragment}
Get predictions of $\psi$ for each row of prediction data.
```{r}
#| label: predpsi
#| echo: true
psi.pred <- predict(fm, newdata=pred.data,
                    type='state', append=TRUE)
```
:::


::: {.fragment}
Get predictions of $p$ for each row of prediction data.
```{r}
#| label: predp
#| echo: true
p.pred <- predict(fm, newdata=pred.data,
                  type='det', append=TRUE)
```
:::

:::


## Prediction in 'unmarked'


View $\psi$ predictions
```{r}
#| label: psi-head
#| echo: true
print(head(psi.pred), digits=2)
```


View $p$ predictions
```{r}
#| label: p-head
#| echo: true
print(head(p.pred), digits=2)
```


## Standardized elevation

```{r}
#| label: pred-psi1
#| fig-width: 7
#| fig-height: 5
#| out-width: '80%'
#| fig-align: 'center'
## #| echo: "[-1]"
par(mai=c(0.9,0.9,0.1,0.1))  
plot(Predicted ~ elev_s, psi.pred, type="l", ylab="Occurrence probability",
     col="blue", xlab="Elevation (standardized)", ylim=0:1) 
lines(lower ~ elev_s, psi.pred, col="grey"); lines(upper ~ elev_s, psi.pred, col="grey")
```



## Elevation on original scale

```{r}
#| label: pred-psi1s
#| fig-width: 7
#| fig-height: 5
#| out-width: '80%'
#| fig-align: 'center'
## #| echo: "[-1]"
par(mai=c(0.9,0.9,0.1,0.1))  
plot(Predicted ~ elev, psi.pred, type="l", ylab="Occurrence probability",
     col="blue", xlab="Elevation (original scale)", ylim=0:1)
lines(lower ~ elev, psi.pred, col="grey"); lines(upper ~ elev, psi.pred, col="grey")
```





## Estimated vs actual occupancy

```{r, psi-actual-est,fig.width=7,fig.height=5,out.width='90%',fig.align='center',echo=FALSE}
par(mai=c(0.9,0.9,0.1,0.1))  
plot(Predicted ~ elev, psi.pred, type="l", ylab="Occurrence probability", col="blue",
     xlab="Elevation (original scale)", ylim=0:1)
lines(lower ~ elev, psi.pred, col="grey"); lines(upper ~ elev, psi.pred, col="grey")
plot(function(x) plogis(beta0 + beta1*x),
     from=min(psi.pred$elev), to=max(psi.pred$elev), add=TRUE,
     lwd=2)
legend(5800, 1, c("Actual", "Estimated"), lwd=c(2,1), col=c("black", "blue"))
```



<!-- \subsection{Bayesian methods} -->




## The BUGS model

```{r}
#| label: bugs
#| comment: ''
writeLines(readLines("occupancy-model-covs.jag"))
```





## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data
#| echo: true
jags.data <- list(y=y, elev=elev_s, forest=forest_s,
                  tempHot=tempHot,
                  nSites=nSites, nOccasions=nVisits)
```


Initial values
  
```{r}
#| label: bugs-inits
#| echo: true
jags.inits <- function() {
    list(beta0=rnorm(1), alpha0=rnorm(1), z=rep(1, nSites))
}
```


Parameters to monitor
  
```{r}
#| label: bugs-pars
#| echo: true
jags.pars <- c("beta0", "beta1", "beta2",
               "alpha0", "alpha1", "alpha2",
               "sitesOccupied")
```



## MCMC

```{r}
#| label: bugs-mcmc
#| echo: true
#| message: false
#| cache: false
library(jagsUI)
jags.post.samples <- jags.basic(data=jags.data, inits=jags.inits,
                                parameters.to.save=jags.pars,
                                model.file="occupancy-model-covs.jag",
                                n.chains=3, n.adapt=100, n.burnin=0,
                                n.iter=2000, parallel=TRUE)
```

## Summarize output

```{r}
#| label: bugs-sum
#| echo: true
summary(jags.post.samples)
```


## Traceplots and density plots

```{r}
#| label: bugs-plot1
#| out-width: "70%"
#| fig-align: 'center'
plot(jags.post.samples[,1:3])
```



## Traceplots and density plots

```{r}
#| label: bugs-plot2
#| out-width: "70%"
#| fig-width: 5
#| fig-height: 6
#| fig-align: 'center'
plot(jags.post.samples[,c(4:6,8)])
```



## Bayesian prediction

Every MCMC iteration represents a sample from the posterior distribution.

<br>
  
Each sample of the occupancy and detection coefficients can be used to make a prediction.
  
<br>
  
The collection of predictions can be used to summarize the posterior  predictive distribution.

<br>

We will look at the distribution of prediction lines.




## Bayesian prediction

First, extract the $\psi$ coefficients
```{r}
#| label: psi-coefs
#| echo: true
psi.coef.post <- as.matrix(jags.post.samples[,c("beta0","beta1","beta2")])
head(psi.coef.post, n=4)
```

Create prediction matrix, one row for each MCMC iteration.

```{r}
#| label: psi-predmat
#| echo: true
n.iter <- nrow(psi.coef.post)  
psi.post.pred <- matrix(NA, nrow=n.iter, ncol=nrow(pred.data))
```

Predict $\psi$ for each MCMC iteration.

```{r}
#| label: psi-pred-bayes
#| echo: true
for(i in 1:n.iter) {
    psi.post.pred[i,] <- plogis(psi.coef.post[i,"beta0"] +
                                psi.coef.post[i,"beta1"]*pred.data$elev_s)
}
```




## Bayesian prediction

Prediction line for first posterior sample
```{r}
#| label: psi-pred1
#| echo: true
#| fig-align: 'center'
#| out-width: '80%'
#| fig-height: 5
#| dev: 'png'
#| dpi: 200
plot(pred.data$elev, psi.post.pred[1,], type="l", xlab="Elevation",
     ylab="Occurrence probability", ylim=c(0, 1), col=gray(0.8))
```


## Bayesian prediction

All samples from the posterior predictive distribution
```{r,echo=-1}
#| label: psi-pred-post
#| fig-align: 'center'
#| out-width: '80%'
#| fig-height: 5
#| dev: 'png'
#| dpi: 200
#| cache: false
plot(pred.data$elev, psi.post.pred[1,], type="l", xlab="elev",
     ylab="Occurrence probability", ylim=c(0, 1), col=gray(0.8))
for(i in 1:n.iter) {
    lines(pred.data$elev, psi.post.pred[i,], col=gray(0.8))
}
```




## Bayesian prediction

Now with posterior mean and 95\% CI
```{r,echo=-(1:2)}
#| label: psi-pred-post-meanCI
#| fig-align: 'center'
#| out-width: '80%'
#| fig-height: 5
#| dpi: 200
#| cache: false
plot(pred.data$elev, psi.post.pred[1,], type="l", xlab="elev",
     ylab="Occurrence probability", ylim=c(0, 1), col=gray(0.8))
for(i in 1:n.iter) {
    lines(pred.data$elev, psi.post.pred[i,], col=gray(0.8))
}
pred.post.mean <- colMeans(psi.post.pred)
pred.post.lower <- apply(psi.post.pred, 2, quantile, prob=0.025)
pred.post.upper <- apply(psi.post.pred, 2, quantile, prob=0.975)
lines(pred.data$elev, pred.post.mean, col="blue")
lines(pred.data$elev, pred.post.lower, col="blue", lty=2)
lines(pred.data$elev, pred.post.upper, col="blue", lty=2)
``` 




## Summary of static occupancy models

::: {style="font-size: 80%;"}

Assumptions are made explicit by the model description, but it's worth emphasizing that: 

*  Occupancy is assumed to remain constant at each site during the sampling period
* The definition of a 'site' is critical, but can difficult in some situations, such as in camera studies.
* Abundance is assumed to be unimportant
* Observations are assumed to be independent, conditional on latent variables and covariates

These assumptions can be relaxed if we have enough data. 

* Dynamic occupancy models

* Occupancy evolves as a function of local colonization and extinction

* $N$-mixture models

* Focus is on local abundance, rather than presence/absence


We'll talk more about design issues later

:::


# Assignment {visibility="hidden"}



## Assignment

::: {style="font-size: 80%;"}

Create a self-contained R script or Rmarkdown file to do the following:

1. Fit an occupancy model in 'unmarked' to the Ruffed Grouse data (`"grouse_data.csv"`). Use all three covariates, with elevation and forest cover affecting occupancy and day of the year affecting detection probability.
  * Response: `grouse1, grouse2, grouse3`
  * Site covs: Elevation (`elev`), forest cover (`fc`)
  * Obs cov: Day of the year (`doy.1, doy.2, doy.3`)
      
2. Make one prediction graph for occupancy and one for detection.

3. Create the same graphs as above, but using Bayesian methods.
    
<center>
    Upload your .R, .Rmd, or .qmd file to ELC by 5:00pm on Tuesday. 
</center>

:::
