---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Multinomial $N$-mixture models'
    slide-number: true
    echo: true
lang: en
---


## {content-align="center" style="font-size: 140%;"}

<br>

<center>
Lecture 7 -- Multinomial $N$-mixture models: <br> simulation, fitting, and prediction

<br>

::: {style='font-size: 80%;'}
WILD(FISH) 8390 <br>
Inference for Models of Fish and Wildlife Population Dynamics
:::

</center>




# Overview {visibility="hidden"}


## Overview

::: {style='font-size: 80%;'}
The objectives are the same as with binomial $N$-mixture models.
  
- Estimate abundance
- Model spatial variation in abundance/density
  
::: {.fragment}
The difference is that, instead of repeated count data at each site,
we have data from sampling methods such as:
  
- Double observer sampling
- Removal sampling
- Mark-recapture
- Distance sampling
- And other methods that yield multinomial counts\dots
:::

:::
  

## Multinomial $N$-mixture model
  
::: {style='font-size: 70%;'}
State model (with Poisson assumption)
$$
  \mathrm{log}(\lambda_i) = \beta_0 + \beta_1 {\color{blue} x_{i1}} + \beta_2 {\color{blue} x_{i2}} + \cdots \\
  N_i \sim \mathrm{Poisson}(\lambda_i)
$$
  
  
Observation model
$$
  \mathrm{logit}(p_{ij}) = \alpha_0 + \alpha_1 {\color{blue} x_{i1}} + \alpha_2 {\color{Purple} w_{ij}} + \cdots \\
  \{y_{i1}, \dots, y_{iK}\}  \sim \mathrm{Multinomial}(N_i, \pi(p_{i1}, \dots, p_{iJ}))
$$
  
Definitions
:::

::: {style='font-size: 60%;'}
$\lambda_i$ -- Expected value of abundance at site $i$ <br>
$N_i$ -- Realized value of abundance at site $i$ <br>
$p_{ij}$ -- Probability of detecting *an individual* at site $i$ on occasion $j$ <br>
$\pi(p)$ -- Function mapping $J$ detection probs to
$K$ multinomial cell probs <br>
$y_{ik}$ -- Multinomial count data (*The final count is not observed!*) <br>
$\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates
$\color{Purple} w$ -- observation covariate
:::



## The multinomial distribution
  
::: {style='font-size: 70%;'}
This is the first multivariate distribution we've covered.
$$
  \{y_{1}, \dots, y_{K}\} \sim \mathrm{Multinomial}(N, \{\pi_i, \dots, \pi_K\})
$$
  
The multinomial describes how $N$ objects are distributed among $K$ classes (also called bins, categories, etc\dots).
  
:::

<br>

::: {.fragment}
::: {style='font-size: 65%;'}
Imagine $N=20$ animals are studied for 1 year and there are 3 possible "fates" with associated probabilities: $\Pr(\mathrm{survived})=0.5$, $\Pr(\mathrm{depredated})=0.3$, $\Pr(\mathrm{starved})=0.2$.
  
Here is one possible outcome:
:::


::: {style='font-size: 85%;'}

```{r}
#| label: multi-sim-concept
#| size: 'footnotesize'
N <- 20
pi <- c(survived=0.5, depredated=0.3, starved=0.2) ## Probs must sum to 1
t(rmultinom(n=1, size=N, prob=pi))                 ## Counts must sum to N
```
:::
:::



## Sampling methods

::: {style='font-size: 70%;'}
Benefits of the multinomial $N$-mixture model
  
- Precision is typically better than binomial $N$-mixture models because there's more information in the data.
- Can be applied to data from many sampling designs, simply by changing how the $\pi$ probabilities are computed.

:::

::: {.fragment}
::: {style='font-size: 65%;'}
Removal sampling
$$
  {\pi(p)} = \{p, (1-p)p, (1-p)^2p, \dots, (1-p)^{K-1}p, (1-p)^K\}
$$

Double observer (*independent observers*)
$$
  {\pi(p)} = \{p_1(1-p_2), (1-p_1)p_2, p_1p_2, (1-p_1)(1-p_2)\}
$$

Double observer (*dependent observers*)
$$
  {\pi(p)} = \{p_1, (1-p_1)p_2, (1-p_1)(1-p_2)\}
$$

:::
:::



# Removal sampling

## Removal sampling
  
::: {style='font-size: 70%;'}
Removal sampling is often used in electrofishing studies.
  
A stream section is surveyed $J$ times, fish are removed on each "pass", and the rate of removal tells us about capture probability.
  
<!-- Definitions -->
  
:::
::: {.fragment}
::: {style='font-size: 70%;'}
- $y_{ij}$ -- number of individuals removed at site $i$ on pass $j$
- $p$ -- probability of catching an individual on a single pass
:::

::: {style='font-size: 60%;'}

  Description                         |  Multinomial cell probability
  ----------------------------------- | ------------------------------
  Pr(first captured on first pass)    |         $\pi_1 = p$
  Pr(first captured on second pass)   |       $\pi_2 = (1-p)p$
  Pr(first captured on third pass)    |   $\pi_3 = (1-p)(1-p)p$
  $\cdots$                            |           $\cdots$
  Pr(first captured on pass $J$)      |    $\pi_J = (1-p)^{J-1}p$
  Pr(not captured)                    |    $\pi_{J+1} = (1-p)^J$

:::
:::

```{r}
#| include: false
set.seed(34889243)
```



## [Simulating removal data, no covariates]{style="font-size: 90%;"}
  
::: {style='font-size: 80%;'}
Abundance
```{r, echo=-1}
#| label: sim-rem-nocov1
set.seed(430)
nSites <- 100
lambda1 <- 4.6  ## Expected value of N
N1 <- rpois(n=nSites, lambda=lambda1)
```

Capture probability and multinomial counts
```{r}
#| label: sim-nocov2
nPasses <- 3
K <- nPasses+1  # multinomial cells
p1 <- 0.4
pi1 <- c(p1, (1-p1)*p1, (1-p1)*(1-p1)*p1, (1-p1)^3)
y1.all <- matrix(NA, nrow=nSites, ncol=K)
for(i in 1:nSites) {
    y1.all[i,] <- rmultinom(n=1, size=N1[i], prob=pi1)    }
```
  
Discard final column of individuals not detected
```{r}
#| label: N1y1
y1 <- y1.all[,-K]
head(y1, n=3)
```
:::


## Simulated removal data

::: {style='font-size: 70%;'}

```{r}
#| label: tbl-kable
#| echo: false
colnames(y1) <- c("Pass 1", "Pass 2", "Pass 3")
rownames(y1) <- paste("Site", 1:nrow(y1))
knitr::kable(y1[1:10,], format="html", caption="")
```

:::


## Removal model, covariates
  
::: {style='font-size: 80%;'}
Covariates
  
```{r}
#| label: sim-cov1
#| size: 'scriptsize'
streamDepth <- rnorm(nSites)
```
Coefficients, $\lambda$, and $p$
  
```{r, echo=1:4}
#| label: nsim-cov2
beta0 <- 1; beta1 <- 0.5
lambda2 <- exp(beta0 + beta1*streamDepth)
alpha0 <- 0; alpha1 <- -1
p2 <- plogis(alpha0 + alpha1*streamDepth)
```

Simulate abundance and removal data
  
```{r}
#| label: sim-cov3
N2 <- rpois(nSites, lambda=lambda2)         ## local abundance 
y2.all <- pi2 <- matrix(NA, nrow=nSites, ncol=K)
for(i in 1:nSites) {
    pi2[i,] <- c(p2[i], (1-p2[i])*p2[i], (1-p2[i])^2*p2[i], (1-p2[i])^3)
    y2.all[i,] <- rmultinom(n=1, size=N2[i], prob=pi2[i,])
}
y2 <- y2.all[,-K] ## Discard final column... individuals not detected
```
:::



## Simulated data

::: {.columns}
::: {.column width='50%'}
      
Observations

```{r}
#| label: sim-nocov-dat
y2[1:19,]
```
:::
::: {.column width='50%'}

<center>
Summary stats
</center>
    
Captures on each pass
  
```{r}
#| label: sim-nocov-ss2
#| size: 'scriptsize'
colSums(y2)
```

Naive abundance
  
```{r}
#| label: sim-nocov-ss3
sum(y2)
```
```{r}
#| label: un
#| include: false
library(unmarked)
```
:::
:::


## Prepare data in 'unmarked'
  
```{r}
#| label: un-umf
umf <- unmarkedFrameMPois(y=y2, siteCovs=data.frame(streamDepth),
                          type="removal")
summary(umf)
```

## Fit the model
 
::: {style='font-size: 70%;'}
`multinomPois` has similar arguments as `occu` and `pcount`. 

```{r}
#| label: un-fit
fm <- multinomPois(~streamDepth ~streamDepth, umf)    
fm
```


Compare to actual parameter values:

```{r}
#| label: un-compare
c(beta0=beta0, beta1=beta1); c(alpha0=alpha0, alpha1=alpha1)
```

:::



##  [Empirical Bayes -- Site-level abundance]{style="font-size: 90%;"}

```{r}
#| label: ranef
#| out-width: '80%'
#| fig-align: 'center'
#| fig-width: 9
re <- ranef(fm, K=15)
plot(re, layout=c(4,3), subset=site%in%1:12, xlim=c(-1, 11), lwd=5)
```



## Total abundance (in surveyed region)

```{r}
#| label: Ntotal
#| out-width: '60%'
#| fig-align: 'center'
N.total.post <- predict(re, func=sum, nsim=1000)
hist(N.total.post, freq=FALSE, main="", xlab="N total", ylab="Probability")
```


## Prediction in 'unmarked'
  
Create `data.frame` with prediction covariates. 

```{r}
#| label: preddat
pred.data <- data.frame(streamDepth=seq(-3, 3, length=20))
```

Get predictions of $\lambda$ for each row of prediction data.
  
```{r}
#| label: predpsi
lambda.pred <- predict(fm, newdata=pred.data, type='state', append=TRUE)
```


View $\lambda$ predictions
  
```{r}
#| label: psi-head
print(head(lambda.pred), digits=2)
```



## Prediction in 'unmarked'

```{r, echo=2:6}
#| label: pred-lam2
#| fig-width: 7
#| fig-height: 5
#| out-width: '80%'
#| fig-align: 'center'
par(mai=c(0.9,0.9,0.1,0.1))
plot(Predicted ~ streamDepth, lambda.pred, ylab="Expected value of abundance",
     ylim=c(0,20), xlab="Stream depth", type="l", lwd=2)
lines(lower ~ streamDepth, lambda.pred, col="grey")
lines(upper ~ streamDepth, lambda.pred, col="grey")
points(rowSums(y2)~streamDepth)
lines(lowess(rowSums(y2)~streamDepth), col="blue", lwd=2)  ## Lowess line for fun (it's way off)
legend(-3, 20, c("Multinomial N-mix model", "Lowess smooth"), lwd=2, col=c("black", "blue"))
```


## [Bayesian multinomial *N*-mixture models]{style="font-size: 90%;"}

::: {style='font-size: 80%;'}
There are several equivalent formulations of the multinomial that can be used to fit the model.
  
- Conditional-on-$N$, missing $y_{iK}$ (Not possible in JAGS)
- Conditional-on-$N$, conditional on $n_i=\sum_{k=1}^{K-1} y_{i,k}$
- Conditional-on-$N$, sequential binomial
- Marginalized $N$
  
These are just probability tricks that can affect MCMC performance. See AHM I for details. Here, we will focus on the second option.
:::


##  Conditional-on-*N* and *n~i~*

::: {style='font-size: 80%;'}
We can break the problem down into two steps by conditioning on $n_i$, the number of individuals captured at site $i$:
  
$$
  n_i \sim \mathrm{Bin}(N_i, 1-\pi_K) \\
  \{y_{i,1}, \dots, y_{i,K-1}\} \sim \mathrm{Multinomial}(n_i, \{\pi_{i,1}, \dots, \pi_{i,K-1}\}/(1-\pi_K))
$$
  
Note that the modified cell probabilities still sum to 1.
:::





##  Conditional-on-*N* and *n~i~*


[removal-mod2.jag](removal-mod2.jag)

```{r}
#| label: bugs-removal2
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("removal-mod2.jag"))
```


## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data2
jags.data.rem2 <- list(y=y2, n=rowSums(y2),
                       streamDepth=streamDepth,
                       nSites=nSites, nPasses=nPasses)
```

Initial values
  
```{r}
#| label: bugs-inits
jags.inits.rem <- function() {
    list(lambda.intercept=runif(1), alpha0=rnorm(1),
         N=rowSums(y2)+rpois(nrow(y2), 2))
}
```


Parameters to monitor
  
```{r}
#| label: bugs-pars
jags.pars.rem <- c("beta0", "beta1",
                   "alpha0", "alpha1", "totalAbundance")
```


## MCMC
  
```{r}
#| label: bugs-mcmc-rem2
#| message: false
#| cache: false
#| results: 'hide'
library(jagsUI); library(coda)
jags.post.rem2 <- jags.basic(data=jags.data.rem2, inits=jags.inits.rem,
                             parameters.to.save=jags.pars.rem, model.file="removal-mod2.jag",
                             n.chains=3, n.adapt=100, n.burnin=0, n.iter=2000, parallel=TRUE)
```

```{r}
#| label: bugs-sum-rem2
summary(jags.post.rem2[,jags.pars.rem])
```


## Traceplots and density plots

```{r}
#| label: bugs-plot1-rem2
#| out-width: "0.7<br>textwidth"
#| fig-align: 'center'
plot(jags.post.rem2[,jags.pars.rem[1:3]])
```

## Traceplots and density plots

```{r}
#| label: bugs-plot2-rem2
#| out-width: "0.7<br>textwidth"
#| fig-align: 'center'
plot(jags.post.rem2[,jags.pars.rem[4:5]])
```


## Convergence diagnostics

Gelman-Rubin diagnostic ($\hat{R}$)
```{r}
#| label: rhat
gelman.diag(jags.post.rem2)
```

Convergence is suggested by $\hat{R}$ close to 1. 



## Effective posterior sample size

It's a good idea to have >500 effective posterior samples. 

```{r}
#| label: ess
data.frame(ESS = effectiveSize(jags.post.rem2))
```



## Closing thoughts on removal models

::: {style='font-size: 80%;'}
In addition to site covariates, you could also include observation
covariates, but it's hard to think of realistic covariates that
would be specific to each pass.

<br>

::: {.fragment}
Removal models get used in many scenarios other than electrofishing:
  
- Point count surveys
  * Observers keep track of individuals during the survey
  * Analyze number of new individuals detected during each time interval
:::
    
::: {.fragment}
- Mark-recapture surveys
  * If behavior responses (trap shyness or trap happiness) are extreme, recaptures can be discarded and removal models can be applied.
:::
:::
      


# Double observer sampling

## [Double observer sampling (independent)]{style="font-size: 85%;"}
  
::: {style='font-size: 65%;'}
Two observers sample the same location at the same time, working independently.
  
After each survey, they compare notes and figure out which individual were detected by observer A only, B only, or by both.
  
::: {.fragment}
- $y_{i1}$ -- number of individuals detected only by observer A
- $y_{i2}$ -- number of individuals detected only by observer B
- $y_{i3}$ -- number of individuals detected by observers A and B
- $p_1$ -- probability that observer A detects an individual 
- $p_2$ -- probability that observer B detects an individual 
  
  Description                            |  Multinomial cell probability
  -------------------------------------- | ------------------------------
  Pr(detected by observer A but not B)   |      $\pi_1 = p_1(1-p_2)$
  Pr(detected by observer B but not A)   |      $\pi_2 = (1-p_1)p_2$
  Pr(detected by observers A and B)      |        $\pi_3 = p_1p_2$
  Pr(not detected)                       |    $\pi_4 = (1-p_1)(1-p_2)$
:::

:::


## [Double observer sampling (dependent)]{style="font-size: 85%;"}
  
::: {style='font-size: 65%;'}
Two observers sample at the same time, but observer B only records what observer A missed.
  
Often used in aerial waterfowl surveys. <br>
  
Observers should switch roles to ensure parameter identifiability. <br>
  
<!-- Definitions -->

::: {.fragment}
- $y_{i1}$ -- number of individuals detected by observer A
- $y_{i2}$ -- number of individuals missed by A but detected by B
- $p_1$ -- probability that observer A detects an individual 
- $p_2$ -- probability that observer B detects an individual 
  

  Description                            |  Multinomial cell probability
  -------------------------------------- | ------------------------------
  Pr(detected by observer A)             |         $\pi_1 = p_1$
  Pr(detected by observer B but not A)   |      $\pi_2 = (1-p_1)p_2$
  Pr(not detected)                       |    $\pi_3 = (1-p_1)(1-p_2)$
:::

:::


## [Simulating dependent double observer data]{style="font-size: 80%;"}
  
::: {style='font-size: 80%;'}
Abundance

```{r, echo=-1}
#| label: sim-doub-N
set.seed(839)
nSites <- 100
lambda <- 4.6  # Expected value of N
N <- rpois(n=nSites, lambda=lambda)
```
  
Detection probability and multinomial counts

```{r}
#| label: sim-doub-counts
observer <- matrix(c("A","B","B","A"),
                   nrow=nSites, ncol=2, byrow=TRUE)
p1 <- 0.4  # Detection prob for observer A
p2 <- 0.3  # Detection prob for observer B
piAfirst <- c(p1, (1-p1)*p2, (1-p1)*(1-p2))
piBfirst <- c(p2, (1-p2)*p1, (1-p1)*(1-p2))
K <- length(piAfirst)
y.all <- matrix(NA, nrow=nSites, ncol=K)
for(i in 1:nSites) {
  pi <- if(observer[i,1]=="A") piAfirst else piBfirst
  y.all[i,] <- rmultinom(n=1, size=N[i], prob=pi)    }
```
  
Discard final column of individuals not detected
```{r}
#| label: y-doub
y <- y.all[,-K]
```

:::


## [Dependent double observer model in JAGS]{style="font-size: 85%;"}

[double-obs-dep.jag](double-obs-dep.jag)
```{r}
#| label: bugs-double
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("double-obs-dep.jag"))
```


## Data, inits, and parameters

Put data in a named list
  
```{r}
#| label: bugs-data-double
jags.data.double <- list(y=y, n=rowSums(y), nSites=nSites,
    observer=ifelse(observer=="A", 1, 2))
```

Initial values
  
```{r}
#| label: bugs-inits-double
jags.inits.doub <- function() {
    list(lambda=runif(1), p1=runif(1), p2=runif(1),
         N=rowSums(y)+rpois(nrow(y), 2))
}
```

Parameters to monitor
  
```{r}
#| label: bugs-pars-doub
jags.pars.doub <- c("lambda", "p1", "p2", "totalAbundance")
```




## MCMC
  
```{r}
#| label: bugs-mcmc-doub
#| message: false
#| cache: false
#| results: 'hide'
jags.post.doub <- jags.basic(data=jags.data.double, inits=jags.inits.doub,
                             parameters.to.save=jags.pars.doub,
                             model.file="double-obs-dep.jag",
                             n.chains=3, n.adapt=100, n.burnin=0,
                             n.iter=2000, parallel=TRUE)
```

```{r}
#| label: bugs-sum-doub
#| size: 'scriptsize'
print(summary(jags.post.doub)$quantiles, digits=3)
```



## [Additional remarks on double observer models]{style="font-size: 80%;"}

::: {style='font-size: 80%;'}
::: {.fragment .semi-fade-out fragment-index=1}
In practice, there are often many observers on a project, even though only two survey at a time.
:::
  
::: {.fragment .fade-in-then-out .semi-fade-out fragment-index=1}
We can deal with this by including "observer ID" as an observation covariate. Everything else stays the same.
:::
  
::: {.fragment .fade-in-then-out .semi-fade-out fragment-index=2}
It can be hard to pull of the independent double observer method because observers often influence each other while working.
:::
  
::: {.fragment .fade-in-then-out .semi-fade-out fragment-index=3}
All of these methods require some sort of *short-term* individual identification of animals, even though they aren't being marked.
:::
  
::: {.fragment .fade-in-then-out .semi-fade-out fragment-index=4}
If short-term individual ID isn't possible, you could fall back on binomial $N$-mixture models.
:::
:::



# Assignment

## Assignment
  
::: {style='font-size: 70%;'}
Create a self-contained R script or Rmarkdown file to do the following:
  
- Simulate *independent* double observer data with the following properties:
  * nSites=200
  * $\lambda=3$
  * $p_1=0.3$ and $p_2=0.5$
- Fit the model using 'unmarked.' You will need to create an observation covariate indicating observer A and B. When you create the `unmarkedFrameMPois`, use `type="double"`.
- Fit the same model in JAGS
- For both unmarked and JAGS, use the point estimates of $p_1$ and $p_2$ to compute the $\pi$ probabilities.
  

<center>
Upload your `.R`, `.Rmd`, or `.qmd` file to ELC by 5:00pm on Tuesday.
</center>

:::

