---
title: "Lecture  13 - Dynamic N-mixture models"
author: "Heather Gaya"
date: "November 12, 2020"
output: beamer_presentation
header-includes: \AtBeginSection[]{\begin{frame}\tableofcontents[currentsection]\end{frame}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(unmarked)
library(coda)
library(jagsUI)
library(knitr)
```
# Models
## Overview

  We've previously seen N-mixture models across only one year or season 
  \pause
  \vfill
  But often we want to know how populations change over time. 
  \pause
  \vfill
  Dynamic $N$-mixture models are useful if we have repeated count data without uniquely marked individuals. 
  \pause
  \vfill
 Dynamic models are also very flexible. We can consider how detection parameters change between years, add temporal autocorrelation, account for spatial variation in density, and/or account for treatment effects.   

## Dynamic N-Mixtures 
State model (with Poisson assumption)
  \begin{gather*}
    \mathrm{log}(\lambda_{it}) = \beta_0 + \beta_1 {\color{blue} x_{it1}} +
    \beta_2 {\color{blue} x_{it2}} + \cdots \\
    N_{it} \sim \mathrm{Poisson}(\lambda_{it})
  \end{gather*}
  \pause
  Observation model
  \begin{gather*}
    \mathrm{logit}(p_{ijt}) = \alpha_0 + \alpha_1 {\color{blue} w_{it1}}
    + \alpha_2 {\color{purple} w_{ijt2}} + \cdots \\
    y_{ijt} \sim \mathrm{Binomial}(N_{it}, p_{ijt})
  \end{gather*}
  \pause
  \small
  $\lambda_{it}$ -- Expected value of abundance at site $i$ in year $t$
  
  $N_{it}$ -- Realized value of abundance at site $i$ in year $t$
  
  $p_{ijt}$ -- Probability of detecting \alert{an individual} at site $i$ on occasion $j$ in year $t$
  
  $y_{ijt}$ -- Count data
  
  $\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates 
  
  $\color{purple} w_1$ and $\color{purple} w_2$ -- observation covariates


## Temporal Autocorrelation
The previous model allows for independence between years.

\pause
\vfill

But we might expect that abundance at year $t$ is related to abundance at year $t-1$.

\pause
\vfill
Dail and Madsen (2011) suggested an expansion to the $N$-mixture framework that allows us to consider populations from this B.I.D.E model perspective.

## Temporal Autocorrelation (Cont')
\small
Year $k = 1$:
\begin{gather*}
    \mathrm{log}(\lambda_{i1}) = \beta_0 + \beta_1 {\color{blue} x_{i1}} \cdots \\
    N_{i1} \sim \mathrm{Poisson}(\lambda_{i1})
  \end{gather*}
  \pause
  \small
Years $k > 1$:
  \begin{gather*}
    S_{it} \sim \mathrm{Binomial}(N_{i(t-1)}, \phi_{it}) \\
    G_{it} \sim \mathrm{Poisson}(N_{i(t-1)}*\gamma_{it}) \\
    N_{it} = S_{it} + G_{it}
  \end{gather*}
  \small
  \pause
$\lambda_{it}$ -- Expected value of abundance at site $i$ in year 1

$N_{it}$ -- Realized value of abundance at site $i$ in year $t$

$\color{blue} x_1$ and $\color{blue} x_2$ -- site covariates

$\phi_{it}$ -- Apparent survival at site $i$ from year $t-1$ to $t$

$\gamma_{it}$ -- Apparent recuritment at site $i$ from year $t-1$ to $t$

$S_{it}$ -- Realized number of individuals that survived/didn't emmigrate

$G_{it}$ -- Realized number of new individuals

# Simulation

## Simulate Site-Specific Abundance
Simulate some data
```{r Simulate Dynamic,eval= TRUE, echo = TRUE}
set.seed(100)
n.sites <- 50
n.years <- 5
elevation <- scale(runif(n.sites,800,1200))
precip <- array(runif(n.sites*n.years, 0,2), 
              dim = c(n.sites, n.years))
```

## Simulate Site-Specific Parameters
Simulate initial N, $\phi$ and $\gamma$
```{r, eval  = TRUE, echo = TRUE}
psi0 <- 2.5
psi1 <- -.25
phi0 <-2
phi1 <- -1
gamma0 <- -1.8
gamma1 <- 1
phi <- plogis(phi0 + phi1*precip)
gamma <- exp(gamma0 + gamma1*as.numeric(elevation))
```

## Simulate Abundance
```{r, echo = TRUE}
N <- S <- G <- array(NA, dim = c(n.sites, n.years))
lambda <- array(NA, dim = n.sites)
for (i in 1:n.sites){
    lambda[i] <- exp(psi0 + psi1*elevation[i])
    N[i,1] <- rpois(1, lambda[i])
 for (t in 2:n.years){
   S[i,t] <- rbinom(1, N[i,t-1], phi[i,t])
   G[i,t] <- rpois(1, N[i,t-1]*gamma[i])
   N[i,t] <- S[i,t] + G[i,t]
 }
}
```
\pause
```{r, echo = T, include=T}
N[1:3,]
```

## Simulate Detection
```{r, echo = T}
n.visit <- 4
p <- rnorm(n.years, mean = .5, .1)
y1 <- array(NA, dim = c(n.sites, n.visit, n.years))
for (i in 1:n.sites){
  for (t in 1:n.years){
    y1[i,,t] <- rbinom(n.visit, size = N[i,t], 
                       prob = p[t])
  }
}
```

# Unmarked 

## Setup in Unmarked
Unmarked wants the observation data in a M x (J*T) matrix
```{r unmarked, echo = T}
y_matrix <- cbind(y1[,,1], y1[,,2], y1[,,3], 
                  y1[,,4], y1[,,5])
precip <- data.frame(precip)
#names(precip) <- c("p1", "p2", "p3", "p4", "p5")
umf <- unmarkedFramePCO(y = y_matrix, 
        siteCovs = data.frame(elevation),
        yearlySiteCovs = list(precip=precip),
        numPrimary = 5)
```
\pause
For some reason this will not run ? Hmm.
```{r, echo = T, eval = F}
m1 <- pcountOpen(~elevation, ~elevation, 
                 ~1, ~1, umf, K=50, 
                 mixture = "P",
                 dynamics = "autoreg")
```
Order of equations is lambda, gamma, apparent survival, detection

## Unmarked Output

If I ever get this to run there will be some output. ugh.

# JAGS

## Code
\tiny
```{r, eval = F, include = T, echo = T}
model {
for (i in 1:n.sites) {
  log(gamma[i]) <- gamma0 + gamma1*elevation[i]
  log(lambda[i]) <- psi0 + psi1*elevation[i]
  
  N[i,1] ~dpois(lambda[i]) 
  
  for (t in 2:n.years){
    logit(phi[i,t]) <- phi0 + phi1*precip[i,t]
    S[i,t] ~ dbin(phi[i,t], N[i,t-1])
    G[i,t] ~ dpois(N[i,t-1]*gamma[i])
             
    N[i,t] <- S[i,t] + G[i,t]
  } #end t
  for (t in 1:n.years){
  for (j in 1:n.visit){
    y1[i,j,t] ~ dbin(p[t], N[i,t])
    } #end j
  
  p[t] ~ dunif(0,1)
  }#end t again
  } #end i
  gamma0 ~ dunif(-5,5)
  gamma1 ~ dunif(-5,5)
  psi0 ~ dunif(-5,5)
  psi1 ~ dunif(-5,5)
  phi0 ~ dunif(-5,5)
  phi1 ~ dunif(-5,5)
}
```

## Send to JAGS
\small
```{r, eval = T, echo = T, include = T}
jd <- list(n.sites =n.sites, n.visit = n.visit,
           n.years = n.years, y1 = y1,
           elevation = as.numeric(elevation),
           precip = precip)
params = c("p", "psi0", "psi1", 
           "phi0", "phi1", 
           "gamma0", "gamma1")
ji <- function() {
  list(psi0 = runif(1,1.5,2.5),psi1 = runif(1, -.5,0), 
    phi0 = runif(1,1.5,2.5), phi1 = runif(1,-1.5,-.5),
    gamma0 = runif(1,-2,-1.5), gamma1 = runif(1,.5,1.5), 
    p = rnorm(n.years, mean = .5, .1))}
```
```{r, eval =T, include = F}
ji <- function(){list(psi0 = psi0, psi1 = psi1, phi0 = phi0, phi1=phi1, gamma0=gamma0, gamma1 = gamma1,N = cbind(N[,1], matrix(NA, nrow = n.sites, ncol = n.years -1)), p=p, S = S, G= G)}
#You wouldn't have this info in real life to use as initials 
#But made it faster to find good inits for example
```
\pause
```{r, eval = T, echo = T,results = 'hide'}
jags.post <- jags.basic(data=jd, inits = ji,
                   parameters.to.save=params,
                   model.file="dynamic_n.txt",
                   n.chains=3, n.adapt=100, n.burnin=2000,
                   n.iter=3000, parallel=TRUE)
```

## JAGS Output 
\small
```{r, echo = T}
kable(summary(jags.post)$quantiles[-1,], 
      digits = 2,align = "c")
```

## Trace Plot
```{r}
plot(jags.post[,c("psi0","psi1", "phi0"),])
```
Need to run for longer

## Compare with Truth
Pretty close to the simulated values
```{r, echo =F}
true <- c(gamma0, gamma1, p, phi0, phi1, psi0, psi1)
est <- summary(jags.post)$quantiles[-1,3]
outs <- data.frame(Simulated = true, JAGS = est)
kable(outs, digits = 2, align = "c")
```

# Assignment 

## Assignment 

Idk if there's any assignment


