---
## title: "Converted Presentation"
format:
  revealjs:
    theme: default
    pagetitle: 'Spatial capture-recapture for closed populations: Part II <br> Density surfaces'
    slide-number: true
    echo: true
lang: en
---

## {content-align="center" style="font-size: 150%;"}

<br>

<center>
  
**Spatial capture-recapture for <br> closed populations: Part II <br> Mapping density surfaces and activity centers**

<br>

::: {style='font-size: 60%;'}
  WILD(FISH) 8390 <br>
  Inference for Models of Fish and Wildlife Population Dynamics <br>
:::
  
</center>



# Overview {visibility="hidden"}



##  Closed population model

::: {style='font-size: 50%;'}
State model (a spatial point process model) 
$$
    \lambda({\bf s}) = \exp(\beta_0 + \beta_1 w_1({\bf s}) + \beta_2 w_2({\bf s}) + \cdots) \\
    \Lambda = \int_{\mathcal{S}} \lambda({\bf s}) \; \mathrm{d}{\bf s} \\
    N \sim \mathrm{Pois}(\Lambda) \\
    {\bf s}_i \propto \lambda({\bf s}) \;\; \mathrm{for}\; i=1,\dots,N 
$$

Observation model (supposing $N$ was known)
$$
    p_{ij} = g_0\exp(-\|{\bf s}_i - {\bf x}_j\|^2/(2\sigma^2))  \;\; \mathrm{for}\, j=1,\dots,J \\
    y_{ijk} \sim \mathrm{Bernoulli}(p_{ij})
$$

:::

::: {style='font-size: 50%;'}
  $\lambda({\bf s})$ -- "Intensity function" describing expected density of individuals at location $\bf s$ <br> 
  $\Lambda$ -- Expected number of individuals in the spatial region $\mathcal{S}$ <br>
  $N$ -- Realized number of individuals (ie, population size) <br>
  ${\bf s}_i$ -- Location of the $i$th activity center <br>
  ${\bf x}_j$ -- Location of trap $j$ <br>
  $\|{\bf s}_i - {\bf x}_j\|$ -- Euclidean distance between ${\bf s}_i$ and ${\bf x}_j$ <br>
  $g_0$ -- Capture probability when distance between activity centers
  and traps is 0 <br>
  $\sigma$ -- Scale parameter of encounter function <br>
  $p_{ij}$ -- Capture probability <br>
  $y_{ijk}$ -- Spatial capture histories <br>
:::



## Spatial point processes

The state model of SCR is a spatial (or spatio-temporal) point process
  
There are many varieties of spatial point processes, including:
    * (In)homogeneous binomial point process
    * (In)homogeneous Poisson point process
    * Cox process
    * Gibbs process
    * Markov point process



## Spatial point processes

All share a few properties
    * The data are a collection of points called a "point pattern"
    * Points are in an area called the state-space ($\mathcal{S}$), or observation window, which is usually two dimensional
    * An intensity function ($\lambda(\bs)$) describes spatial variation in the density of points
    * The area under this function is the expected number of points (a.k.a, $N$) in the region:
  
  
$$
  E(N) = \Lambda = \int_{\mathcal{S}} \lambda({\bf s}) \;\mathrm{d}{\bf s}
$$


## Poisson point process

Properties
    * If density is constant throughout the state-space: $\lambda(\bs) = \lambda$, the process is said to be "homogeneous"
    * Otherwise, the process is "inhomogeneous"
    * The number of points in any region of the state-space is Poisson distributed
    * Points are independent of one another (no attraction or repulsion) 
  
An inhomogeneous point process allows for spatial variation in density, which can be modeled using [spatial covariates]{style="color: red;"}, such as raster layers.


## Density surfaces
  
We can use a fitted model to make maps of the following: 

1. Expected density surface%, $\lambda({\bf s})$
    * This is an estimate of $\lambda({\bf s})$.
    * It's the expected value of density at  each location.
    * It's good for describing and predicting density at a different point in time, assuming the processes controlling density don't change.
2. Realized density surface
    * This is an estimate of the number of activity centers per unit area [during the sampling period]{style="color: red;"}.
    * It's an estimate of the realized point pattern.
    * It will be equivalent to the expected density surface far from the traps. 
3. Activity center locations
    * The posterior distribution of each activity center.
    * This is not a home range map.
      
  

# Simulation

## Inhomogeneous Poisson point process

First, let's import a raster layer

```{r}
#| label: ippp1
#| fig-width: 7.2
#| out-width: "60%"
#| fig-align: "center"
#| results: "hide"
#| message: false
library(raster)
elevation <- raster("elevation.tif")
delta <- res(elevation)[1]  ## resolution
plot(elevation, col=topo.colors(100), main="Elevation")
```






## Inhomogeneous Poisson point process
  
Second, let's pick some coefficients and create a density surface

```{r}
#| label: ippp2
#| fig-width: 7.2
#| out-width: "60%"
#| fig-align: "center"
beta0 <- -15
beta1 <- 0.01 #0.005
lambda <- exp(beta0 + beta1*elevation) # Intensity function
plot(lambda, col=terrain.colors(100), main="Density surface", zlim=c(0,0.18))
```




## Inhomogeneous Poisson point process
  
Third, simulate $N$
  
```{r}
#| label: ippp3
set.seed(538)  
ds <- 1                          ## Pixel area is 1 ha
lambda.values <- values(lambda)  ## Convert raster to vector
Lambda <- sum(lambda.values*ds)  ## E(N)
(N <- rpois(1, Lambda))          ## Realized N
```

Fourth, simulate ${\bf s}_1, \dots, {\bf s}_N$. To do this, we'll pick pixels proportional to density. Then we'll jitter each point inside its pixel. 
  
```{r}
#| label: ipp4
n.pixels <- length(lambda)
jitter <- 0.005                    ## Half width of pixel 
s.pixels <- sample(n.pixels, size=N, replace=TRUE,
                   prob=lambda.values/Lambda)
elevation.xyz <- as.data.frame(elevation, xy=TRUE)
s <- elevation.xyz[s.pixels,c("x","y")] +
    cbind(runif(N, -jitter, jitter),runif(N, -jitter, jitter))
```




## Inhomogeneous Poisson point process

```{r}
#| label: ippp5
#| fig-width: 7.2
#| out-width: "70%"
#| fig-align: "center"
plot(lambda, col=terrain.colors(100),
     main="Density surface with activity centers")
points(s, pch=16, cex=1, col="blue")
```





## Traps

```{r}
#| label: traps1
#| fig-width: 7.2
#| out-width: "60%"
#| fig-align: "center"
x <- cbind(rep(seq(0.15, 0.85, by=0.1), each=8),
           rep(seq(0.15, 0.85, by=0.1), times=8))  ## Trap locations
plot(lambda, col=terrain.colors(100),
     main="Density surface with activity centers and traps")
points(s, pch=16, col="blue") ## Activity center locations
points(x, pch=3)              ## Trap locations
```





## Distance between traps and activity centers

Compute distances between activity centers (${\bf s}_1, \dots, {\bf s}_N$) and traps (${\bf x}_1, \dots, {\bf x}_J$).
```{r}
#| label: dist1
J <- nrow(x)                 ## nTraps
dist.sx <- matrix(NA, N, J)  
for(i in 1:N) {
    dist.sx[i,] <- sqrt((s[i,1]-x[,1])^2 + (s[i,2]-x[,2])^2)
}
```


Look at distances between first 4 individuals and first 5 traps.

```{r}
#| label: dist2
dist.sx[1:4,1:5]
```




## Capture probability

Compute capture probability
```{r}
#| label: p1
g0 <- 0.2
sigma <- 0.05
p <- g0*exp(-dist.sx^2/(2*sigma^2))
```



Look at capture probs for first 4 individuals and first 5 traps.
```{r}
#| label: p2
print(p[1:4,1:5], digits=3)
```



## Capture histories

Simulate capture histories for all $N$ individuals
```{r}
#| label: y1
K <- 5                          # nOccasions
y.all <- array(NA, c(N, J, K))
for(i in 1:N) {
    for(j in 1:J) {
        y.all[i,j,] <- rbinom(K, 1, prob=p[i,j])
    }
}
```


  Discard individuals not captured
```{r}
#| label: y2
captured <- rowSums(y.all)>0
y <- y.all[captured,,]
```


  Capture histories for first 2 individuals and first 5 traps
  on first occasion.
```{r}
#| label: y3
y[1:2,1:5,1]
```




# Likelihood



```{r}
#| label: write
#| include: false
#| echo: false
#| results: "hide"
#| purl: false
ch.out <- data.frame(session=1,
                     individual=rep(slice.index(y, 1), y),
                     occasion=rep(slice.index(y, 3), y),
                     trap=rep(slice.index(y, 2), y))
write.table(ch.out, file="encounter_data_file.csv",
            row.names=FALSE, col.names=FALSE, sep=",")
traps.out <- data.frame(trap=1:nrow(x), x*1000)
write.table(traps.out, file="trap_data_file.csv",
            row.names=FALSE, col.names=FALSE, sep=",")
library(secr)
```



## R package 'secr'


Import data from two text files.
```{r}
#| label: secr-in
#| warning: false
#| results: 'hide'
#| message: false
library(secr)  
sch <- read.capthist(captfile="encounter_data_file.csv",
                     trapfile="trap_data_file.csv",
                     detector="proximity", fmt="trapID")
```
  
  Create the ``habitat mask'', which defines the state-space and the
  spatial covariates. First, prepare the elevation covariate. 
```{r}
#| label: format-elevation
#| size: 'scriptsize'
elevation.xyz <- as.data.frame(elevation, xy=TRUE)
elevation.xyz$y <- round(elevation.xyz$y, 3) ## Fix numerical fuzz
elevation.xyz.m <- elevation.xyz
elevation.xyz.m$x <- elevation.xyz$x*1000  ## Convert units to meters
elevation.xyz.m$y <- elevation.xyz$y*1000
elevation.xyz.m$elevation <- scale(elevation.xyz$elevation) ## Standardize
elevation.m <- rasterFromXYZ(elevation.xyz.m)
```
  
  Now create the mask.
```{r}
#| label: make-mask
#| size: 'scriptsize'
library(sp)
elev.sp <- as(elevation.m, "SpatialGridDataFrame")
trp <- traps(sch)
mask <- make.mask(trp, buffer=150, spacing=10)
mask <- addCovariates(mask, spatialdata=elev.sp)
```





## Habitat mask

```{r}
#| label: plot-mask
#| fig-height: 5
#| out-width: "95%"
#| fig-align: 'center'
#| echo: -1
#| size: 'scriptsize'
par(mai=c(0.1, 0.1, 0.1, 0.1))
plot(mask, covariate="elevation")
plot(trp, add=TRUE)
```





## R package 'secr'

Model density as a function of elevation. 
```{r}
#| label: secr-elev
#| cache: true
fm.elev <- secr.fit(sch, model=list(D=~elevation, g0=~1, sigma=~1),
                    mask=mask, trace=FALSE)  ## Don't use 'buffer'
coef(fm.elev)
```


Estimates on original scale (at average covariate values).
```{r}
#| label: secr-M0-real
#| size: 'scriptsize'
predict(fm.elev)
```




## Estimated density surface

```{r}
#| label: secr-dsurf
#| fig-height: 6.8
#| warning: false
## #| fig.show: 'hide'
dsurf <- predictDsurface(fm.elev)
dsurf.r <- raster(dsurf, covariate="D.0")
pix.area <- (delta*1000)^2
plot(dsurf.r/pix.area, col=terrain.colors(100), zlim=c(0,0.18),
     main="Estimated density surface")
```

<!-- [width=0.49\textwidth]{figure/secr-dsurf-1} -->
<!-- [width=0.49\textwidth]{figure/ippp2-1} <br> -->
<!-- %  -->
<!-- %The units differ because we transformed to meters. <br> -->






## Estimated activity centers
  
Contour map for the Empirical Bayes posterior distribution of the activity center for individual 1. 

```{r}
#| label: fxi
#| eval: true
#| out-width: '65%'
#| fig-align: 'center'
fxiContour(fm.elev, i=1)
```





## R package 'secr'

We can still estimate $N$ as before.
  
`E.N` is the expected value of $N$. `R.N` is the realized value of $N$. 
  
```{r}
#| label: regionN-M0
region.N(fm.elev)
```




# Bayes



## Data augmentation model

DA version of SCR model with inhomogeneous point process <br>
  
Version 1 

$$
  \lambda(\bs) = \exp(\beta_0 + \beta_1 \mathrm{ELEV}(\bs)) \\
  \bsi \propto \lambda(\bs) \\
  \Lambda = E(N) = \int_{\mathcal{S}} \lambda(\bs)\; \mathrm{d}\bs \\
  \psi = \Lambda/M \\
  z_i \sim \mathrm{Bern}(\psi) \\
  p_{ij} = g_0\exp(-\|\bsi-\bxj\|^2/(2\sigma^2)) \\
  y_{ijk} \sim \mathrm{Bern}(z_i\times p_{ij}) \\
  N=\sum_{i=1}^M z_i
$$
   
This version is relatively slow and doesn't mix that well.



## SCR with IPP -- version 1

[SCR-elev-v1.jag](SCR-elev-v1.jag)

```{r}
#| label: bugs-SCR-elev
writeLines(readLines("SCR-elev-v1.jag"))
```




<!-- %   ## Add code to convert from secr format to 3D array -->



## Data augmentation model

DA version of SCR model with inhomogeneous point process <br>
  
  Version 2 <br>
  \begin{gather*}
    \tilde{\lambda}(\bs) = \exp(\beta_1 \mathrm{ELEV}(\bs)) <br>
    \bsi \propto \tilde{\lambda}(\bs) <br>
%    \Lambda = E(N) = \int_{\mathcal{S}} \lambda(\bs)\; \mathrm{d}\bs <br>
%    \psi \sim \mathrm{Unif}(0,1) <br>
    z_i \sim \mathrm{Bern}(\psi) <br>
    p_{ij} = g_0\exp(-\|\bsi-\bxj\|^2/(2\sigma^2)) <br>
    y_{ijk} \sim \mathrm{Bern}(z_i\times p_{ij}) <br>
    N=\sum_{i=1}^M z_i
  \end{gather*}
%   
  This version is faster and mixes better. <br>








## SCR with IPP -- version 2

```{r}
#| label: bugs-SCR-elev-v2-R
#| size: 'tiny'
#| eval: false
writeLines(readLines("SCR-elev-v2.jag"))
```
```{r}
#| label: bugs-SCR-elev-v2
#| size: 'tiny'
#| echo: false
#| comment: ''
#| background: 'beige'
writeLines(readLines("SCR-elev-v2.jag"))
```






## SCR with IPP -- version 3

There's another version where we treat space as discrete, rather than continuous. 
   
Under this formulation, activity centers are snapped to pixel centers, and a categorical prior is used instead of a continuous distribution.
   
This version is similar to what is used in likelihood-based methods.
   
It works fine, but it's slow in JAGS and it adds an extra level of approximation.



## Inhomogeneous point process in JAGS

We need to do the following to use version 2:
    * Provide the spatial covariate as a vector
    * Loop over all pixels to compute the *conditional*-on-$N$ density surface. 
    * Retain the uniform prior on the activity centers, but add a second probability component to force them to be distributed according to the density surface.
    * This requires that we create a "lookup" table to map points in space to pixel IDs
    * Then we use the "zero's trick" to implement the non-standard probability distribution
    * Put a uniform prior on $\psi$.
  
If we were using version 1, we would define $\psi=E(N)/M$ and we'd include $\beta_0$ in the intensity function.


```{r}
#| label: rjags
#| include: false
#| results: "hide"
library(rjags)
```



## SCR with IPP

Create the lookup table
```{r}
#| label: lookup
#| fig-show: 'hide'
#| echo: -1
#par(mai=c(0.1,0.1,0.3,0.1))  
G <- nrow(elevation.xyz)    ## nPixels  
lookup <- matrix(1:G, nrow=nrow(elevation), ncol=ncol(elevation),
                 byrow=TRUE) ## To be consistent with raster package
delta <- res(elevation)[1]  ## resolution
image(seq(delta/2, 1-delta/2, length=100),
      seq(delta/2, 1-delta/2, length=100), lookup, asp=1,
      xlab="x", ylab="y", frame=FALSE, col=0, main="Pixel ID")
text(elevation.xyz$x[seq(5, G, by=10)],
     elevation.xyz$y[seq(5, G, by=10)], seq(5, G, by=10), cex=0.4)
```



## Lookup table
  
[width=0.8\textwidth]{figure/lookup-1}






## SCR with IPP

Data
  
```{r}
#| label: jd-SCR-elev-faster
M <- 150  
y.tilde <- apply(y, c(1,2), sum)
n <- nrow(y)
jags.data.SCR.elev <- list(
    y.tilde=y.tilde, n=n, M=M, J=J, G=G, delta=delta,
    lookup=lookup, pixelArea=delta^2*1e4, ## Convert to hectares
    elevation=elevation.xyz$elevation, z=c(rep(1, n), rep(NA, M-n)),
    zeros=rep(0, M), K=K, zero.cap=rep(0, M), x=x, xlim=c(0,1), ylim=c(0,1))
```

Inits and parameters

```{r}
#| label: jp-ji-SCR-elev
jp.SCR.elev <- c("beta0", "beta1", "g0", "sigma", "EN", "N")
ji.SCR.elev <- function() {
    list(z=c(rep(NA, n), rep(0,M-n)),#psi=runif(1),beta0=runif(1,-20,-15),
         beta1=0.01, s=cbind(runif(M), runif(M)),
         g0=runif(1), sigma=runif(1, 0.05, 0.1)) }
```

MCMC
  
```{r}
#| label: mcmc-SCR-elev
#| results: 'hide'
#| cache: true
#| message: false
library(jagsUI)  
jags.post.SCR.elev <- jags.basic(
    data=jags.data.SCR.elev, inits=ji.SCR.elev,
    parameters.to.save=jp.SCR.elev, model.file="SCR-elev-v2.jag",
    n.chains=3, n.adapt=100, n.iter=1000, parallel=TRUE)
```




## Posterior summaries

```{r}
#| label: summary-mcmc-SCR0-faster
summary(jags.post.SCR.elev)
```



## Traceplots and density plots

```{r}
#| label: plot-mcmc-SCR-elev1
#| out-width: "70%"
#| fig-align: 'center'
plot(jags.post.SCR.elev[,jp.SCR.elev[1:3]])
```






## Traceplots and density plots

```{r}
#| label: plot-mcmc-SCR-elev2
#| out-width: "70%"
#| fig-align: 'center'
plot(jags.post.SCR.elev[,jp.SCR.elev[4:6]])
```




## Predicted density surface in JAGS

Extract posterior samples of $\beta_0$ and $\beta_1$. 
```{r}
#| label: extract-beta
beta.post <- as.matrix(jags.post.SCR.elev[,c("beta0","beta1")])
n.samples <- nrow(beta.post)
```

Compute density surface for every MCMC sample

```{r}
#| label: lambda-samples
lambda.post <- matrix(NA, n.pixels, n.samples)
for(i in 1:n.samples) {
    lambda.post[,i] <- exp(
        beta.post[i,"beta0"] +
        beta.post[i,"beta1"]*elevation.xyz$elevation)
}
lambda.post.mean <- rowMeans(lambda.post)
lambda.post.lower <- apply(lambda.post, 1, quantile, prob=0.025)
lambda.post.upper <- apply(lambda.post, 1, quantile, prob=0.975)
```




## Predicted density surface

Make multi-panel figure showing posterior mean density surface, along with 95\% CI.
```{r}
#| label: lambda-post
#| fig-height: 5
#| fig-show: "hide"
#| dpi: 400
library(latticeExtra)
trellis.par.set(regions=list(col=hcl.colors(100))) ## Colors
panel1 <- levelplot(lambda.post.lower ~ x+y, elevation.xyz, aspect="iso",
                    at=seq(0,0.20,0.005), ## Resolution of color key
                    colorkey=list(space="bottom"),
                    xlab="", ylab="")
panel2 <- levelplot(lambda.post.mean ~x+y, elevation.xyz, aspect="iso",
                    at=seq(0,0.20,0.005))
panel3 <- levelplot(lambda.post.upper ~x+y, elevation.xyz, aspect="iso",
                    at=seq(0,0.20,0.005))
panels <- c(panel1, panel2, panel3)
panels <- update( 
    panels, scales=list(draw=FALSE), layout=c(3,1), 
    xlab="", ylab="", 
    strip=strip.custom(bg=gray(0.8),
        factor.levels=c("Lower CI","Posterior mean","Upper CI")))
plot(panels)
```




## Predicted density surface

Posterior mean and 95% CI

[width=\textwidth]{figure/lambda-post-1}  <br>






## Realized density surface

We need to monitor $z$ and $\bsi$.

```{r}
#| label: mcmc-SCR-elev-sz
#| results: 'hide'
#| cache: true
jags.post.SCR.elev.sz <- jags.basic(
    data=jags.data.SCR.elev, inits=ji.SCR.elev,
    parameters.to.save=c(jp.SCR.elev,"s","z"),
    model.file="SCR-elev-v2.jag",
    n.chains=3, n.adapt=100, n.iter=1000, parallel=TRUE)
```
 
Extract posterior samples of $z$ and $\bsi$. 

```{r}
#| label: extract-sz
s1.post <- as.matrix(jags.post.SCR.elev.sz[,paste0("s[", 1:M, ",1]")])
s2.post <- as.matrix(jags.post.SCR.elev.sz[,paste0("s[", 1:M, ",2]")])
z.post <- as.matrix(jags.post.SCR.elev.sz[,paste0("z[", 1:M, "]")])
```
 
Chop activity centers into discrete intervals, then tabulate.

```{r}
#| label: cut-s
lambda.r.post <- array(NA, c(sqrt(n.pixels), sqrt(n.pixels), n.samples))
for(i in 1:n.samples) {
    si1.post.d <- cut(s1.post[i,z.post[i,]==1], breaks=seq(0, 1, delta))
    si2.post.d <- cut(s2.post[i,z.post[i,]==1], breaks=seq(0, 1, delta))
    pixel.counts <- table(si2.post.d, si1.post.d)
    lambda.r.post[,,i] <- pixel.counts[sqrt(n.pixels):1,]
}
```



## Realized density surface

```{r, echo=-1}
#| label: lambda-r
#| fig-height: 3
#| out-width: '99%'
par(omi=c(0,0,0,0.2))  
lambda.r.mean <- raster(apply(lambda.r.post, c(1,2), mean))
lambda.e.mean <- raster(matrix(lambda.post.mean, sqrt(n.pixels), byrow=TRUE))
lambda.re <- stack(lambda.r.mean, lambda.e.mean)
names(lambda.re) <- c("Realized", "Expected")
plot(lambda.re, zlim=c(0,0.18))
```





## Posterior distribution of $\bsi$
  
Posterior distribution of $\bs_1$. 

```{r, echo=-1}
#| label: si-post
#| fig-show: 'hide'
par(mai=c(0.0,0.0,0.0,0.0))
plot(x, asp=1, xlim=0:1, ylim=0:1, pch=3, axes=FALSE, ann=FALSE)
points(s1.post[,1], s2.post[,1], pch=16, col=rgb(0,0,1,0.1))
contour(MASS::kde2d(s1.post[,1], s2.post[,1]), add=TRUE)
points(x[y.tilde[1,]>0,,drop=FALSE], pch=3, col="red", lwd=2, cex=2)
```


[width=0.75\textwidth]{figure/si-post-1} <br>



## Posterior distribution of $\bsi$
  
Posterior distribution of $\bs_i$

```{r}
#| label: si-post-i
#| echo: false
#| fig-show: 'hide'
for(i in 1:n) {
    par(mai=c(0.0,0.0,0.0,0.0))
    plot(x, asp=1, xlim=0:1, ylim=0:1, pch=3, axes=FALSE, ann=FALSE)
    points(s1.post[,i], s2.post[,i], pch=16, col=rgb(0,0,1,0.1))
    contour(MASS::kde2d(s1.post[,i], s2.post[,i]), add=TRUE)
    points(x[y.tilde[i,]>0,,drop=FALSE], pch=3, col="red", lwd=2, cex=2)
}
```
%


%\only<1>{[width=0.85\textwidth]{figure/si-post-i-2} <br>}
%\only<2>{[width=0.85\textwidth]{figure/si-post-i-3} <br>}
% \only<3>{[width=0.85\textwidth]{figure/si-post-i-4} <br>}
\foreach \n in {1,...,30} {%
  \only<\n>{[width=0.85\textwidth]{figure/si-post-i-\n}<br>}
}






## Posterior distribution of * **s**~i~ *
  
Posterior distribution of $\bsi$ for an indiviudal \alert{not} captured. 

```{r, echo=-(1:2)}
#| label: si-post-y0
#| fig.show: 'hide'
par(mai=c(0.0,0.0,0.0,0.0))
## plot(x, asp=1, xlim=0:1, ylim=0:1, pch=3, axes=FALSE, ann=FALSE)
image(MASS::kde2d(s1.post[z.post[,150]==1,150],
                  s2.post[z.post[,150]==1,150]))
points(x, asp=1, pch=3)
##points(s1.post[z.post[,150]==1,150],
##       s2.post[z.post[,150]==1,150], pch=16, col=rgb(0,0,1,0.8))
```


[width=0.5\textwidth]{figure/si-post-y0-1} <br>




## Summary

Just as we can estimate both the realized and expected value of $N$, we can also estimate the realized and expected value of abundance (or density) at each location in the region of interest.
   
Sometimes, the region of interest will be different than $\cal S$, and we can easily obtain estimates of abundance in subsets of $\mathcal{S}$.
   
When we get back to SCR later, we'll talk about modeling both spatial and temporal variation in density.




# Assignment {visibility="hidden"}


## Assignment

<!-- % Create a self-contained R script or Rmarkdown file to do the -->
<!--   % following:  -->
<!--   %  -->
<!--   %  -->
<!--   %   * Fit a ``local behavioral response'' model, -->
<!--   %     rather than a ``global behavioral response'' model in secr. -->
<!--   %   * Fit a model in JAGS where $g_0$ varies among -->
<!--   %     occasions. Compare estimates of $E(N)$ and $N$ to the estimates -->
<!--   %     from secr for the same model (which we fit earlier). -->
<!--   %  -->
<!--   %  -->
<!--   % Upload your {\tt .R} or {\tt .Rmd} file to ELC before Tuesday. -->

No assignment this week. Work on your paper. <br>







